---
title: Exploratory - Afghanistan Rainfall and ONI
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  include: true
  echo: true
  warning: false
  message: false
  eval: true
  results: "asis"
  out.width: "100%"
  code-fold: true
editor: visual
project:
  execute-dir: project
---

## Intro

**Document objectives:**

- Geography - province level with emphasis on **Hirat**
- Explore CHIRPS data to better understand seasonality of rainfall
- Explore relationship between CHIRPS data and ENSO ONI anomaly


### Data Sets

- Monthly rainfall by province derived from CHIRPS daily + FAO GAUL admin 1
- [Monthly ONI from Climate Prediction Center (CPC) NOAA](https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/)

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(sf)
library(ggiraph)
library(glue)
library(gghdx)
library(here)
library(patchwork)

source(here("R/utils.R"))


gghdx()

fps <- proj_paths() # convenience func to load paths

# load gaul geodf/sf
gdf_adm1_global <- sf::st_read(fps$ADM1_GAUL,
                        "gaul1_asap",quiet = T)

# filter to afg
gdf_adm1 <-  gdf_adm1_global |> 
  filter(name0=="Afghanistan")

# load montly oni data
df_oni <- readr::read_table(fps$ONI_LINK) |> 
  clean_names()

# load monthly chirps data (downloaded from GEE)
df_chirps <- read_csv(fps$ADM1_CHIRPS_MONTHLY) |> 
  clean_names()

```


```{r}
#| echo: false
#| include: false
#| eval: false

# Here i interactively map and define groups for admin 1 regions to easy # visualizatons/plotting
# this is a quick way i could quickly group them based on geography  -- i save the output w/ datapasta as a hard coded table to be joined to the chirps data. The hardcoded table is saved in the next chunk so that we can repeat the analysis/visualizations without this interactive step. Saving the interactive step as it's a useful/cool trick someone (maybe me!?) might like to borrow in the future.

nw_grp <-mapedit::selectFeatures(x = gdf_adm1)
sw_grp <-  mapedit::selectFeatures(x = gdf_adm1 |> 
                                     filter(!name1 %in% nw_grp$name1))

ne_grp <-  mapedit::selectFeatures(x = gdf_adm1 |> 
                                     filter(!name1 %in% 
                                              c(sw_grp$name1,nw_grp$name1)))
central_grp <-  mapedit::selectFeatures(x = gdf_adm1 |> 
                                     filter(!name1 %in% 
                                              c(sw_grp$name1,
                                                nw_grp$name1,
                                                ne_grp$name1)))

lst(nw_grp,sw_grp,ne_grp,central_grp) |> 
  imap(\(gdft,nm_tmp){
    gdft |> 
      st_drop_geometry() |> 
      select(name1) |> 
      mutate(
        grp = nm_tmp
      )
  }
  ) |> 
  list_rbind() |> 
  datapasta::tribble_paste()
```
  
```{r grpLookup}
# data pasta output from above

adm1_grp_lookup<- tibble::tribble(
           ~name1,          ~grp,
          "Farah",      "nw_grp",
          "Hirat",      "nw_grp",
        "Badghis",      "nw_grp",
         "Faryab",      "nw_grp",
        "Jawzjan",      "nw_grp",
          "Balkh",      "nw_grp",
      "Sar-e-Pul",      "nw_grp",
       "Samangan",      "nw_grp",
           "Ghor",      "nw_grp",
       "Kandahar",      "sw_grp",
        "Hilmand",      "sw_grp",
         "Nimroz",      "sw_grp",
          "Zabul",      "sw_grp",
       "Daykundi",      "sw_grp",
        "Uruzgan",      "sw_grp",
     "Badakhshan",      "ne_grp",
         "Takhar",      "ne_grp",
         "Kunduz",      "ne_grp",
        "Baghlan",      "ne_grp",
       "Nuristan",      "ne_grp",
       "Panjsher",      "ne_grp",
        "Paktika", "central_grp",
         "Ghazni", "central_grp",
          "Khost", "central_grp",
         "Paktya", "central_grp",
          "Logar", "central_grp",
          "Kabul", "central_grp",
         "Kapisa", "central_grp",
        "Laghman", "central_grp",
      "Nangarhar", "central_grp",
          "Kunar", "central_grp",
         "Wardak", "central_grp",
         "Parwan", "central_grp",
         "Bamyan", "central_grp"
     )
```

```{r wrangleChirpsONI}

# wrangle ONI & chirps
df_oni <- df_oni |> 
  mutate(
    # create propoer date col
    date= ym(paste0(yr,".",mon))
  )

# df_processed 
df_proc <- df_chirps |> 
  group_by(adm1_name,mo= month(date)) |> 
  mutate(
    # creat absolute anomaly
    rainfall_abs_anom = value - mean(value, na.rm=T),
    # as well as one divide by SD
    rainfall_s_anom = rainfall_abs_anom/sd(value,na.rm=T)
  ) |> 
  ungroup() |> 
  left_join(adm1_grp_lookup, by =c("adm1_name"="name1")) |> 
  left_join(df_oni) |> 
  mutate(
    enso_class= case_when(anom<=-.5~"La Nina",
                          anom< 0.5~"Neutral",
                          anom>= 0.5 ~"El Nino")
  ) 

```

## Rainfall By Month

Below I've (somewhat arbitrarily) split the provinces in to larger regional groups and plotted the average monthly rainfall by province and month. As Hirat, is our initial Area of Interest (AOI), I've highlighted it in the plot.

```{r plotSeasonl}
df_proc |> 
  group_by(grp,adm1_name, mo = month(date,label =T, abbr=T)) |> 
  summarise(
    rainfall = mean(value,na.rm=T)
  ) |> 
  mutate(
    prov_label = case_when(
      adm1_name =="Hirat"~adm1_name,
      TRUE~"other"
    ),
    grp_label = 
      case_when(
        str_detect(grp,"ne")~"North East",
        str_detect(grp,"nw")~"North West",
        str_detect(grp,"sw")~"South West",
        str_detect(grp,"central")~"Central"
      )
  ) |> 
  ggplot(
    aes(x=mo, y=rainfall, 
        group=adm1_name, 
        color = prov_label)
  )+
  geom_line(aes(alpha= prov_label, 
                color = prov_label))+
   scale_alpha_manual(values=c(1,0.1))+  
  scale_color_manual(values=c("red","black"))+
  facet_wrap(~grp_label)+
  labs(title= "Monthly Average Rainfall By Province",
       subtitle= "Afghanistan",
       y= "rainfall (mm)")+
  theme(
    # legend.position = "none",
    legend.title= element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

```{r}
df_hirat <-  
  df_proc |> 
  filter(
    adm1_name=="Hirat"
  )
```


```{r}
#| echo: false
#| eval: false
#| include: false





p_hirat <-
  df_hirat |> 
   ggplot(
    aes(x= date, 
        y= value
        )
  )+
  geom_point_interactive(
    aes(color=enso_class,
        tooltip=glue("date: {date}
                     oni score: {anom},
                     enso class: {enso_class}")
        )
    
    )+
  geom_line(alpha=0.3)+
  theme(
    
  )

girafe(ggobj = p_hirat)

```

## Rainfall & ONI

### Overall

- Let's focus on `Hirat` to simplify the research questions and visuals.

- Below we look at absolute rainfall anomaly and ONI. When we look at all months no clear relationships pop out.

```{r}
p_scatter_hirat <- df_hirat |> 
  ggplot(
    aes(x= anom, y= rainfall_abs_anom)
  )+
  geom_point(alpha= 0.5)+
  labs(
    x = "ENSO ONI Anomaly",
    y= "Rainfall Anomaly (mm)",
    title = "Monthly ONI anomaly and rainfall not strongly correlated",
    subtitle= "Hirat Afghanistan",
    caption = "Average total monthly rainfall anomaly across Hirat province againt global monthly ONI anomaly (1981-2024)"
    )+
  scale_y_continuous()+
  # expand_limits (y= c(0,0))+
  theme(
    plot.title = element_text(size=10),
    plot.subtitle  = element_text(size=10),
    axis.title = element_text(size=8),
    axis.text= element_text(size=8),
    plot.caption = element_text(hjust= 0, size=8)
  )

p_boxplot_hirat <- df_hirat |> 
  ggplot(
    aes(x= enso_class, y= rainfall_abs_anom)
  )+
  geom_boxplot()+
  labs(
    y= "Rainfall Anomaly (mm)",
    title = "Monthly rainfall anomaly not well differentiated by ENSO class",
    subtitle= "Hirat Afghanistan",
    caption = "Average total monthly rainfall anomaly by ONI scores classified an ONI score of:
  - ≥ 0.5 -> El Nino
  - <0.5 to -0.5 -> Neutral
  - ≤ -0.5 -> La Nina"
  )+
  scale_y_continuous()+
  # expand_limits (y= c(0,0))+
  theme(
    plot.title = element_text(size=10),
    plot.subtitle  = element_text(size=10),
    axis.title = element_text(size=8),
    axis.text= element_text(size=8),
    axis.title.x= element_blank(),
    plot.caption = element_text(hjust= 0, size=8)
  )
  
```

```{r}
p_scatter_hirat + p_boxplot_hirat
```

### Winter wheat growing season

- Perhaps different seasons/months have different relationships (stronger/less strong) and this is adding noise.
- A this stage of the project I believe the main season of interest is the **Winter Wheat Season**.
  + [FAO Crop Calendar](https://www.fao.org/giews/countrybrief/country.jsp?code=AFG&lang=ar#:~:text=Harvesting%20of%20the%202023%20winter,conditions%20in%202021%20and%202022) shows that this growing season is Dec-April (inclusive). Therefore let's filter down to these months.
  
```{r}

# let's just copy the code above and filter by to the winter wheat growing season and change plot title

df_hirat_ww <- df_hirat |> 
  filter(month(date)%in% c(12,1,2,3,4))

p_scatter_hirat_ww <- df_hirat_ww |> 
  ggplot(
    aes(x= anom, y= rainfall_abs_anom)
  )+
  geom_point(alpha= 0.5)+
  labs(
    x = "ENSO ONI Anomaly",
    y= "Rainfall Anomaly (mm)",
    title = "Monthly ONI anomaly and rainfall not strongly correlated",
    subtitle= "Hirat Afghanistan - Winter Wheat Growing Season",
    caption = "Average total monthly rainfall anomaly across Hirat province againt global monthly ONI anomaly (1981-2024)"
    )+
  scale_y_continuous()+
  # expand_limits (y= c(0,0))+
  theme(
    plot.title = element_text(size=10),
    plot.subtitle  = element_text(size=10),
    axis.title = element_text(size=8),
    axis.text= element_text(size=8),
    plot.caption = element_text(hjust= 0, size=8)
  )

p_boxplot_hirat_ww <- df_hirat_ww |> 
  ggplot(
    aes(x= enso_class, y= rainfall_abs_anom)
  )+
  geom_boxplot()+
  labs(
    y= "Rainfall Anomaly (mm)",
    title = "Monthly rainfall anomaly not well differentiated by ENSO class",
    subtitle= "Hirat Afghanistan - Winter Wheat Growing Season",
    caption = "Average total monthly rainfall anomaly by ONI scores classified an ONI score of:
  - ≥ 0.5 -> El Nino
  - <0.5 to -0.5 -> Neutral
  - ≤ -0.5 -> La Nina"
  )+
  scale_y_continuous()+
  # expand_limits (y= c(0,0))+
  theme(
    plot.title = element_text(size=10),
    plot.subtitle  = element_text(size=10),
    axis.title = element_text(size=8),
    axis.text= element_text(size=8),
    axis.title.x= element_blank(),
    plot.caption = element_text(hjust= 0, size=8)
  )
  
```

- Plots are still relatively similar after filtering

```{r}
p_scatter_hirat_ww + p_boxplot_hirat_ww
```

- Instead of looking at monthly values lets' look at the whole season as one value (anomaly and rainfall)
- So we take the average ONI and sum of rainfall for all seasons Dec-April

```{r}
#| include: true
#| eval: true
#| echo: true

df_hirat_ww_seas <- df_hirat_ww |> 
  select(date,adm1_name,value, contains("rainfall"),"anom",everything()) |> 
  group_by(
    adm1_name,
  ) |> 
  mutate(
    mean_oni = zoo::rollmeanr(anom,4,fill=NA),
    rainfall = zoo::rollsumr(value,4,fill=NA),
  ) |> 
  filter(
    month(date)==4
  ) |> 
  group_by(
    adm1_name
  ) |> 
  mutate(
    rainfall_abs_anom = rainfall - mean(rainfall, na.rm=TRUE),
    rainfall_sd_anom = rainfall_abs_anom/sd(rainfall,na.rm=T),
    enso_class= case_when(
      mean_oni >= 0.5 ~ "El Nino",
      mean_oni < 0.5 & mean_oni > -0.5 ~ "Neutral",
      mean_oni <= -0.5 ~ "La Nina"
    )
    
  ) |> 
  ungroup()


pal_enso <-  c("La Nina"= "#FBB4AE" ,
                 "El Nino" = "#B3CDE3" ,
                 "Neutral" ="#CCEBC5"
)

df_hirat_ww_seas |> 
   ggplot(
    aes(x= mean_oni, y= rainfall_abs_anom, color=enso_class)
  )+
  geom_point(alpha= 1)+
  scale_color_manual(values = pal_enso)+
  labs(
    x = "ENSO ONI Anomaly",
    y= "Rainfall Anomaly (mm)",
    title = "Monthly ONI anomaly and rainfall not strongly correlated",
    subtitle= "Hirat Afghanistan - Winter Wheat Growing Season",
    caption = "Average rainfall anomaly across Hirat winter wheat growing season (Dec-April) plotted against global ONI anomaly averaged for each season (1981-2024)"
    )+
  scale_y_continuous()+
  # expand_limits (y= c(0,0))+
  theme(
    plot.title = element_text(size=10),
    plot.subtitle  = element_text(size=10),
    axis.title = element_text(size=8),
    axis.text= element_text(size=8),
    plot.caption = element_text(hjust= 0, size=8),
    legend.title = element_blank()
  )
  
```




```{r}
#| eval: false
#| include: false

# DESCRIPTION: Warm (red) and cold (blue) periods based on a threshold of +/- 0.5oC for the Oceanic Niño Index (ONI) [3 month running mean of ERSST.v5 SST anomalies in the Niño 3.4 region (5oN-5oS, 120o-170oW)], based on centered 30-year base periods updated every 5 years.
# since El Nino seems 
# https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php

# looks good.
df_oni |> 
  mutate(
    oni_mean3= zoo::rollmeanr(anom,3,fill=NA)
  )

df_oni |> 
  ggplot(
    aes(x= date, y=anom)
  )+
  geom_point()+
  geom_line()+
  scale_x_date(date_breaks= "1 year",date_labels = "%y")+
  theme(axis.text.x = element_text(angle = 90))

```

```{r}


```


