---
title: "SEAS5 MAM Aggregation: R vs Python Comparison"
format: html
---

Verify that the Python monitoring pipeline (`w1_monitor_seas5.py`) produces
identical regional MAM aggregates to the R pipeline used for threshold
derivation (`data-raw/16_prepare_2026_model_feature_datasets.R`).

## Setup

```{r}
#| label: setup-r
library(dplyr)
library(lubridate)
library(cumulus)
library(reticulate)
library(ggplot2)
# Point reticulate at the uv-managed venv
use_virtualenv(here::here(".venv"), required = TRUE)
```

```{python}
#| label: setup-python
import sys
from pathlib import Path

# ensure project root is on path for imports
sys.path.insert(0, str(Path.cwd()))

from calendar import monthrange

import numpy as np
import ocha_stratus as stratus
import pandas as pd
from sqlalchemy import text
```

## Constants

```{r}
#| label: constants
PROVINCES_AOI <- c("Faryab", "Sar-e-Pul", "Jawzjan", "Balkh", "Badghis")
```

## R Pipeline

Replicate the exact logic from `data-raw/16_prepare_2026_model_feature_datasets.R`:

1. Load SEAS5 from DB with `cumulus::pg_load_seas5_historical()` (which applies `mean * days_in_month`)
2. Aggregate MAM months with `cumulus::seas5_aggregate_forecast()`
3. Join area weights from ERA5 blob
4. Compute `weighted.mean()` across provinces

```{r}
#| label: r-seas5-load
# Load SEAS5 with unit conversion (daily rate -> monthly mm)
df_seas5 <- pg_load_seas5_historical(
  iso3 = "AFG",
  adm_name = PROVINCES_AOI,
  adm_level = 1,
  convert_units = TRUE
)

# Aggregate to MAM total per province per issued_date (March only)
df_seas5_mam <- seas5_aggregate_forecast(
  df = df_seas5,
  value = "mean",
  valid_months = c(3, 4, 5),
  by = c("iso3", "pcode", "name", "issued_date")
) |>
  filter(month(issued_date) == 3) |>
  rename(adm1_name = name, mam_mm = mean)

glimpse(df_seas5_mam)
```

```{r}
#| label: r-area-weights
# Load area weights from ERA5 blob (same source as Python pipeline)
df_era5 <- blob_read(
  name = "ds-aa-afg-drought/raw/vector/historical_era5_land_ndjfmam_lte2025.parquet",
  stage = "dev",
  container_name = "projects"
) |>
  janitor::clean_names()

df_area_lookup <- df_era5 |>
  filter(adm1_name %in% PROVINCES_AOI) |>
  distinct(adm1_name, shape_area)

df_area_lookup
```

```{r}
#| label: r-regional-aggregate
# Area-weighted regional mean per year
df_r_regional <- df_seas5_mam |>
  left_join(df_area_lookup, by = "adm1_name") |>
  group_by(issued_date) |>
  summarise(
    r_mam_mm = weighted.mean(mam_mm, w = shape_area, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(year = year(issued_date)) |>
  filter(year >= 1984) |>
  select(year, r_mam_mm) |>
  arrange(year)

df_r_regional
```

## Python Pipeline

Same logic as `w1_monitor_seas5.py`:

1. Query SEAS5 from prod DB
2. `mean * days_in_month` -> monthly mm
3. Sum MAM per province
4. `np.average` with `shape_area` weights

```{python}
#| label: py-load-data
# Load area weights
df_areas = stratus.load_parquet_from_blob(
    blob_name="ds-aa-afg-drought/raw/vector/historical_era5_land_ndjfmam_lte2025.parquet",
    stage="dev",
    container_name="projects",
)

df_areas.columns = df_areas.columns.str.lower().str.replace(" ", "_")
df_areas = df_areas.loc[df_areas["adm1_name"].isin(
    ["Faryab", "Sar-e-Pul", "Jawzjan", "Balkh", "Badghis"]
)][["adm1_name", "shape_area"]].drop_duplicates()

# Load admin lookup
df_lookup = stratus.load_parquet_from_blob(
    blob_name="admin_lookup.parquet",
    stage="dev",
    container_name="polygon",
)
df_lookup.columns = df_lookup.columns.str.lower()
df_lookup = df_lookup.loc[
    (df_lookup["iso3"] == "AFG")
    & (df_lookup["adm_level"] == 1)
    & (df_lookup["adm1_name"].isin(
        ["Faryab", "Sar-e-Pul", "Jawzjan", "Balkh", "Badghis"]
    )),
    ["adm1_name", "adm1_pcode"],
].drop_duplicates()
pcodes = df_lookup["adm1_pcode"].tolist()
print(f"Pcodes: {pcodes}")
print(f"Area weights:\n{df_areas}")
```

```{python}
#| label: py-query-aggregate
# Query SEAS5 from prod
engine = stratus.get_engine(stage="prod")
pcode_list = ", ".join(f"'{p}'" for p in pcodes)
sql = text(f"""
    SELECT iso3, pcode, valid_date, issued_date, leadtime, mean
    FROM seas5
    WHERE iso3 = 'AFG'
      AND adm_level = 1
      AND pcode IN ({pcode_list})
      AND EXTRACT(MONTH FROM issued_date) = 3
      AND EXTRACT(MONTH FROM valid_date) IN (3, 4, 5)
    ORDER BY issued_date, pcode, valid_date
""")
with engine.connect() as conn:
    df = pd.read_sql(sql, conn)

# daily rate -> monthly mm
df["days"] = df["valid_date"].apply(lambda d: monthrange(d.year, d.month)[1])
df["precip_mm"] = df["mean"] * df["days"]

# MAM sum per province
df_province = (
    df.groupby(["issued_date", "pcode"], as_index=False)["precip_mm"]
    .sum()
    .rename(columns={"precip_mm": "mam_mm"})
)
df_province = df_province.merge(
    df_lookup.rename(columns={"adm1_pcode": "pcode"}), on="pcode", how="left"
)

# area-weighted regional mean
df_province = df_province.merge(df_areas, on="adm1_name", how="left")

def _weighted_mean(g):
    return np.average(g["mam_mm"], weights=g["shape_area"])

df_py_regional = (
    df_province.groupby("issued_date", as_index=False)
    .apply(_weighted_mean, include_groups=False)
    .rename(columns={None: "py_mam_mm"})
)
df_py_regional["year"] = pd.to_datetime(df_py_regional["issued_date"]).dt.year
df_py_regional = (
    df_py_regional.loc[df_py_regional["year"] >= 1984][["year", "py_mam_mm"]]
    .sort_values("year")
    .reset_index(drop=True)
)
print(df_py_regional)
```

## Comparison

```{r}
#| label: compare
# Pull Python results into R via reticulate
df_py <- py$df_py_regional

# Merge
df_compare <- df_r_regional |>
  inner_join(df_py, by = "year") |>
  mutate(
    diff_mm = r_mam_mm - py_mam_mm,
    diff_pct = diff_mm / r_mam_mm * 100
  )

cat("Max absolute difference (mm):", max(abs(df_compare$diff_mm)), "\n")
cat("Max relative difference (%):", max(abs(df_compare$diff_pct)), "\n")
cat("Number of years compared:", nrow(df_compare), "\n")
```

```{r}
#| label: compare-table
df_compare |>
  select(year, r_mam_mm, py_mam_mm, diff_mm) |>
  mutate(across(where(is.numeric) & !matches("year"), ~round(.x, 4))) |>
  knitr::kable(
    col.names = c("Year", "R (mm)", "Python (mm)", "Diff (mm)"),
    caption = "SEAS5 MAM regional aggregate: R vs Python"
  )
```

```{r}
#| label: compare-plot
#| fig-width: 8
#| fig-height: 4


ggplot(df_compare, aes(x = r_mam_mm, y = py_mam_mm)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.7) +
  labs(
    x = "R pipeline (mm)",
    y = "Python pipeline (mm)",
    title = "SEAS5 MAM Regional Aggregate: R vs Python",
    subtitle = "Points on the dashed line = perfect agreement"
  ) +
  theme_minimal()
```
