---
title: AOI Selection - Prioritization
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  include: true
  echo: true
  warning: false
  message: false
  eval: true
  results: "asis"
  out.width: "100%"
  code-fold: true
editor: visual
project:
  execute-dir: project
---

## Intro

Preliminary **work**

This document contains exploratory work done to support area targeting for drought monitoring and AA activity implementation. Initially the discussion centered around drought monitoring over the winter wheat agricultural season in Afghanistan, later this was revised to the Spring wheat growing season

While targeting decisions should ultimately be made by in-country experts and implementing partners to reflect operational realities, this document aims to explore several additional data sets to support this process and encourage discussion.

The document focuses on 3 main data sets:

1.  Winter wheat/Spring wheat crops mask (TBD)
2.  Rainfall (CHIRPS)
3.  Food Insecurity (IPC Phase Classification)

### Key Takeaways

1.  To date no winter/spring wheat crop mask provided. Therefore, the identification of key wheat growing areas has been left to in-country experts
2.  Rainfall analysis provides useful clustering of provinces based on historical scarcity in alignment with general feedback from in-country partners
3.  Up to date IPC Phase Classifications have not shown to be useful for prioritization due to homogeneity of the latest classification which offers little discriminatory power for this project

Working group feedback:

1.  The initial group decision was to focus the pilot on `Faryab` province in the north
2.  As of December 2024, the group has tentatively decided to expand to include `Faryab` ,`Takhar`, `Badakhshan`, `Badghis`, `Sar-e-Pul` provinces.

```{r libsData}
box::use(dplyr[...])
box::use(purrr[...])
box::use(stringr[...])
box::use(forcats[...])
box::use(lubridate[...])
box::use(ggplot2[...])
box::use(ggfx[...])
box::use(gghdx[...])
box::use(patchwork[...])
box::use(terra[...])
box::use(sf[...])
box::use(janitor[...])

box::use(ggiraph)
box::use(ripc)
box::use(cumulus)

box::use(blob = ../R/blob_connect)

gghdx()

# load admin boundaries

gdf_adm1 <- cumulus$download_fieldmaps_sf(iso3= "AFG",layer= "afg_adm1")$afg_adm1 |>
  clean_names() |>
  select(matches("adm\\d_[pe]"))


gdf_adm2 <- cumulus$download_fieldmaps_sf(iso3= "AFG",layer= "afg_adm2")$afg_adm2 |>
  clean_names() |>
  select(matches("adm\\d_[pe]"))

# load chirps rainfall
fps <- blob$proj_blob_paths()

df_chirps_adm2 <- cumulus$blob_read(
  name = fps$DF_ADM2_CHIRPS_WFP,
  container ="projects",
  stage= "dev",
  progress_show =FALSE
)
```

## Winter Wheat

Data source: **TBD** (waiting for officially recommended winter wheat crop mask layer).

Idea: On basis that activities/framework will be centered around winter wheat agricultural season we should target areas where winter wheat is grown. If we have a crop mask we could analyze total area/% area growing winter wheat at an admin level and use that as an input for the targeting/prioritization analysis.

## Rainfall

Data source: CHIRPS

```{r weightedZonalStats}

# Get area of adm2 as % of adm1 for weighting
df_adm2_w_area <- gdf_adm2 %>%
  mutate(
    area = st_area(.)
  ) |>
  group_by(adm1_pcode, adm1_en) %>%
  mutate(
    adm1_area = sum(area)
  ) |>
  ungroup() |>
  mutate(
    pct_area = area / adm1_area
  ) |>
  st_drop_geometry()

# get monthly rainfall by admin 2
df_chirps_monthly_adm2 <- df_chirps_adm2 |>
  mutate(
    mo_date = floor_date(date,"month"),.before = "date"
  ) |>
  group_by(
    across(
      matches("adm\\d_[pe]")
    ),
    mo_date
  ) |>
  summarise(
    precip = sum(rfh)
  ) |>
  left_join(
    df_adm2_w_area
  )


# aggregate to admin 1
df_chirps_monthly_adm1 <- df_chirps_monthly_adm2 |>
  group_by(
    across(
      matches("adm[01]_[pe]")
    ),
    mo_date
  ) |>
  summarise(
    precip = weighted.mean(x= as.numeric(precip), w=as.numeric(pct_area)),
    .groups="drop"
  )


# aggregate to yearly
df_chirps_yearly_adm1 <- df_chirps_monthly_adm1 |>
  group_by(
    across(
      matches("adm\\d_[pe]")
    ),
    yr_date = floor_date(mo_date, "year")
  ) |>
  summarise(
    precip = sum(precip)
  )



baseline_years <- c(1981:2020)

# get average precip per year per admin using baseline
yearly_baseline_avg <- df_chirps_yearly_adm1 |>
  filter(
    year(yr_date) %in% baseline_years
  ) |>
  group_by(
    across(
      matches("adm\\d_[pe]")
    )
  ) |>
  summarise(
    avg_precip = mean(precip)
  )


# calculate anomaly
df_chirps_adm1_anomaly <- df_chirps_yearly_adm1 |>
  left_join(
    yearly_baseline_avg
  ) |>
  mutate(
    anom_abs = precip - avg_precip
  )



gdf_adm1_ranked <- gdf_adm1 |>
  left_join(
    df_chirps_adm1_anomaly |>
      filter(year(yr_date) %in% c(2020:2023)) |>
      group_by(
        yr_date
      ) |>
      arrange(
        yr_date,
        desc(-anom_abs)
      ) |>
      mutate(
        yr = as_factor(year(yr_date)),
        rank = row_number(),
        top5_dry = rank %in% c(1:7),
        top5_anom = if_else(top5_dry,anom_abs,NA_real_)
      )
  )


limit <- max(abs(gdf_adm1_ranked$top5_anom)) * c(-1, 1)

gdf_top5_diss <- gdf_adm1_ranked |>
  group_by(yr) |>
  filter(top5_dry) |>
  summarise()
```

```{r}
#| fig.height: 7
#| label: fig-plot_chirps_2020_2023

ggplot()+
  ggiraph$geom_sf_interactive(
    data= gdf_adm1_ranked,
    aes(
      fill = anom_abs,
      tooltip = adm1_en
        ),
    show.legend = c(color =FALSE, fill=TRUE)
  )+
  with_shadow(
    geom_sf(
      data = gdf_top5_diss,
      # aes(fill= diff),
      fill = NA,
      alpha = 1,
      color = "black",
      lwd = 0.7
    ),
    sigma = 3,
    x_offset = 0.5,
    y_offset = 0.25
  ) +
  scale_fill_gradient2(low = "#FD7446", high = "#709AE1") +
  facet_wrap(~yr)+
  labs(
    title = "Yearly rainfall anomaly",
    subtitle = "Driest 7 provinces highlighted",
    caption = "Data source: CHIRPS"
  )+
  theme(
    legend.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    axis.line.x.bottom = element_blank()
  )
```

```{r}
#| label: fig-plot_chirps_2020_2023_avg


gdf_adm1_ranked_overall<- gdf_adm1_ranked |>
  mutate(
    # these weights are subjective first go
    # rationale for weighting is that more recent years
    # should be weighted higher
    weights = case_when(
      yr ==2020~0.15,
      yr==2021~0.2,
      yr==2022~0.3,
      yr==2023~0.35
    )
  ) |>
  group_by(
    across(matches("adm\\d_[pe]"))
  ) |>
  summarise(
    weighted_anom = weighted.mean(x= anom_abs, w=weights)
  ) |>
  ungroup() |>
  arrange(
    desc(-weighted_anom)
  ) |>
  mutate(
    rank = row_number(),
    top5_dry = rank %in% c(1:7),
    top5_anom = if_else(top5_dry,weighted_anom,NA_real_)
  )

gdf_top5_overall_diss <- gdf_adm1_ranked_overall |>
  filter(top5_dry) |>
  summarise()


ggplot()+
  geom_sf(
    data= gdf_adm1_ranked_overall,
    aes(fill = weighted_anom),
    show.legend = c(color =FALSE, fill=TRUE)
  )+
  with_shadow(
    geom_sf(
      data = gdf_top5_overall_diss,
      # aes(fill= diff),
      fill = NA,
      alpha = 1,
      color = "black",
      lwd = 0.7
    ),
    sigma = 3,
    x_offset = 0.5,
    y_offset = 0.25
  ) +
  scale_fill_gradient2(low = "#FD7446", high = "#709AE1")+
  labs(
    title = "Average yearly rainfall anomaly (2020-2023)",
    subtitle = "Driest 7 provinces highlighted",
    caption = "Data source: CHIRPS"
  )+
  theme(
    legend.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    axis.line.x.bottom = element_blank()
  )

```

```{r chirpsBarplot}
#| echo: false
#| include: false
df_chirps_adm1_anomaly |>
  filter(year(yr_date) %in% c(2020:2023)) |>
  group_by(
    yr_date
  ) |>
  arrange(
    yr_date,
    desc(-anom_abs)
  ) |>
  mutate(
    yr = as_factor(year(yr_date)),
    rank = row_number()
  ) |>
  ggplot(
    aes(
      x= reorder(adm1_en,anom_abs),
      y= anom_abs,
      fill= rank
    )
  )+
  geom_bar(
    stat="identity"
  )+
  scale_fill_continuous(type = "viridis",direction =-1)+
  coord_flip()+
  facet_wrap(~yr, scales = "free")
```

## Food Insecurity (IPC)

Data source (IPC)

**preliminary analysis** - needs further validation

```{r}
# bugfix/pop-areas
ipc_get_areas_current_sf <- function(country){

  # get full list of data sets
  ipc_pop_area <- ripc$ipc_get_population(country = country)$area

  # grab area analysis ids
  area_analysis_ids <- unique(ipc_pop_area$analysis_id)

  gdf_ipc_area <- map(
    area_analysis_ids,
    \(id_tmp){

      gdf_area <- ripc$ipc_get_areas(
        id = id_tmp,
        period = "C",
        return_format = "geojson"
      )

      gdf_area_merged <- gdf_area |>
        left_join(
          ipc_pop_area |>
            mutate(
              analysis_id = as.character(analysis_id),
              area_id = as.character(area_id)
            ) |>
            filter(
              period == "current"
            ) |>
            select(
              analysis_id,
              area_id,
              analysis_period_start,
              analysis_period_end,
              analysis_date,
              title,
              period
            ),
          by = c("analysis_id","area_id"),
          relationship="one-to-one"
        )
      gdf_area_merged
    }
  )
  gdf_ipc_area |>
    list_rbind()

}

ipc_fill_pal <- c(
  "0" = "#FFFFFF",
  "1" = "#CDFACD",
  "2" = "#FAE61E",
  "3" = "#E67800",
  "4" = "#C80000",
  "5" = "#640000"
)
```

### Latest IPC classification

Below we plot the latest IPC phase classifications produced:

```{r}
#| label: fig-plot_ipc_latest

gdf_ipc_sf_current <- ipc_get_areas_current_sf(country = "AF")

latest_analysis_id <- gdf_ipc_sf_current |>
  filter(
    analysis_period_end == max(analysis_period_end)
  ) |>
  pull(
    analysis_id
  ) |>
  unique()



# could put warning if latest_analysis_id > 1 -- here

gdf_ipc_sf_projected_latest <- ripc$ipc_get_areas(
  id = latest_analysis_id,
  period = "P",
  return_format = "geojson"
)


gdf_ipc_sf_current_latest <- gdf_ipc_sf_current |>
  filter(
    analysis_id == latest_analysis_id
  )


gdf_ipc_sf_latest <- bind_rows(
  gdf_ipc_sf_current_latest,
  gdf_ipc_sf_projected_latest
) |>
  mutate(
    ipc_period_label = if_else(ipc_period =="C","Current","Projected")
  )


gdf_ipc_sf_latest_poly <- gdf_ipc_sf_latest |>
  filter(
    !(str_detect(st_geometry_type(geometry), "POINT") )
  ) |>
  st_as_sf() |>
  mutate(
    overall_phase = fct_relevel(fct_expand(as_factor(overall_phase),"0","1","2","3","4","5"),"0","1","2","3","4","5")
  )


p_title <- unique(gdf_ipc_sf_latest_poly$title)
p_title <- p_title[!is.na(p_title)]
ggplot()+
  geom_sf(
    data= gdf_ipc_sf_latest_poly,
    aes(
      fill = overall_phase,
    ),show.legend = c(fill=TRUE)
  )+
  scale_fill_manual(
    values = ipc_fill_pal,
    drop =FALSE
  )+
  facet_wrap(
    ~ipc_period_label
  )+
  labs(
    title = p_title,
    subtitle = "Afghanistan: latest available classifiations",
    caption = "Data source: IPC"
  )+
  guides(
    fill = guide_legend(
      label.position = "bottom",
      nrow=1, byrow=TRUE
    )
  )+
  theme(
    legend.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    axis.line.x.bottom = element_blank(),
    legend.key.spacing  = unit(0, 'cm'),
    legend.text = element_text(margin = margin(t = 3))
  )
```

### IPC Recent Years

Next we filter to last 5 years (2020-2024) to get an overview of how the situation has evolved recently

```{r ipc}

gdf_ipc_sf_current_polys <- gdf_ipc_sf_current |>
  filter(
    !(str_detect(st_geometry_type(geometry), "POINT") )
  ) |>
  st_as_sf()


gdf_areas_filt <- gdf_ipc_sf_current_polys |>
  mutate(
    analysis_period_start_year =year(analysis_period_start),
    analysis_period_end_year =year(analysis_period_end),
    start_end = paste0(
      format(analysis_period_start,"%e %b %y"),"-",
      format(analysis_period_end,"%e %b %y")
    )
  ) |>
  filter(
    year(analysis_period_start) %in% 2020:2024
  )

df_lookup_factor_levels <- gdf_areas_filt |>
  st_drop_geometry() |>
  distinct(
    analysis_period_start,
    start_end
  ) |>
  arrange(
    analysis_period_start
  )


# factorize start_end so that factors are in same order as analysis_period_start
gdf_areas_filt <- gdf_areas_filt |>
  mutate(
    start_end = factor(start_end, levels = df_lookup_factor_levels$start_end),

    overall_phase_factor = fct_relevel(fct_expand(as_factor(overall_phase),"0","1","2","3","4","5"),"0","1","2","3","4","5")
  )
```

```{r}
#| fig.height: 7
#| label: fig-plot_ipc_2020_current_snapshots

# map each time period
ggplot()+
  geom_sf(
    data= gdf_areas_filt,
    aes(
      fill = overall_phase_factor,
    ),show.legend = c(fill=TRUE)
  )+
  scale_fill_manual(
    values = ipc_fill_pal,
    drop =FALSE
  )+
  facet_wrap(
    ~start_end
  )+
  guides(
    fill = guide_legend(
      label.position = "bottom",
      nrow=1, byrow=TRUE
    )
  )+
  theme(
    legend.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    axis.line.x.bottom = element_blank(),
    legend.key.spacing  = unit(0, 'cm'),
    legend.text = element_text(margin = margin(t = 3))
  )

```

### Recent Years Aggregated

Finally we attempt to aggregate recent years by two methods as an attempt to try to understand where certain areas could be more vulnerable due to accumulated food insecurity stress over the recent years.

```{r}
#| fig.height: 8
#| label: fig-plot_ipc_weighted_mean

mode <- function(codes){
  which.max(tabulate(codes))
}

gdf_areas_filt_summarised<- gdf_areas_filt |>
  mutate(
    # go back to numeric
    overall_phase = as.numeric(overall_phase),
    yr = year(analysis_period_start),
    weights = case_when(
      yr == 2020~ 0.1,
      yr== 2021~ 0.15,
      yr== 2022~ 0.2,
      yr== 2023~ 0.25,
      yr==2024~0.3
    )
  )|>
  group_by(
    # area_id,
    area_name
  ) |>
  summarise(
    overall_phase_mean = mean(overall_phase,na.rm=T),
    overall_phase_median = median(overall_phase,na.rm=T),
    overall_phase_mode = mode(overall_phase),
    overall_phase_mean_weighted = weighted.mean(as.numeric(overall_phase),weights,na.rm=T),
    .groups="drop"
  ) |>
  mutate(
    overall_phase_mode = fct_relevel(fct_expand(as_factor(overall_phase_mode),"0","1","2","3","4","5"),"0","1","2","3","4","5")
  )



p_ipc_mode <- ggplot()+
  geom_sf(
    data= gdf_areas_filt_summarised,
    aes(
      fill = overall_phase_mode
    ),show.legend = c(fill=TRUE)
  )+
  scale_fill_manual(
    values = ipc_fill_pal,
    drop =FALSE
  )+
  guides(
    fill = guide_legend(
      label.position = "bottom",
      nrow=1, byrow=TRUE
    )
  )+
  labs(
    subtitle= "Mode"
  )+
  theme(
    legend.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    axis.line.x.bottom = element_blank(),
    legend.key.spacing  = unit(0, 'cm'),
    legend.text = element_text(margin = margin(t = 3))
  )


p_ipc_weighted_mean <- ggplot()+
  geom_sf(
    data= gdf_areas_filt_summarised,
    aes(
      fill = overall_phase_mean_weighted
    ),
    show.legend = c(fill=TRUE)
  )+
  scale_fill_gradient(
    low = "white",high="red"
    ) +
  guides(
    fill = guide_legend(
      label.position = "bottom",
      nrow=1, byrow=TRUE
    )
  )+
  labs(
    subtitle = "Weighted Mean"
  )+
  theme(
    legend.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    axis.line.x.bottom = element_blank(),
    legend.key.spacing  = unit(0, 'cm'),
    legend.text = element_text(margin = margin(t = 3))
  )



p_ipc_mode +
  p_ipc_weighted_mean+
  plot_layout(
    nrow= 2, ncol =1
  )+
  plot_annotation(
    title = "IPC Phase Aggregated 2020-2024",
  )
```
