---
title: Snow Cover/Water Equivalent
subtitle: Initial Analysis - NDSI & SWE
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  include: true
  echo: true
  warning: false
  message: false
  eval: true
  results: "asis"
  out.width: "100%"
  code-fold: true
editor: visual
project:
  execute-dir: project
---

**WIP**

## Intro

In code below we try to assess if monthly remotely sensed snow indicators have predictive power on vegetative indicators Agriculture Stress Index (ASI) and Vegetative Health Index (FAO) over the March-April-May (MAM) planting season in Afghanistan.

**Data Sources**

-   Normalized Difference Snow Index (NDSI): MODIS Terra Satellite
-   Snow Water Equivalent (SWE): Famine Early Warning Systems Network (FEWS NET) Land Data Assimilation System (FLDAS)
-   Agricultural Stress Index (ASI): FAO
-   Vegetative Health Index (VHI): FAO

## Results

-   These results concur with the previous [analysis](https://docs.google.com/presentation/d/1pfqpEGx-MB_-1A8PiCnDmcOwifUQ-EXyTDuqQyBiw1c/edit?pli=1#slide=id.g2e1f646d655_0_126) of Faryab province that indicated a low correlation/predictive between driest SWE years and least healthy ASI years within
-   Feb-Mar snow indicators (SWE & NDSI) show a more intuitive relationship with ASI than Dec-Jan snow indicators. This could indicate that later season hydro-dynamics play a more important role in agricultural MAM drought than the actual the total quantity of snow detected in the heart of the winter months (D-J).
-   SWE & NDSI are moderately correlated with each other as expected

## Discussion & Next Steps

-   Simple N-D-J aggregated snow indicators (NDSI & SWE) do not appear to be useful in trigger design.
-   To confirm or support the importance of late-season hydro-dynamics we may consider further integrating snow-melt dynamics into the analysis as started [output here](https://rpubs.com/zackarno/snow_melt_date_vs_asi)

```{r}
strip_text_size <- 8
pt_size <- 0.5
box::use(
  ../R/blob_connect,
  loaders =../R/load_funcs[...]
)

box::use(
  dplyr[...],
  forcats[...],
  ggplot2[...],
  gghdx[...],
  janitor[...],
  lubridate[...],
  purrr[...],
  readr[...],
  scales[...],
  stringr[...],
  glue[...],
  tidyr[...],
  scales
)
gghdx()


df_env <- loaders$load_cleaned_env_features(mo= c(11,12,1,2,3,4,5,6),include_cumulative = F)


df_fao <- df_env |>
  filter(parameter %in% c("asi","vhi"))

df_snow <- df_env |>
  filter(parameter %in% c("NDSI_Snow_Cover_mean","SWE_inst","era5_land_snow_cover","era5_land_snow_depth_water_equivalent")) |>
   mutate(
    type = case_when(
      parameter == "NDSI_Snow_Cover_mean"~"NDSI",
      parameter == "SWE_inst"~"SWE",
      parameter == "era5_land_snow_cover" ~"Snow Cover ERA",
      parameter == "era5_land_snow_depth_water_equivalent" ~"SDWE ERA",
      .default = NA
      )
  )

```

## Snow Indicators

### Overview

Here we see the monthly distributions of snow indicators `NDSI` & `SWE` based on historical data.

```{r}
#| label: fig-plot_snow_overview

df_snow |>
  filter(adm1_name == "Faryab", type %in% c("NDSI","SWE")) |>
  mutate(
    mo = month(date, label = T, abbr=T)
  ) |>

  mutate(
    mo_reorder = factor(mo,levels=month(c(11:12,1:10), label =T, abbr= T))
  ) |>
  ggplot(
    aes(x= mo_reorder, y= value, fill =type)
  )+
  geom_boxplot()+
  facet_wrap(~adm1_name)+
  theme(
    axis.text.x = element_text(angle = 90),legend.title = element_blank(),
    axis.title.x = element_blank()
  )

```

```{r}

df_snow_proc <- df_snow |>
  group_by(
    parameter,
    adm1_name,
    mo = month(date)
  ) |>
  mutate(
    anom = value - mean(value) ,
    anom_z = anom /sd(value)
  ) |>
  group_by(adm1_name) |>
  arrange(adm1_name,date) |>
  mutate(
    value_change= value -lag(value)
  ) |>
  ungroup()
```

### Comparison

sense check that NDSI and SWE are somewhat aligned. They seem reasonably so - giving confidence

```{r plot-scatterSnow}
#| fig.height: 10
#| label: fig-plot_ndsi_vs_swe

df_snow_proc_adj <- df_snow_proc |>
  mutate(
    yr = year(date),
    yr_adj = ifelse(mo>7, yr+1, yr)
  )

# gut check that SWE & NDSI roughly line up
df_snow_proc_adj |>
    pivot_wider(
    id_cols = c("date","adm1_name"),
    values_from = value,
    names_from = type
  ) |>
  ggplot(
    aes(
      x = NDSI,
      y= SWE,
    )
  )+
  geom_point(
    size = pt_size
  )+
  facet_wrap(~adm1_name, scales = "free")+
  labs(
    title = "NDSI vs SWE"
  )+
  theme(
    strip.text = element_text(size = 10)
  )
```

## ASI & VHI

```{r}

df_fao_yr_mo <- df_fao |>
  group_by(
    type= parameter,
    adm1_name,
    mo_date = floor_date(date,"month"),
  ) |>
  summarise(
    month_mean = mean(value),
    last_dekad = last(value)
  ) |>
  ungroup()

```

```{r}
#| eval: false
#| echo: false
#| include: false
#| label: fig-plot_asi_vhi_timeseries

# plot of ASI & VHI by month - could be another chapter

df_fao_yr_mo |>
  group_by(
      type,
    adm1_name,
    mo = month(mo_date,label =T, abbr= T)
  ) |>
  mutate(
    month_mean = ifelse(type == "asi",month_mean/100,month_mean)
  ) |>
  summarise(
    across(.cols = c("month_mean","last_dekad"),\(x)mean(x,na.rm=T))
  ) |>
  ggplot(
    aes(x= mo, y= month_mean, color = type, group = type)
  )+
  geom_point()+
  geom_line()+
  facet_wrap(~adm1_name)+
  theme(
    axis.text.x = element_text(angle =90)
  )

```

```{r}

df_fao_yr_mo_anoms <- df_fao_yr_mo |>
  mutate(
    yr_date= floor_date(mo_date, "year")
  ) |>
  group_by(
    type,
    adm1_name,
    mo= month(mo_date)
  ) |>
  mutate(
    mo_mean_anom =  month_mean - mean(month_mean),
    mo_mean_anom_z = mo_mean_anom/sd(month_mean),
    mo_end_anom = last_dekad - mean(last_dekad),
    mo_end_anom_z = mo_end_anom/sd(mo_end_anom)
  ) |>
  ungroup()

mam_int <- c(3:5)
mamj_int <- c(3:6)
df_fao_yr_mo_anoms |> glimpse()
df_fao_yr_mo_anoms |>
  filter(is.na(last_dekad))

df_fao_yr <- df_fao_yr_mo_anoms |>
  group_by(
    type,
    adm1_name,
    yr_date,
    yr= year(yr_date)
  ) |>
  summarise(
    mam_mean_value = mean(month_mean[mo %in% mam_int]),
    mamj_mean_value = mean(month_mean[mo %in% mamj_int]),
    may_last_value = mean(last_dekad[mo==5]),
    june_last_value = mean(last_dekad[mo==6]),
    mam_mean_anom = mean(mo_mean_anom[mo %in% mam_int]),
    mamj_mean_anom = mean(mo_mean_anom[mo %in% mamj_int]),
    mam_mean_z = mean(mo_mean_anom_z[mo %in% mam_int]),
    mamj_mean_z = mean(mo_mean_anom_z[mo %in% mamj_int]),
    may_last_value = mean(last_dekad[mo==5]),
    june_last_value = mean(last_dekad[mo==6]),
  )
```

quick check to make sure VHI and ASI are reasonably aligned. Here is a scatter plot of the two indicators for all provinces. It makes sense

```{r}
#| label: fig-plot_asi_vhi_scatter

# quick comparison of VHI vs ASI - makes sense
df_fao_yr |>
  pivot_wider(
    id_cols = c(
      "adm1_name","yr_date"),
    names_from = "type", values_from = "mam_mean_value"
              ) |>
  ggplot(
    aes(x= vhi, y= asi)
  )+
  geom_point()+
  labs(
    title = "VHI vs ASI (average MAM value)",
    subtitle = "Afghanistan - All provinces"
  )

```

## Snow vs Vegetation

```{r}
#' this section is a bit overly complicated as I wanted to run the correlations for all combinations of engineered features for ASI, VHI, NDSI, SWE.
# In the end I only show a few of the resulting plots, but I think the rest may be nice to have at some point - there are tons!!!


df_snow_ready <- df_snow_proc_adj |>
  mutate(
    mo_label = month(mo,label=T, abbr=T),
    snow_type = type
  ) |>
  select(-yr) |>
  rename(
  yr= yr_adj
  )


merge_snow_veg <- function(
    df_snow,
    snow_param,
    df_veg,
    veg_param,
    start_year = 2000
    ){


  inclusion_range<- df_veg |>
    pull(yr) |>
    range()
  if(is.null(start_year)){
    start_year <- inclusion_range[1]
  }

  snow_filt <- df_snow |>
    filter(
      snow_type == snow_param
    ) |>
     filter(yr>=start_year, yr<=inclusion_range[2]) |>
      pivot_wider(id_cols = c("adm1_name",
                              "yr"), # pivot so 1 row per year
                      names_from = mo_label,
                      values_from = c("value","anom","anom_z","value_change"),names_prefix ="snow_"
          )
   veg_filt <-  df_veg |>
    filter(
      type == veg_param
    )
  left_join(snow_filt,veg_filt,by=c("adm1_name",
                                    "yr"))

}


corr_all_snow <- function(df_snow=df_snow_ready,
                          snow_param = "NDSI",
                          snow_feature="value",
                          df_veg=df_fao_yr,
                          veg_param = "asi",
                          fao_feature ="mam_mean_value",
                          start_year = 2000
                          ){
  mo_labs <- month(1:12, label = T, abbr= T)
  mo_rgx <- glue_collapse(mo_labs,sep = "|")
  p_title = glue("Snow ({snow_param}: {snow_feature}) vs Vegetation ({toupper(veg_param)}: {fao_feature})")

  df_merged <- merge_snow_veg(
    df_snow =df_snow,
    snow_param = snow_param,
    df_veg = df_veg,
    veg_param = veg_param,
    start_year = start_year
  ) |>
    filter(yr<2025)

  df_corrs <- df_merged |>
    group_by(
      type,
      adm1_name,
    ) |>
    summarise(
      across(
        .cols = any_of(ends_with(c("Nov","Dec","Jan","Feb","Mar","Apr"))),
        .fns = ~cor(.,!!sym(fao_feature),use= "pairwise.complete.obs")
      ),.groups="drop"
    ) |>
    mutate(
      fao_feature = {{fao_feature}}
    )
   col_rgx <- glue_collapse(glue("^{snow_feature}_snow_{mo_labs}"),sep = "|")
   df_corrs_long <- df_corrs |>
     select(
       all_of(c("type",
                "adm1_name")),matches(col_rgx)
  ) |>
  pivot_longer(
    cols = matches(col_rgx)
  ) |>
     mutate(
       mo = factor(str_extract(name,mo_rgx),levels=c("Nov","Dec","Jan","Feb","Mar","Apr"))
     )
   df_corrs_long |>
     ggplot(
                aes(x= mo,
                    y= adm1_name,
                    fill = value)
              )+
              geom_tile()+
              scale_fill_gradient2(
                low = hdx_hex("tomato-hdx"),      # Color for negative values
                mid = "white",    # Color for zero
                high = hdx_hex("mint-hdx"),   # Color for positive values
                midpoint = 0      # Set midpoint at zero

              )+
              geom_tile(
                data= df_corrs_long |>
                  filter(adm1_name == "Faryab",
                         name %in% c("Nov","Dec","Jan","Feb","Mar","Apr")),
                fill = NA, color ="black", lwd = 1.5
              )+

              geom_text(
                aes(label = round(value,2))
              )+
              labs(
                title = p_title,
                # subtitle = "Afghanistan by Province",
                y= "Province"
              )+
              theme(
                axis.title.x = element_blank(),
                legend.title= element_blank(),
                plot.title = element_text(size = 12),
                plot.subtitle = element_text(size = 12),
                legend.text = element_text(angle=90)
              )

}


all_feature_combos <- expand_grid(
  snow_features = c("value",
                    "anom",
                    "anom_z",
                    "value_change"),
  fao_features = c( "mam_mean_value",
                    "mamj_mean_value",
                    "may_last_value",
                    "june_last_value",
                    "mam_mean_anom",
                    "mamj_mean_anom",
                    "mam_mean_z",
                    "mamj_mean_z"
                    )
) |>
  mutate(
    label = glue ("Snow: {snow_features} vs Vegetation: {fao_features}")
    )







lps_ndsi_asi <- map(
  set_names(all_feature_combos$label,all_feature_combos$label),
  \(feature_combo){
    df_feature<- all_feature_combos |>
      filter(label == feature_combo)

    corr_all_snow(
    df_snow= df_snow_ready,
    snow_param ="NDSI",
    snow_feature = df_feature$snow_features,
    df_veg= df_fao_yr,
    veg_param ="asi",
    fao_feature =df_feature$fao_features,
    start_year = 2000
    )
  }
)



lps_swe_asi <- map(
  set_names(all_feature_combos$label,all_feature_combos$label),
  \(feature_combo){
    df_feature<- all_feature_combos |>
      filter(label == feature_combo)

    corr_all_snow(
    df_snow= df_snow_ready,
    snow_param ="SWE",
    snow_feature = df_feature$snow_features,
    df_veg= df_fao_yr,
    veg_param ="asi",
    fao_feature =df_feature$fao_features,
    start_year = 2000
    )
  }
)
```

### NDSI vs ASI

Here we see the correlation between monthly NDSI values and the cumulative `ASI` value for june (3rd dekad)

```{r}
#| fig.height: 8
#| label: fig-plot_ndsi_value_june_asi

lps_ndsi_asi$`Snow: value vs Vegetation: june_last_value`

```

Here we see the correlation between monthly NDSI values and the cumulative `ASI` value for May (3rd dekad)

```{r}
#| fig.height: 8
#| label: fig-plot_ndsi_value_may_asi

lps_ndsi_asi$`Snow: value vs Vegetation: may_last_value`
```

### SWE vs ASI

We see the same relation ship when we look at SWE - here we have filtered SWE to the same date range as MODIS for comparison (>= 2000)

```{r}
#| fig.height: 8
#| label: fig-plot_swe_value_june_asi

lps_swe_asi$`Snow: value vs Vegetation: june_last_value`

```

```{r}
#| fig.height: 8
#| label: fig-plot_swe_value_may_asi

lps_swe_asi$`Snow: value vs Vegetation: may_last_value`

```

We then run it on the full historical record of FLDAS SWE that overlaps w/ ASI >=1984
```{r}

lps_swe_asi_full <- map(
  set_names(all_feature_combos$label,all_feature_combos$label),
  \(feature_combo){
    df_feature<- all_feature_combos |>
      filter(label == feature_combo)

    corr_all_snow(
    df_snow= df_snow_ready,
    snow_param ="SWE",
    snow_feature = df_feature$snow_features,
    df_veg= df_fao_yr,
    veg_param ="asi",
    fao_feature =df_feature$fao_features,
    start_year = NULL
    )
  }
)

lps_swe_asi_full$`Snow: value vs Vegetation: may_last_value`
```


### Snow Discussion

-   Oddly the Dec & Jan snow indicators seem if anything positively correlated with ASI which is the exact opposite of the hypothesis. The charts above show that higher snow fractions/and snow water equivalents are associated with higher end of season agricultural stress (May & June).
-   There could be a variety of factors that causing this unexpected relationship. A few hypothesized below:
    -   More Dec & Jan snow associated with later planting season - poor vegetative outcomes
    -   More Dec & Jan snow associated with colder seasons - less productive harvests
    -   More Dec & Jan snow associated with warmer springs - earlier snow melt
    -   More Dec & Jan snow associated with less spring rains
    -   It does seem like the Feb-Mar the relationship shifts back to the directionality we would expect: More snow:less stress. This could indicate that it's not simply the total snowfall impacting water availability and crop health, but perhaps there is a later season dynamic (snow melt?) that impacts water availability. Perhaps if there is some (x?) amount of snow available, it keeps a consistent supply of water to crops, but if it disappears too early, than the agriculture suffers

### Update

When working on the later included `indicator-expansion` chapter, some snow related indicators/bands from ERA5-LAND image collection were extracted.These include 2 bands  specifically:

- snow_cover
- snow_depth_water_equivalent

The correlations for these two bands from the ERA5-LAND sensor image collection are shown below just for the 5 provinces under initial consideration (data only extracted to these provinces)

We see that the counter-intuitive relationship observed above is less obvious and that many of the provinces have a mild negative relationship

```{r}
lps_era_snow_asi <- map(
  set_names(all_feature_combos$label,all_feature_combos$label),
  \(feature_combo){
    df_feature<- all_feature_combos |>
      filter(label == feature_combo)

    corr_all_snow(
    df_snow= df_snow_ready,
    snow_param ="Snow Cover ERA",
    snow_feature = df_feature$snow_features,
    df_veg= df_fao_yr,
    veg_param ="asi",
    fao_feature =df_feature$fao_features,
    start_year = 2000
    )
  }
)

lps_era_sdwe_asi <- map(
  set_names(all_feature_combos$label,all_feature_combos$label),
  \(feature_combo){
    df_feature<- all_feature_combos |>
      filter(label == feature_combo)

    corr_all_snow(
    df_snow= df_snow_ready,
    snow_param ="SDWE ERA",
    snow_feature = df_feature$snow_features,
    df_veg= df_fao_yr,
    veg_param ="asi",
    fao_feature =df_feature$fao_features,
    start_year = 2000
    )
  }
)
lps_era_snow_asi$`Snow: value vs Vegetation: may_last_value`
```

```{r}
lps_era_sdwe_asi$`Snow: value vs Vegetation: may_last_value`
```

We then run it on the full historical record of ERA5-LAND that overlaps w/ ASI >=1984
```{r}
lps_era_snow_asi_full <- map(
  set_names(all_feature_combos$label,all_feature_combos$label),
  \(feature_combo){
    df_feature<- all_feature_combos |>
      filter(label == feature_combo)

    corr_all_snow(
    df_snow= df_snow_ready,
    snow_param ="Snow Cover ERA",
    snow_feature = df_feature$snow_features,
    df_veg= df_fao_yr,
    veg_param ="asi",
    fao_feature =df_feature$fao_features,
    start_year = NULL
    )
  }
)

lps_era_sdwe_asi_full <- map(
  set_names(all_feature_combos$label,all_feature_combos$label),
  \(feature_combo){
    df_feature<- all_feature_combos |>
      filter(label == feature_combo)

    corr_all_snow(
    df_snow= df_snow_ready,
    snow_param ="SDWE ERA",
    snow_feature = df_feature$snow_features,
    df_veg= df_fao_yr,
    veg_param ="asi",
    fao_feature =df_feature$fao_features,
    start_year = NULL
    )
  }
)
lps_era_snow_asi_full$`Snow: value vs Vegetation: may_last_value`
```

```{r}
lps_era_sdwe_asi_full$`Snow: value vs Vegetation: may_last_value`
```
