---
title: Indicator Expansion

format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  include: true
  echo: true
  warning: false
  message: false
  eval: true
  results: "asis"
  out.width: "100%"
  code-fold: true
editor: visual
project:
  execute-dir: project
---

## Intro
```{r}

aoi_adm1 <- c(
  "Takhar",
  # "Badakhshan",
  # "Badghis",
  "Sar-e-Pul" ,
  "Faryab"
  )

box::use(
  ../R/blob_connect,
  ../R/utils,
  seas5 = ../R/seas5_utils,
  loaders = ../R/load_funcs
)

# not using box as it doesn't play nice w/ meta pacakges like `{tidymodels}`
# in refactoring I should probably just laod this specific packages necessary
# for tidymodels and ultimately move all the heavy tidymodels work to a new
# notebook

library(tidyverse)
library(cumulus)
library(gt)
library(gghdx)
library(glue)
library(janitor)
library(ggiraph)
library(ggh4x)
library(ggrepel)
library(sf)
library(tidymodels)
library(vip)
gghdx()
```


```{r}

label_parameters <- function(df){
  df |>
    mutate(
     parameter_label = case_when(
      str_detect(parameter, "era5_land_volumetric_soil")~ "Soil Moisture (ERA5)",
      str_detect(parameter,"NDSI")~"NDSI",
      str_detect(parameter,"asi")~"ASI",
      str_detect(parameter,"vhi")~"VHI",

      str_detect(parameter,"cumu_chirps_precipitation_sum")~"Precip cumu (CHIRPS) ",
      str_detect(parameter,"chirps_precipitation_sum")~"Precip (CHIRPS)",
      str_detect(parameter,"cumu_era5_land_total_precipitation_sum")~"Precip cumu (ERA)",
      str_detect(parameter,"era5_land_total_precipitation_sum")~"Precip (ERA)",
      str_detect(parameter,"mean_2m_air_temperature")~"Temp (ERA)",
      str_detect(parameter,"era5_land_snow_depth_water_equivalent")~"SDWE (ERA5)",
      str_detect(parameter,"era5_land_snow_cover")~"Snow Cover (ERA5)",
      str_detect(parameter,"era5_land_snowmelt_sum")~"Snow Melt (ERA5)",

      str_detect(parameter,"runoff_max")~"Runoff max (ERA5)",
      str_detect(parameter,"runoff_sum")~"Runoff sum (ERA5)",
      str_detect(parameter,"SWE_inst")~"SWE (FLDAS)",


      .default = parameter
    )
    )
}



filter_to_validation_range <- function(df,df_valid){
  df |>
    filter(
      yr_season >= min(df_valid$yr_season),
      yr_season <= max(df_valid$yr_season)
    )
}

load_aggregated_forecast <- function(aoi_adm1,valid_months=c(3,4,5)){
   valid_months_chr <- lubridate::month(valid_months,abbr=T,label=T)
   parameter_label <- glue::glue("SEAS5-{glue::glue_collapse(str_sub(valid_months_chr,1,1),sep='')}")


   df_seas5 <- cumulus::pg_load_seas5_historical(
    iso3 ="AFG",
    adm_level = 1,
    adm_name = aoi_adm1,
    convert_units = T
  )

  df_agg <- cumulus::seas5_aggregate_forecast(
    df_seas5,
    valid_months = valid_months,
    value= "mean",
    by = c("iso3","pcode","name","issued_date")
  )

  df_agg |>
    mutate(
        yr_season = lubridate::floor_date(issued_date+months(leadtime),"year"),
        parameter = parameter_label
      )  |>
    select(
       yr_season, adm1_name=name, pub_mo_date =issued_date, value=mean,parameter
    )



}

load_compiled_indicators <- function(aoi_adm1) {
  df_observational <- loaders$load_cleaned_env_features(mo = c(11,12,1:6))
  df_obs_clean <- loaders$prep_observational_data(df= df_observational)
  df_validation_set <- df_obs_clean |>
    filter(parameter == "asi")

  df_seas5_forecast <- load_aggregated_forecast(aoi_adm1 = aoi_adm1,valid_months  = c(3,4,5))

  df_obs_pred <- bind_rows(
    df_obs_clean,
    df_seas5_forecast
    )


  df_obs_pred |>
    filter_to_validation_range(df_valid = df_validation_set) |>
    filter(adm1_name %in% aoi_adm1) |>
    group_by(adm1_name, parameter, pub_mo_label = month(pub_mo_date,label = T, abbr = T)) |>
    arrange(desc(value), .by_group = TRUE) |>
    mutate(rank = row_number(), q_rank = rank/(max(rank)+1), rp_max_direction = 1/q_rank) |>
    arrange(value, .by_group = TRUE) |>
    mutate(rank = row_number(), q_rank = rank/(max(rank)+1), rp_min_direction = 1/q_rank) |>
    ungroup() |>
    mutate(rp_relevant_direction  = case_when(
      str_detect(parameter,"asi|temp")~ rp_max_direction,
      .default = rp_min_direction
    ))

}


```










```{r}

df_compiled_indicators <- load_compiled_indicators(aoi_adm1 = aoi_adm1)

df_indicators_flagged <- df_compiled_indicators |>
  mutate(
    flag = rp_relevant_direction>=3
  )

# get validation set flagged w/ RP breaches
df_valid_flag <-
  df_indicators_flagged |>
  filter(
    # May ASI has been used as our validation data set
    month(pub_mo_date) ==6 &
    str_detect(parameter,"asi")
  ) |>
  rename(
    valid_flag = flag
  ) |>
  select(
   yr_season,adm1_name,valid_flag
  )

df_valid_flag |>
  arrange(
    adm1_name,yr_season
  ) |>
  print(n=100)


df_env_compare <- df_indicators_flagged |>
  left_join(
    df_valid_flag,
    by = c("yr_season","adm1_name")
  ) |>
  filter_to_validation_range(df_valid = df_validation_set)

df_env_model <-
  df_env_compare |>
    select(
    date,
    yr_season,
    pub_mo_date,
    pub_mo_label,
    adm1_name,
    parameter,
    value
    ) |>
  filter(
    !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max","era5_land_runoff_max"),
    !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d"),

    # already have these from ERA land up to 2024 rather than 2020
    !(parameter %in% c("total_precipitation","mean_2m_air_temperature"))

  )




df_perf_metrics <- df_env_compare |>
  mutate(
    TP = flag & valid_flag,
    TN = !flag & !valid_flag,
    FP = flag & !valid_flag,
    FN = !flag & valid_flag
  ) |>
  group_by(
    adm1_name, pub_mo_label = month(pub_mo_date,label=T,abbr=T),parameter
  ) |>
  summarise(
    precision = sum(TP)/(sum(TP)+sum(FP)),
    recall = sum(TP)/(sum(TP)+sum(FN)),
    accuracy = (sum(TP)+sum(TN))/(sum(TP)+sum(TN)+sum(FP)+sum(FN)),
    .groups="drop_last"
  ) |>
  mutate(
    f1 = (2 * precision * recall) / (precision + recall),
    # when precision & recall are 0 we get NaN - replace w/ 0 for easier interpretation
    f1 = ifelse(precision==0&recall==0,0,f1),
    rank_correct= dense_rank(accuracy)
  ) |>
    mutate(
    parameter_simplified = case_when(
      # str_detect(parameter, "era5_land_volumetric_soil")~ "Soil Moisture (ERA5)",
      str_detect(parameter,"NDSI")~"NDSI",
      str_detect(parameter,"asi")~"ASI",
      str_detect(parameter,"vhi")~"VHI",

      str_detect(parameter,"cumu_chirps_precipitation_sum")~"Precip cumu (CHIRPS) ",
      str_detect(parameter,"chirps_precipitation_sum")~"Precip (CHIRPS)",
      str_detect(parameter,"cumu_era5_land_total_precipitation_sum")~"Precip cumu (ERA)",
      str_detect(parameter,"era5_land_total_precipitation_sum")~"Precip (ERA)",
      str_detect(parameter,"era5_land_temperature_2m")~"Temp (ERA)",
      str_detect(parameter,"era5_land_snow_depth_water_equivalent")~"SDWE (ERA5)",
      str_detect(parameter,"era5_land_snow_cover")~"Snow Cover (ERA5)",
      str_detect(parameter,"era5_land_snowmelt_sum")~"Snow Melt (ERA5)",
      str_detect(parameter,"cumu_era5_land_temperature_2m")~"Temp cumu (ERA5)",

      str_detect(parameter,"runoff_max")~"Runoff max (ERA5)",
      str_detect(parameter,"runoff_sum")~"Runoff sum (ERA5)",
      str_detect(parameter,"SWE_inst")~"SWE (FLDAS)",
      .default = parameter
    )
  ) |>
  ungroup()


```

```{r}
aoi_adm1 <- set_names(aoi_adm1,aoi_adm1)

p_metrics_seasonal <-
  aoi_adm1 |>
  map(\(adm_tmp){

    df_perf_metrics |>
      ungroup() |>
      filter(
        adm1_name == adm_tmp
      ) |>
      mutate(
        trigger_type = factor(case_when(
          pub_mo_label %in% c("Jan","Feb","Mar")~"Pre-Season",
          pub_mo_label %in% c("Mar","Apr","May")~"Mid-Season",
          pub_mo_label %in% c("Jun","Jul")~"Post-Season"
        ), levels = c("Pre-Season","Mid-Season","Post-Season")
        )
      ) |>
      pivot_longer(cols = precision:accuracy) |>
      filter(pub_mo_label %in% c("Jan","Feb","Mar","Apr","May","Jun","Jul")) |>
      ggplot(
        aes(
          y = reorder(parameter,value),
          x= value,
          color = name
        )
      )+
      geom_point_interactive(
        aes(tooltip= glue(
          "parameter: {parameter},
      metric: {name}
      value: {scales::label_percent(accuracy = 1)(value)}")
        ),alpha=0.4
      )+
      facet_nested(
        rows=vars(trigger_type, pub_mo_label),
        # nest_line = element_line(colour = "red"),
        scales="free")+

      scale_x_continuous(labels =scales::label_percent())+
      scale_y_discrete(expand= c(.05,.05))+
      labs(
        title =  "How well do RS parameters (thresholded - 5 year RP) predict cumulative May ASI breaching 5 year RP",
        subtitle =glue("{adm_tmp} Province")

      )+
      theme(
        # strip.text = element_text(angle=90),
        ggh4x.axis.nestline = element_line(linetype = 2),
        panel.border = element_rect(fill=NA, color ="grey"),
        legend.title = element_blank(),
        axis.text.y= element_text(size=6),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),
        plot.title = element_text(size =9),
        plot.subtitle = element_text(size =8),strip.background = element_rect(fill="lightgrey")
      )
  }
  )
```

```{r}
aoi_adm1 <- set_names(aoi_adm1,aoi_adm1)

lp_tile_metrics_seasonal <-
  aoi_adm1 |>
  map(\(adm_tmp){

    df_temp <- df_perf_metrics |>
      ungroup() |>
      filter(
        adm1_name == adm_tmp
      ) |>
      mutate(
        trigger_type = factor(case_when(
          pub_mo_label %in% c("Jan","Feb","Mar")~"Pre-Season",
          pub_mo_label %in% c("Mar","Apr","May")~"Mid-Season",
          pub_mo_label %in% c("Jun","Jul")~"Post-Season"
        ), levels = c("Pre-Season","Mid-Season","Post-Season")
        )
      ) |>
      filter(pub_mo_label %in% c("Jan","Feb","Mar","Apr","May","Jun","Jul"))

    df_temp |>
      ggplot(
        aes(
          x= pub_mo_label,
          y = parameter,
          fill = f1
        )
      )+
      geom_tile_interactive(
        aes(tooltip= glue(
          "parameter: {parameter},
      metric: {f1}
      value: {scales::label_percent(accuracy = 1)(f1)}")
        ),alpha=1
      )+
        geom_tile(
    data = df_temp|>
      filter(f1>=0.7),
      # group_by(pub_mo_label) |>
      # slice_max(
      #   f1, n= 3
      # ),
      fill = NA, color = "red", linewidth = 1
  )+
      labs(
        title =  "How well do RS parameters (thresholded - 5 year RP) predict cumulative May ASI breaching 5 year RP",
        subtitle =glue("{adm_tmp} Province")

      )+
      theme(
        # strip.text = element_text(angle=90),
        # ggh4x.axis.nestline = element_line(linetype = 2),
        panel.border = element_rect(fill=NA, color ="grey"),
        legend.title = element_blank(),
        axis.text.y= element_text(size=6),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),
        plot.title = element_text(size =9),
        plot.subtitle = element_text(size =8),
        strip.background = element_rect(fill="lightgrey")
      )
  }
  )
```

## Indicator Performance by Trigger/Framework Stage

-   Below the we look at measured indicators performance. How well does a 5 year RP or greater event measured across each indicator perform in predicting a 5 year RP ASI drought level as measured at the end of March.
    -   The indicators are broken up into 3 major categories (`pre-season`, `mid-season`, `post-season`), based on when each indicator/trigger moment will become available. Note the months represented in the strips on right represent the month the data will be available. Therefore, they are typically put as 1 month after the date the indicator is monitoring. For example NDVI for March will be fully available in April. This is true for all indicators except for the seasonal forecast (`seas5`) where a forecast for the current month is provided at the start of the month the data is published (5th of month)

**Quick Initial Takeaways:**

-   All indicators pre-season indicators perform weakly in January across all admin units

-   `SEAS5` (seasonal forecast) is the best pre-season indicator that we have the earliest in `Feb` in most provinces

-   More indicators become better predictors in March (labelled as pre-season, but not really sufficient leadtime for agricultural activities and operational leadtimes received).

-   In March SEAS5 prediction for MAM is still the most predictive preciptiation indicator we have.

-   By March/April, soil moisture looks very promising as well as other combinations with snow indicators and precipitation

-   From April onwards observed precipitation becomes more predictive. With CHIRPS leading over ERA5 by a usually a minor except in Bagdhis where ERA5 is much more performant.

### Interpreting plots:

Below the performance metrics are given for each parameter and admin unit. The following metrics are calculated:

-   Accuracy
-   Precision
-   Recall

As we are predicting relatively rare events the accuracy metric should be supported by interpretation of precision and recall.

Below we show all trigger moment/months and there `f1 score` (harmonic mean of precision and recall) for predicting end of season ASI (dekad 3 may).

```{r}
gdf_adm1 <- cumulus::download_fieldmaps_sf(iso3= "afg",layer= "afg_adm1")

df_pin <- cumulus::blob_read(
  name= "ds-aa-afg-drought/raw/vector/HNRP_2025_PiN_Targets_Severity.xlsx",
  container = "projects"
)

df_pin_wts <- df_pin |>
  clean_names() |>
  filter(province %in% aoi_adm1) |>
  group_by(province) |>
  summarise(
    fsc_pin_wt = sum(pin_fsc)
  ) |>
  rename(
    adm1_name = "province"
  )

df_area_wt_table <- gdf_adm1$afg_adm1 |>
  clean_names() |>
  filter(adm1_en %in% aoi_adm1) %>%
  mutate(
    area_wt = as.numeric(st_area(.))
  ) |>
  st_drop_geometry() |>
  select(adm1_name = adm1_en,adm1_pcode,area_wt) |>
  left_join(
    df_pin_wts
  ) |>
  mutate(
    area_x_fsc_wt =area_wt * fsc_pin_wt
  )


generalize_f1 <- function(df, aoi, weight=NULL){
  df_agg <- df |>
    filter(
      adm1_name %in% aoi
    ) |>
    group_by(
      pub_mo_label, parameter,parameter_simplified
    )
  if(is.null(weight)){
    ret <-  df_agg |>
    summarise(
      across(
        .cols = c("f1","precision","recall"), \(x) mean(x,na.rm=T)
      ),
      .groups="drop"
      # f1 = mean(f1, na.rm = TRUE),
      # precision = mean(precision, na.rm = TRUE),
      # recall = mean(precision, na.rm = TRUE),.groups="drop"
    )
  }
  if(!is.null(weight)){
    ret <- df_agg |>
      summarise(
        across(
         .cols = c("f1","precision","recall"), \(x) weighted.mean(x,,w =!!sym(weight),na.rm=T)
        ),

         .groups="drop"
        # f1 = weighted.mean(f1,w=!!sym(weight),na.rm=T),
      )
  }
  ret
}

wt_set <- list(NO_WT = NULL, AREA_WT = "area_wt", AREA_FSC_PIN_WT = "area_x_fsc_wt")
df_perf_metrics_w_wt <- df_perf_metrics |>
  left_join(df_area_wt_table)

df_perf_metrics_w_wt |> count(parameter) |> print(n=23)

df_perf_metrics_w_wt_filt <- df_perf_metrics_w_wt |>
        filter(
          !parameter %in% c("NDSI_Snow_Cover_min",
                            "NDSI_Snow_Cover_max",
                            "era5_land_runoff_max"),
          !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d")
        )

perf_param <- c("precision","recall","f1")[3]

lp_f1_combined <- imap(
  wt_set,
  \(wt_temp,nmt){
    dft <- generalize_f1(
      df= df_perf_metrics_w_wt_filt,
      aoi =  c("Takhar","Sar-e-Pul","Faryab"),
      weight=wt_temp
    )
    dft_filt <-  dft |>
      filter(
        pub_mo_label %in% c("Feb","Mar","Apr","May","Jun")
      ) |>
      mutate(
        f1_gte_0.5 = !!sym(perf_param)>=0.5
      )

    subtitle_tmp <- switch(nmt,
            "NO_WT"= "Unweighted",
            "AREA_WT"= "Weighted by province area",
            "AREA_FSC_PIN_WT"= "Weighted by province area & FSC PiN"
    )

    dft_filt |>
      ggplot(
        aes(
          x= pub_mo_label,
          y = parameter_simplified,
          fill = !!sym(perf_param)
        )
      )+
      geom_tile_interactive(
        aes(tooltip= glue(
          "parameter: {parameter},
      F1 score: {scales::label_percent(accuracy = 1)(f1)}"
      )
        ),alpha=1
      )+
      labs(
        title =  subtitle_tmp

      )+
      geom_tile(
        data = dft_filt |>
          group_by(pub_mo_label) |>
          slice_max(
            !!sym(perf_param),
            n= 5
          ),
        aes(color = f1_gte_0.5),
        fill = NA, show.legend=F,
        # color = "red",
        linewidth = 1
      )+
      scale_fill_viridis_c()+
      scale_color_manual(
        values = c(
          "TRUE"= hdx_hex("tomato-hdx"),
          "FALSE"="grey"
          )
      )+
      theme(
        # ggh4x.axis.nestline = element_line(linetype = 2),
        panel.border = element_rect(fill=NA, color ="grey"),
        legend.title = element_blank(),
        axis.text.y= element_text(size=10),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),

        strip.background = element_rect(fill="lightgrey")
      )

  }

)
lp_f1_combined$AREA_WT
lp_f1_combined$NO_WT


ldf_f1_combined <- imap(
  wt_set,
  \(wt_temp,nmt){
     wt_label <- switch(nmt,
            "NO_WT"= "Unweighted",
            "AREA_WT"= "Weighted by province area",
            "AREA_FSC_PIN_WT"= "Weighted by province area & FSC PiN"
    )
    dft <- generalize_f1(
      df= df_perf_metrics_w_wt |>
        filter(
          !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max"),
          !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d")
        ),
      aoi =  c("Takhar","Sar-e-Pul","Faryab"),
      weight=wt_temp
    )
    dft |>
      filter(
        pub_mo_label %in% c("Feb","Mar","Apr","May","Jun")
      ) |>
      mutate(
        f1_gte_0.5 = f1>=0.4,
        weight_type = wt_label
      )
  }) |>
  list_rbind()
ldf_f1_combined

lp_ndsi_params <- map(
  wt_set,
  \(wt_temp){
    dft <- generalize_f1(
      df= df_perf_metrics_w_wt |>
        filter(
          parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_mean","NDSI_Snow_Cover_max")
        ),
      aoi =  c("Takhar","Sar-e-Pul","Faryab"),
      weight=wt_temp
    )
    dft_filt <-  dft |>
      filter(
        pub_mo_label %in% c("Feb","Mar","Apr","May","Jun")
      ) |>
      mutate(
        f1_gte_0.5 = f1>=0.4
      )

    dft_filt |>
      ggplot(
        aes(
          x= parameter,
          y = f1
        )
      )+
      geom_point()+
      coord_flip()+
      facet_wrap(~pub_mo_label)+
      theme(
        # strip.text = element_text(angle=90),
        ggh4x.axis.nestline = element_line(linetype = 2),
        panel.border = element_rect(fill=NA, color ="grey"),
        legend.title = element_blank(),
        axis.text.y= element_text(size=10),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),
        # plot.title = element_text(size =9),
        # plot.subtitle = element_text(size =8),
        strip.background = element_rect(fill="lightgrey")
      )

  }

)

lp_ndsi_params$NO_WT
library(patchwork)
lp_f1_combined$NO_WT +
  lp_f1_combined$AREA_WT +
  lp_f1_combined$AREA_FSC_PIN_WT + plot_layout(nrow = 1)+
  plot_annotation(title= "Average F1 Score Across Provinces")

girafe(ggobj =lp_f1_combined$AREA_FSC_PIN_WT)

```

```{r}
#| fig.height: 9

lp_tile_metrics_seasonal$`Sar-e-Pul`
```

```{r}
#| fig.height: 9

lp_tile_metrics_seasonal$Takhar
```

```{r}
#| fig.height: 9

lp_tile_metrics_seasonal$Faryab
```

```{r}
#| eval: false
#| fig.height: 9

# no longer in AOI
# lp_tile_metrics_seasonal$Badakhshan
```

```{r}
girafe(ggobj= p_metrics_seasonal$Faryab)
```

```{r}
#| fig.height: 9

girafe(ggobj= p_metrics_seasonal$`Sar-e-Pul`)
```

```{r}
#| fig.height: 9

girafe(ggobj= p_metrics_seasonal$Takhar)
```

```{r}
lp_metric <-
  aoi_adm1 |>
  map(\(adm_tmp){

    df_perf_metrics |>
      filter(adm1_name == adm_tmp) |>
      pivot_longer(cols = precision:accuracy
      ) |>
      ggplot(
        aes(
          x=reorder(parameter,value),
          y= value,
          color = name)
      )+
      geom_point_interactive(
        aes(tooltip= glue(
          "parameter: {parameter},
      metric: {name}
      value: {scales::label_percent(accuracy = 1)(value)}")
        ),alpha=0.4
      )+
      coord_flip()+
      scale_y_continuous(labels =scales::label_percent())+
      labs(
        title =  "How well do RS parameters (thresholded - 5 year RP) predict cumulative May ASI breaching 5 year RP",
        subtitle =glue("{adm_tmp} Province")

      )+
      theme(
        legend.title = element_blank(),
        axis.text.y= element_text(size=6),
        axis.title.x= element_blank(),
        axis.title.y= element_blank(),
        plot.title = element_text(size =9),
        plot.subtitle = element_text(size =8)
      )
  }
  )


girafe(ggobj = lp_metric$Takhar)
```

Here we can see how closer to the end of season our performance metrics improve.

```{r}

df_metrics_simple <- df_perf_metrics |>
  filter(adm1_name!="Badghis",adm1_name!="Badakhshan") |>
    filter(
          pub_mo_label %in% c("Feb","Mar","Apr","May","Jun"),
          !str_detect(parameter,"runoff_max"),

           !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max","total_precipitation"),
          !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d")
        ) |>
  # filter(!is.nan(f1)) |>
  label_parameters()





df_metrics_filt <-  df_metrics_simple |>
        filter(
          pub_mo_label %in% c("Feb","Mar","Apr","May","Jun"),
          !str_detect(parameter,"runoff_max"),

           !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max","total_precipitation"),
          !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d")
        )

df_top_metrics_label <- df_metrics_filt |>
  mutate(
    f1_lgl = f1>0.5
  ) |>
  group_by(adm1_name, pub_mo_label) |>
  mutate(
    num_gte = sum(f1_lgl)
  ) |>
  arrange(
   adm1_name, pub_mo_label, desc(f1)
  ) |>
  mutate(
    include = row_number()
  ) |>
  print(n=25) |>
  filter(include%in% c(1:3) & f1_lgl) |>
  ungroup()

df_top_metrics_label |>
  filter(adm1_name == "Sar-e-Pul")

df_metrics_top3 <- df_metrics_filt |>
        group_by(adm1_name, pub_mo_label) |>
        slice_max(f1, n= 3, with_ties = T)


viridis_colors <- viridisLite::viridis(15, option = "A")
length(c25)
c25 <- c(

  "green4",
  "#E31A1C",
  "dodgerblue2",
  "#6A3D9A", # ndsi purple
  "#FF7F00", # orange
  "skyblue2",
  "gold1",
  "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70",
  "khaki2",
  "maroon",
  "orchid1",
  "deeppink1",
  "blue1", # soil mositure
  "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown"
)





p_ts_f1 <- df_metrics_top3 |>
  ggplot(
    aes(
      x= pub_mo_label, y= f1
    )
  )+
  geom_point(
    data = df_metrics_filt,
    alpha=0.5, color = "darkgrey"
  )+
  geom_point(
    aes(color =parameter_simplified)
  )+
  geom_label_repel(
    data = df_metrics_top3,
    aes(
      label = parameter_simplified,
      fill = parameter_simplified
    ),
    max.overlaps = 20,
    color = "black"
    )+
  # scale_fill_brewer(palette = "Set3")+
  scale_fill_manual(values = c25) +
  labs(
    # title = "F1 Scores: How well do indicators predict worst end of season ASI values",
    title = "F1 Scores: By Province & Trigger Moment",
    subtitle = "Only Top indicators labelled"
  )+
  scale_color_manual(values = c25) +
  facet_wrap(~adm1_name)+
  theme(
    panel.grid.major.x = element_line(color = "black"),
    legend.position = "none"

  )

# whta if we just show values > 0.5
# p_ts_f1_gte0.5 <- df_metrics_filt |>
#    filter(
#           !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max"),
#           !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d")
#         ) |>
#   filter(f1>=0.5) |>
#   filter(
#     adm1_name != "Badkhshan"
#   # ) |>



ggplot(
  df_metrics_filt,
  aes(
    x= pub_mo_label, y= f1
  )
)+
  geom_point(
    alpha=0.5,
    color = "darkgrey"
  )+

  geom_point(
    data= df_top_metrics_label,
    aes(color =parameter_simplified)
  )+
  geom_label_repel(
    data =  df_top_metrics_label ,
    aes(
      label = parameter_simplified,
      fill = parameter_simplified
    ),
    max.overlaps = 20,
    color = "black"
    )+
  # scale_fill_brewer(palette = "Set3")+
  scale_fill_manual(values = c25) +
  scale_color_manual(values = c25)+
  labs(
    # title = "F1 Scores: How well do indicators predict worst end of season ASI values",
    title = "Predictiveness of Indicators by Province and Trigger Timing (F1 Score)",
    subtitle = "Only Top 3 Indicators Labelled Per Publication Month and Province"
  )+
  facet_wrap(~adm1_name,ncol=1)+
  theme(
    panel.grid.major.x = element_line(color = "black"),
    legend.position = "none"

  )

p_ts_f1_gte0.5
library(patchwork)
# box::use(patchwork[...])

p_ts_f1 +
  p_perf_tile_combined+

  plot_layout(widths = c(2,1))+
  plot_annotation(
    title = "How well do RS indicators predict worst end of season ASI values?"
  )
```



### Heuristic Weighting

Based on the above let's select some indicators and make a weighting

```{r}
library(tidymodels)
library(usemodels)
library(vip)

df_env_model <-
  df_env_compare |>
    select(
    date,
    yr_season,
    pub_mo_date,
    pub_mo_label,
    adm1_name,
    parameter,
    value
    ) |>
  filter(
    !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max","era5_land_runoff_max"),
    !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d"),

    # already have these from ERA land up to 2024 rather than 2020
    !(parameter %in% c("total_precipitation","mean_2m_air_temperature"))

  )

# just some manual inspectino stuff here
df_env_model |>
  filter(str_detect(parameter,"^asi|^vhi")) |>
  group_by(parameter, pub_mo_label) |>
  count() |>
  pivot_wider(names_from=pub_mo_label, values_from = n)

date_ranges <- df_env_model |>
  group_by(parameter) |>
  summarise(
    start = min(pub_mo_date),
    end = max(pub_mo_date)
  )


# df_jun_asi <- df_env_model |>
#   filter(parameter == "asi",month(pub_mo_date)==6) |>
#   select(
#     yr_season,adm1_name, value
#   ) |>
#   rename(
#     asi_Jun = value
#   )






#' create_weight_grid
#' @description
#' function to create valid weight sets in long data.frame that can be
#' iterated through
#' @param x
#' @param wt_vals
#'
#' @returns
#' @export
#'
#' @examples
#' params_set_gte_1984 <- c(
#'   "era5_land_soil_moisture_1m",
#'   "cumu_era5_land_total_precipitation_sum",
#'   "vhi",
#'   "era5_land_snow_cover", # toying w/ adding this or not
#'   "asi"
#'   )
#' weight_combos <- create_weight_grid(params_set_gte_1984,wt_vals = c(0,seq(0.1, 1, by = 0.1)))

create_weight_grid <- function(x,wt_vals){

  df <- map(x,\(xt){
    tibble(
      !!sym(xt) :=wt_vals
    )
  }
  )|>
    list_cbind()

  df_expanded <- expand_grid(!!!df)

  df_valid_combinations <- df_expanded[rowSums(df_expanded) == 1, ]

    df_valid_combinations |>
      mutate(
        wt_id = row_number()
      ) |>
      pivot_longer(
        -wt_id, names_to = "parameter", values_to ="weight"
      )
}



#' Title
#'
#' @param df
#' @param params_included
#' @param earliest_year
#'
#' @returns
#' @export
#'
#' @examples
#' df_env_model |>
#'   normalize_to_z(params_included = l_params$gte1984,earliest_year = 1984)

normalize_to_z <- function(df,params_included, earliest_year){
      df |>
        filter(
          year(yr_season)>=earliest_year
        ) |>
        mutate(
        pub_mo_label = as.character(pub_mo_label)
      ) |>
      filter(
        parameter %in% params_included,
        pub_mo_label %in% c(month.abb[4:6])
      ) |>
      group_by(
        pub_mo_label, adm1_name, parameter
      ) |>
      mutate(
        zscore = scale(value,center=T,scale=T)[,1],
        zscore = ifelse(parameter != "asi",zscore*-1,zscore)
      )  |>
      ungroup()
}

#' Title
#'
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
#' df_env_model |>
#'   normalize_to_z(params_included = l_params$gte2000,earliest_year = 2001) |>
#'   extract_truth_set()

extract_truth_set <- function(df){
  df |>
    filter(
      month(pub_mo_date)==6,
      parameter == "asi"
    ) |>
    select(
      yr_season, adm1_name, value,zscore_asi_Jun=zscore
    ) |>
    distinct()
}

#' Title
#'
#' @param df
#' @param params_included
#'
#' @returns
#' @export
#'
#' @examples
summarise_z <- function(
    df,
    params_included
    ){

  df_weight_grid <- create_weight_grid(x = params_included, wt_vals = c(0,seq(0.1, 1, by = 0.1)))

  df_filt <- df |>
    filter(
      parameter %in% params_included
    )
  split(df_filt,df_filt$pub_mo_label) |>
    map(\(dft){

      split(
        df_weight_grid,
        df_weight_grid$wt_id
      ) |>
        map(
          \(dft_w){

            dft_weighted <- dft |>
              left_join(
                dft_w, by = "parameter"
              )
            dft_summarised <- dft_weighted |>
              group_by(
                yr_season ,
                pub_mo_label,
                adm1_name,
                pub_mo_date,wt_id
              ) |>
              summarise(
                zscore = weighted.mean(zscore,w=weight,na.rm=T)
                ,.groups="drop"
              ) |>
              mutate(
                wt_set = list(dft_w)
              )
          }
        ) |>
        list_rbind()
    }
    ) |>
    list_rbind()

}



#' Title
#'
#' @param df
#' @param params_included
#' @param earliest_year
#' @param rp
#'
#' @returns
#' @export
#'
#' @examples
weighted_classify <- function(df,
                              params_included,
                              earliest_year,
                              rp=3 ){
  df_normalized <- df |>
    normalize_to_z(
      params_included = params_included,
      earliest_year = earliest_year
    )

  df_truth <- df_normalized |>
    extract_truth_set()

  df_z_weighted <- df_normalized |>
    summarise_z(params_included = params_included)

   df_z_weighted |>
      left_join(
        df_truth |>
          select(-value)
      ) |>
      utils$threshold_var(
        var= "zscore",
        by = c("pub_mo_label","adm1_name","wt_id"),
        rp_threshold = rp
      ) |>
      utils$threshold_var(
        var= "zscore_asi_Jun",
        by = c("pub_mo_label","adm1_name","wt_id"),
        rp_threshold = rp
      ) |>
      arrange(adm1_name, yr_season,wt_id) |>
      ungroup()

}

single_indicator_performance <- function(df, parameter,earliest_year,rp){
  df_normalized <- df |>
    normalize_to_z(
      params_included = parameter,
      earliest_year = earliest_year
    )

  df_truth <- df_normalized |>
    extract_truth_set()



  df_classified <- df_normalized |>
    left_join(
      df_truth |>
        select(-value)
    ) |>
    utils$threshold_var(
      var= "zscore",
      by = c("pub_mo_label","adm1_name"),
      rp_threshold = rp
    ) |>
    utils$threshold_var(
      var= "zscore_asi_Jun",
      by = c("pub_mo_label","adm1_name"),
      rp_threshold = rp
    ) |>
    arrange(adm1_name, yr_season) |>
    ungroup()
  summarise_performance(df_classified,by = c("pub_mo_label","adm1_name"))
}
#' Title
#'
#' @param df
#' @param by
#'
#' @returns
#' @export
#'
#' @examples
summarise_performance <-  function(
    df,
    by=c("pub_mo_label","adm1_name","wt_id","wt_set"),
    parameter_subset = NULL
    ){
  if(!is.null(parameter_subset)){
    df |>
      filter(
        df_all_weights_flagged
      )
  }
  df |>
    mutate(
      across(ends_with("_flag"),\(x) factor(x,levels = c("TRUE","FALSE")))
    ) |>
    group_by(
      across({{by}})
    ) |>
    f_meas(zscore_asi_Jun_flag, zscore_flag, estimator= "binary",event_level = "first") |>
    ungroup()

}

#' Title
#'
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
top_performance_per_moment <- function(df,n=1){
  # df=  ldf_perf_all_models$gte1984,n=1
  df_max <- df |>
    group_by(pub_mo_label, wt_id) |>
    summarise(
      avg_estimate = mean(.estimate)
    ) |>
    slice_max(
      order_by = avg_estimate,
      n= n
    )|>
    ungroup()

  inner_join(
    df_max,
    select(df,
           any_of(c("adm1_name","pub_mo_label", "wt_id", "wt_set",".estimate","asi_f1"))
           ),
    by = c("pub_mo_label","wt_id")
  )
}

#' Title
#'
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
plot_optimal_compositions <-  function(df, label_plot = F){

  df_labelled <- df|>
    select(
      pub_mo_label,adm1_name,wt_set,avg_estimate,.estimate
    ) |>
    unnest(wt_set) |>
    label_parameters() |>
    mutate(
      p_label = glue(
      "id: {wt_id}
      f1: {scales::label_number(accuracy =0.001)(.estimate)}
      avg f1: {scales::label_number(accuracy =0.001)(avg_estimate)}"
      ),
      weight_pct_label = scales::percent(weight,accuracy = 1, trim = FALSE),
    )

  p <- df_labelled |>
    group_by(pub_mo_label, adm1_name) |>
    mutate(
      pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
      id = dense_rank(wt_id)
    ) |>
    ungroup() |>
    ggplot(
      aes(x= id, y= weight,fill = parameter_label)
    )+
    geom_bar(
      stat= "identity", color = "black"
    )+

    scale_fill_manual(values=rand_pal)+
    facet_grid(cols= vars(pub_mo_facet),
               rows = vars(adm1_name)
               ,scales="free")+
    labs(
      x = "Different indicator weightings"

    )+
    scale_y_continuous(labels=scales::label_percent())+
    theme(
      legend.title = element_blank(),
      axis.text.x = element_blank()
    )
  if(label_plot){
    # df <- ldf_tops2$gte1984
    df |>
      mutate(
        p_label = glue("{wt_id}: {scales::label_number(accuracy =0.01)(avg_estimate)}")
      )
    p <- p +
      # composition plot
      geom_text(
        aes(x= id, y= weight, label = weight_pct_label),
        position = position_stack(vjust = 0.5), color ="black"
      )+
      # wt plot
      geom_label(aes(x= id, y= 1.1, label = p_label), color ="black",fill="beige",alpha=0.4)+
      # theme(
      #   panel.spacing.y = unit(4, "lines")
      # )+
      expand_limits(y = 1.2)
  }
  p
}



#' Title
#' helper func to find weight ids that are simple and optimal
#' not a perfect solution, but gets us 90% there.
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
plot_low_var_optimal <-  function(df){
  dfp <- df|>
  select(
    pub_mo_label, wt_set
  ) |>
  unnest(wt_set) |>
    group_by(
      pub_mo_label,wt_id
    ) |>
    mutate(
      sd = sd(weight)

    ) |>
    group_by(pub_mo_label) |>
  slice_min(sd) |>
  group_by(pub_mo_label) |>
  mutate(
    pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
    id = dense_rank(wt_id)
  ) |>
  ungroup() |>
  label_parameters()

  dfp |>
  distinct() |>
  mutate(
    weight_pct_label = scales::percent(weight,accuracy = 1, trim = FALSE),
    weight_id_label = glue("weight id: {wt_id}")

  ) |>
  filter(pub_mo_label!= "Jun") |>
  ggplot(
    aes(x= id, y= weight,fill = parameter_label), position = position_stack(vjust = 0.5)
  )+
  geom_bar(
    stat= "identity"
  )+
  geom_text(aes(x= id, y= weight, label = weight_pct_label),
            position = position_stack(vjust = 0.5), color ="white")+
  geom_text(aes(x= id, y= 1.01, label = weight_id_label), color ="black")+
  scale_fill_manual(values=rand_pal)+
  facet_grid(cols= vars(pub_mo_facet)
             ,scales="free")+
  labs(
    # title = "May publication weight combos that give avg 0.90 f1"
  )
}


#' Title
#'
#' @param df
#' @param param_simple
#'
#' @returns
#' @export
#'
#' @examples
get_simple_weight_id <- function(df,param_simple){
    df |>
    select(wt_set) |>
    unnest(wt_set) |>
    distinct() |>
    filter(
      parameter == param_simple ,
      weight == 1
    ) |>
    pull(wt_id) |>
    unique()

}
#' Title
#'
#' @param df
#' @param wt_id_apr
#' @param wt_id_may
#' @param param_simple
#'
#' @returns
#' @export
#'
#' @examples
compare_to_simple_model <- function(df,wt_id_apr, wt_id_may,param_simple){
  simple_model_id <-get_simple_weight_id(df= df, param_simple = param_simple)

    df_chosen <- df |>
      summarise_performance() |>
      filter(
        (pub_mo_label == "Apr" & wt_id==wt_id_apr)|
        (pub_mo_label == "May" & wt_id==wt_id_may)|
        wt_id == simple_model_id

      ) |>
      filter(pub_mo_label != "Jun")

    df_simple <- df_chosen |>
      filter(
        wt_id == simple_model_id
      )
    anti_join(df_chosen,df_simple) |>
      left_join(
        df_simple |>
          select(
            pub_mo_label,adm1_name,
            estimate_simple =.estimate),
        by = c("pub_mo_label","adm1_name")
      )

}

# setup as many sets of parameters as we want... currently we have 2
# One that is limited from data gte 2000 because of MODIS
params_set_gte_2001 <- c(
  "era5_land_soil_moisture_1m",
  "cumu_era5_land_total_precipitation_sum",
  "NDSI_Snow_Cover_mean",
  "vhi",
  "asi"
)

# and one that takes data going back to 1984
params_set_gte_1984 <- c(
  "era5_land_soil_moisture_1m",
  "cumu_era5_land_total_precipitation_sum",
  "vhi",
  "era5_land_snow_cover", # toying w/ adding this or not
  "asi"
)
params_set_gte_1984_no_snow <- c(
  "era5_land_soil_moisture_1m",
  "cumu_era5_land_total_precipitation_sum",
  "vhi",
    # "era5_land_snow_cover", # toying w/ adding this or not
  "asi"
)

l_params <- list(
  "gte2001"= params_set_gte_2001,
  "gte1984" = params_set_gte_1984,
  "gte_1984_no_snow" = params_set_gte_1984_no_snow
)

df_parameter_configs <- tibble(
  param_set_id = names(l_params),
  parameter_sets = l_params,
  earliest_year =parse_number(param_set_id)
)

box::reload(utils)
box::use(yardstick[...])

ldf_perf <- pmap(
  list(
    df_parameter_configs$param_set_id,
    df_parameter_configs$parameter_sets,
    df_parameter_configs$earliest_year
    ),
  \(id,params, start_yr){

    # cat("reclassifying\n")
    df_all_weights_flagged <- weighted_classify(
      df = df_env_model,
      earliest_year = start_yr,
      params_included = unlist(params),
      rp=3
    )

    df_asi_perf <-  single_indicator_performance(
      df_env_model,
      parameter = "asi",
      earliest_year = start_yr,
      rp=3
    )


    df_perf <- df_all_weights_flagged |>
      summarise_performance() |>
       left_join(
        df_asi_perf |>
          rename(asi_f1 = .estimate)
        )


    df_top_perf <- df_perf |>
      top_performance_per_moment()
      # left_join(
      #   df_asi_perf |>
      #     rename(asi_f1 = .estimate)
      #   )

    list(
      ALL_MODELS = df_perf,
      TOP_MODELS = df_top_perf,
      HISTORICAL_WEIGHTED = df_all_weights_flagged
    )

  }
)

ldf_perf_all_models <- ldf_perf |>
  map("ALL_MODELS") |>
   set_names(
    df_parameter_configs$param_set_id
  )


ldf_top3 <- ldf_perf |>
  map("ALL_MODELS") |>
  map(\ (dft){
    dft |>
      top_performance_per_moment(n=3)
  }) |>
  set_names(
    df_parameter_configs$param_set_id
  )

ldf_historical <-  ldf_perf |>
  map("HISTORICAL_WEIGHTED") |>
   set_names(
    df_parameter_configs$param_set_id
  )


ldf_top3$gte1984 |>
  filter(pub_mo_label!="Jun") |>
  group_by(pub_mo_label, adm1_name) |>
  slice_max(avg_estimate,with_ties = F) |>
  # left_join(df_asi_performance |> rename(asi_f1 = .estimate)) |>
  pivot_longer (cols = c(".estimate","asi_f1")) |>
  mutate(
    indicator_label = ifelse(name == ".estimate", "Combined Indicator","Simple Indicator (ASI)" )
  ) |>
  ggplot(
    aes(x = adm1_name, y =value, fill = indicator_label)
  )+
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Indicator (ASI)"=hdx_hex("sapphire-light"))
  )+
  facet_wrap(~pub_mo_label)+
  scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: Composite Indicator vs ASI Only",
    subtitle= "Analysis 1984-present (including ERA5 snow)",
    caption= "Composite indicator contains optimal weighted aggregations of Snow Cover (ERA5), Soil Moisture (ERA5), ASI (FAO), VHI (FAO), and cumulative precipitation (ERA 5)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )
plot_low_var_optimal(
  ldf_top3$gte1984
)

plot_optimal_compositions(
  ldf_top3$gte1984  |>
    filter(pub_mo_label != "Jun"),label_plot = T
  ) +
  labs(
      title ="Indicator weighting options that result in max performance"
    )


df_compare <- compare_to_simple_model(
  ldf_historical$gte1984,
  wt_id_apr = 379, #14 is optimal
  wt_id_may = 142, #142 & 84 are optimal for May, but dont include soil moisture
  param_simple = "asi"
  )

df_compare |>
  pivot_longer(cols = c(".estimate","estimate_simple")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","Simple Model (ASI ONLY)")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge", color = "lightgrey") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Model (ASI ONLY)"=hdx_hex("sapphire-light"))
  )+
  labs(
    title = "F1 Score Comparison: May Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  facet_wrap(~pub_mo_label)+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )




ldf_top3$gte1984 |>
  group_by(pub_mo_label,adm1_name) |>
  slice_max(avg_estimate, n=1)

plot_optimal_compositions(
  ldf_top3$gte1984  |>
    filter(pub_mo_label != "Jun") |>
    group_by(pub_mo_label,adm1_name) |>
  slice_max(avg_estimate, n=2)


  ,


  label_plot = T
  ) +
  labs(
      title ="Indicator weighting options that result in max performance"
    )

plot_optimal_compositions(
  ldf_top3$gte1984  |>
    filter(
      pub_mo_label != "Jun",
      (pub_mo_label == "May"&wt_id%in%c(142,84))|(pub_mo_label=="Apr"& wt_id==14)
      ),label_plot = T
  ) +
  labs(
      title ="Indicator weighting options that result in max performance"
    )

simple_model_id <- get_simple_weight_id(
  df= ldf_historical$gte1984,
  param_simple = "asi"
  )


optimal_model_wt_id <-  c(14,379)[1]
df_simple_historical <- ldf_historical$gte1984 |>
  select(
    yr_season,wt_id,wt_set, pub_mo_label, adm1_name,ends_with("flag")

  ) |>
  pivot_longer(cols = ends_with("flag")) |>
  filter(wt_id ==simple_model_id,name== "zscore_flag") |>
  mutate(
    weight_type = "Simple (ASI only)"
  )

df_historical_weights_sel <- ldf_historical$gte1984 |>
  select(
    yr_season,wt_id,wt_set, pub_mo_label, adm1_name,ends_with("flag")

  ) |>
  pivot_longer(cols = ends_with("flag")) |>
  filter(
    wt_id == optimal_model_wt_id & pub_mo_label=="Apr"
  ) |>
  mutate(
    weight_type =ifelse(name == "zscore_flag","Combined","EOS ASI\n'Truth'")
  )


bind_rows(
  df_historical_weights_sel,
  df_simple_historical
) |>
  filter(adm1_name == "Takhar") |>
  mutate(
    yr = year(yr_season),
    weight_type = factor(weight_type, c("Simple (ASI only)","Combined","EOS ASI\n'Truth'"))
  ) |>
  ggplot()+
  geom_tile(
    aes(x = weight_type,y= yr,  fill = value), color = "white",linewidth = 0.7
  ) +
  scale_fill_manual(
    values = c("TRUE"=hdx_hex("tomato-hdx"), "FALSE"="grey")
  )+
  scale_y_reverse(breaks= 1984:2024,
                  expand= c(0,0)
                  # limits = c(2024,1984)
                  )+
  scale_x_discrete(expand =c(0,0))+
  labs(
    title = "Historical analyses of indicators",
    subtitle = "April - Takhar Province",
    caption = glue("optimal model id: {optimal_model_wt_id}")
  )+
  facet_wrap(~adm1_name)+
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )

df_cerf <- cumulus::blob_read("ds-aa-afg-drought/raw/vector/afg_drought_cerf_allocations.xlsx") |>
  clean_names() |>
  mutate(
    allocation_date = as_date(allocation_date)
  )
df_cerf_yr <-  df_cerf |>
  group_by(yr = year(allocation_date)) |>
  summarise(
    n=n()
  ) |>
  complete(
    yr = 2006:2024,
    fill = list(n=0)
  )


rp_compare <- 3
df_eos_asi_flagged <- ldf_historical$gte1984 |>
  select(yr_season,
         strata= adm1_name,
         zscore_asi_Jun,
         ) |>
  # select(yr_season, strata= adm1_name,value= zscore_asi_Jun_flag) |>
  distinct() |>
  utils$threshold_var(
    var="zscore_asi_Jun",
    by= c("strata"),
    rp_threshold = rp_compare
  ) |>
    mutate(
    yr = year(yr_season)
  ) |>
  rename(value=zscore_asi_Jun_flag)



bind_rows(
df_eos_asi_flagged,
  df_cerf_yr |>
    mutate(
      strata = "CERF Allocations",
      value = n>0
    )
) |>
  mutate(
    value_fill = case_when(
      value==F ~ "no activations",
      value == T & strata != "CERF Allocations"~"activation",
      value == T  & strata == "CERF Allocations" & n ==1~ "1 activation",
      value == T & strata == "CERF Allocations" & n ==2~ "2 activations",
      .default = "?"

    ),
    strata = factor(strata, c("Faryab","Sar-e-Pul","Takhar","CERF Allocations")),
  ) |>
  ggplot(
    aes(x= strata, y= yr, fill = value_fill)
  )+
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c("no activations"=hdx_hex("grey-medium"),
               "activation" = hdx_hex("tomato-light"),
               "1 activation"=hdx_hex("tomato-hdx"),
               "2 activations"= hdx_hex("tomato-dark"))
  )+
  scale_y_reverse(expand= c(0,0), breaks= 1984:2024)+
  scale_x_discrete(expand= c(0,0))+
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )+
  labs(
    title = glue("End of Season ASI - Thresholded at {rp_compare} year RP and Historical CERF Activations")
  )


ldf_top3$gte2001 |>
  filter(pub_mo_label!="Jun") |>
    group_by(pub_mo_label, adm1_name) |>
  slice_max(avg_estimate,with_ties = F) |>
  # left_join(df_asi_performance |> rename(asi_f1 = .estimate)) |>
  pivot_longer (cols = c(".estimate","asi_f1")) |>
  mutate(
    indicator_label = ifelse(name == ".estimate", "Combined Indicator","Simple Indicator (ASI)" )
  ) |>
  ggplot(
    aes(x = adm1_name, y =value, fill = indicator_label)
  )+
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Indicator (ASI)"=hdx_hex("sapphire-light"))
  )+
  facet_wrap(~pub_mo_label)+
  scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: Composite Indicator vs ASI Only",
    subtitle= "Analysis 2001-present (including MODIS Snow -NDSI)",
    caption= "Composite indicator contains optimal weighted aggregations of NDSI Snow Cover (MODIS), Soil Moisture (ERA5), ASI (FAO), VHI (FAO), and cumulative precipitation (ERA 5)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

ldf_top3$gte_1984_no_snow |>
  filter(pub_mo_label!="Jun") |>
    group_by(pub_mo_label, adm1_name) |>
  slice_max(avg_estimate,with_ties = F) |>
  # left_join(df_asi_performance |> rename(asi_f1 = .estimate)) |>
  pivot_longer (cols = c(".estimate","asi_f1")) |>
  mutate(
    indicator_label = ifelse(name == ".estimate", "Combined Indicator","Simple Indicator (ASI)" )
  ) |>
  ggplot(
    aes(x = adm1_name, y =value, fill = indicator_label)
  )+
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Indicator (ASI)"=hdx_hex("sapphire-light"))
  )+
  facet_wrap(~pub_mo_label)+
  scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: Composite Indicator vs ASI Only",
    subtitle= "Analysis 1984-Present (No Snow included)",
    caption= "Composite indicator contains optimal weighted aggregations of  Soil Moisture, ASI, VHI, and cumulative precipitation",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

ldf_historical$gte1984 |>
  group_by(
    adm1_name,
    #yr_season,
    pub_mo_label,
    wt_id,
    wt_set,
  ) |>
  summarise(
    activation_rate_wt = mean(zscore_flag),
    activation_rate_truth = mean(zscore_asi_Jun_flag)
  )
ldf_historical$gte2001 |>
  group_by(
    adm1_name,
    #yr_season,
    pub_mo_label,
    wt_id,
    wt_set,
  ) |>
  summarise(
    activation_rate_wt = mean(zscore_flag),
    activation_rate_truth = mean(zscore_asi_Jun_flag)
  )

ldf_historical$gte_1984_no_snow |>
  group_by(
    adm1_name,
    #yr_season,
    pub_mo_label,
    wt_id,
    wt_set,
  ) |>
  summarise(
    activation_rate_wt = mean(zscore_flag),
    activation_rate_truth = mean(zscore_asi_Jun_flag)
  )





df_truth_joint <- df_all_weights_flagged |>
  distinct(
    yr_season,adm1_name, zscore_asi_Jun_flag
  ) |>
  group_by(
    yr = year(yr_season)
  ) |>
  summarise(
    flag = any(zscore_asi_Jun_flag)
  ) |>
  mutate(
    indicator_type = "Truth"
  )

 df_apr_joint <- df_all_weights_flagged |>
  filter(
    (pub_mo_label == "Apr" &
    wt_id %in% c(1,684)
    )) |>
    mutate(
      yr= year(yr_season),
    indicator_type = case_when(
      wt_id == 1 ~ "Simple (ASI only)",
      # name== "zscore_asi_Jun_flag"~"Truth",
      .default= "Composite",
      # .default = "combined"

  )
    ) |>
   select(yr,indicator_type, pub_mo_label, adm1_name,zscore_flag) |>
   group_by(
     yr, indicator_type, pub_mo_label
   ) |>
   summarise(
     flag = any(zscore_flag)
   )
df_apr_joint |>
  bind_rows(
    df_truth_joint
  ) |>
  group_by(
    indicator_type
  ) |>
  summarise(
    flag = mean(flag)
  )
df_apr_joint |>
  bind_rows(
    df_truth_joint
  ) |>
  ggplot()+
  geom_tile(
    aes(x = indicator_type,y= yr,  fill = flag), color = "white"
  ) +
  scale_fill_manual(
    values = c("TRUE"=hdx_hex("tomato-hdx"), "FALSE"="grey")
  )+
  labs(
    title = "April Joint"
  )

df_historical_joint_admins <- df_all_weights_flagged |>
  filter(
    (pub_mo_label == "Apr" &
    wt_id %in% c(1,684)
    )|
      (pub_mo_label =="May" &
         wt_id %in% c(1, 678))
  ) |>
  mutate(
    yr= year(yr_season)
  ) |>
  group_by(wt_id,yr, pub_mo_label) |>
  summarise(
    zscore_flag= any(zscore_flag)
  ) |>
  # pivot_longer(
  #   cols = c("zscore_flag","zscore_asi_Jun_flag")
  # ) |>
  mutate(
    indicator_type = case_when(
      wt_id == 1 ~ "Simple (ASI only)",
      # name== "zscore_asi_Jun_flag"~"Truth",
      .default= "Combined",
      # .default = "combined"

  ),
  adm1_name = "all_admins"
  )
left_join(df_historical_joint_admins,df_truth_joint, by ="yr") |>
  pivot_longer(cols = c("zscore_flag","truth_flag")) |>
   ggplot()+
  geom_tile(
    aes(x = name,y= yr,  fill = value), color = "white"
  ) +
  scale_fill_manual(
    values = c("TRUE"=hdx_hex("tomato-hdx"), "FALSE"="grey")
  )+
  facet_wrap(~pub_mo_label)

bind_rows(df_historical_joint_admins,df_truth_joint) |>
  ggplot()+
  geom_tile(
    aes(x = name,y= yr,  fill = zscore_flag), color = "white"
  ) +
  scale_fill_manual(
    values = c("TRUE"=hdx_hex("tomato-hdx"), "FALSE"="grey")
  )+
  facet_grid(
    cols = vars(adm1_name),
    # rows = vars(pub_mo_label)
  )

df_all_weights_flagged |>
  filter(
    (pub_mo_label == "Apr" &
    wt_id %in% c(1,684)
    )|
      (pub_mo_label =="May" &
         wt_id %in% c(1, 678))
  ) |>
  mutate(
    yr= year(yr_season)
  ) |>
  select(yr,pub_mo_label,adm1_name,wt_id,zscore_flag,zscore_asi_Jun_flag ) |>
  pivot_wider(names_from = wt_id, values_from=zscore_flag) |>
  pivot_longer(zscore_asi_Jun_flag:`684`,values_drop_na = TRUE) |>
  mutate(
    name = case_when(
      name == "1" ~ "Simple (ASI only)",
      name== "zscore_asi_Jun_flag"~"Truth",
      .default= "Combined",
      # .default = "combined"
      ),

    # yr= year(yr_season)
  ) |>
  bind_rows(
    df_historical_joint_admins
  )
  mutate(
        name = factor(name,c("Simple (ASI only)","Combined","Truth" ))
  )
  ggplot()+
  geom_tile(
    aes(x = name,y= yr,  fill = value), color = "white"
  ) +
  scale_fill_manual(
    values = c("TRUE"=hdx_hex("tomato-hdx"), "FALSE"="grey")
  )+
  facet_grid(
    cols = vars(adm1_name),
    rows = vars(pub_mo_label)
  )



```



```{r}
# okay -- above we get best weights per adm and month.
# can we instead test each set of weights against all provinces and get the highest f1 score
# that works per weight

l_top_wt_sets_per_mo <- ldf_conf_metrics  |>
  map(
    \(dft){
      dft |>
        group_by(pub_mo_label, wt_id) |>
        summarise(
          avg_estimate = mean(.estimate)
        ) |>
        slice_max(
          order_by = avg_estimate,
          n= 1
        )

    }

  )

ldf_best_metrics_overall<- map2(ldf_conf_metrics,l_top_wt_sets_per_mo,
  \(dft_conf,dft_top){
          dft_conf |>
        inner_join(
          dft_top, by = c("pub_mo_label","wt_id")
          )
  }
)
ldf_best_metrics_overall<- map(named_weight_sets,
  \(nmt){
    dft_conf <- ldf_conf_metrics[[nmt]]
    dft_top <- l_top_wt_sets_per_mo[[nmt]]
    dft_conf |>
        inner_join(
          dft_top, by = c("pub_mo_label","wt_id")
          )
  }
)
rand_pal <- c("ASI"="green4",
  "VHI"="#CAB2D6",
  "NDSI"= "#6A3D9A",
  "Precip cumu (ERA)"= "skyblue2",
  "era5_land_soil_moisture_1m"= "dodgerblue2",
  "Snow Cover (ERA5)"= "white"
)

ldf_top_weights_nested <-  map2(l_top_wt_sets_per_mo,l_weight_combos,
     \(df_top,df_wt_set){
     df_top_wt_vals <- df_top |>
         inner_join(
           df_wt_set |>
             group_by(wt_id) |>
             nest()
           # by = "wt_id"
           # wt_id %in% df_top$wt_id
         )
         # label_parameters()
     }
 )


ldf_best_metrics_overall$gte1984 |> select(wt_set) |> unnest(wt_set) |> count(parameter)



lp_optimal_weight_composition <- map(ldf_best_metrics_overall,
     \(dft){

       dft <- dft|>
         select(
           pub_mo_label,adm1_name,wt_set,avg_estimate,.estimate
         ) |>
         unnest(wt_set) |>
         label_parameters()

       dft |>
         group_by(pub_mo_label, adm1_name) |>
         mutate(
           pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
           id = dense_rank(wt_id)
         ) |>
         ungroup() |>
         ggplot(
           aes(x= id, y= weight,fill = parameter_label)
         )+
         geom_bar(
           stat= "identity", color = "grey"
         )+

         scale_fill_manual(values=rand_pal)+
         facet_grid(cols= vars(pub_mo_facet),
                    rows = vars(adm1_name)
                    ,scales="free")+
         labs(
           # title = "May publication weight combos that give avg 0.90 f1"
         )


     }
)
lp_optimal_weight_composition$gte2000
lp_optimal_weight_composition$gte1984

lp_weight_performance_per_adm <- map(
  ldf_best_metrics_overall,\(dft){

         dft <- dft |>
           mutate(
             pub_mo_facet = factor(pub_mo_label, levels = c("Apr","May","Jun"))
           )

         dft |>
           ggplot(
             aes(
               x= adm1_name,
               y= .estimate,
               group = wt_id
               # color =as_factor(wt_id)
             )
           )+
           geom_point(alpha=0.3,size=4)+
           geom_line(alpha=0.3)+
           facet_wrap(~pub_mo_facet)+
           geom_hline(
             data = dft |>
               select(
                 pub_mo_facet,
                 adm1_name,
                 avg_estimate
                 ) |> distinct(),
             aes(yintercept = avg_estimate),
             linetype = "dashed",
             color = hdx_hex("tomato-hdx")
           )+
           labs(
             y= "F1 Performance"
           )+
           theme(
             panel.border = element_rect(fill=NA, color = "grey")
           )

     }
)


lp_weight_performance_per_adm$gte2000
lp_weight_performance_per_adm$gte1984



# i like these, but problem is faryab actually performs worse
dft_low_sd_f1 <- ldf_best_metrics_overall$gte2000 |>
  select(
    pub_mo_label, wt_set
  ) |>
  unnest(wt_set) |>
  group_by(
    pub_mo_label,wt_id
  ) |>
  mutate(
    sd = sd(weight)
  ) |>
  group_by(pub_mo_label) |>
  slice_min(sd) |>
  group_by(pub_mo_label) |>
  mutate(
    pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
    id = dense_rank(wt_id)
  ) |>
  ungroup() |>
  label_parameters()

dft_low_sd_f1 |>
  distinct() |>
  mutate(
    weight_pct_label = scales::percent(weight,accuracy = 1, trim = FALSE),
    weight_id_label = glue("weight id: {wt_id}")

  ) |>
  filter(pub_mo_label!= "Jun") |>
  ggplot(
    aes(x= id, y= weight,fill = parameter_label), position = position_stack(vjust = 0.5)
  )+
  geom_bar(
    stat= "identity"
  )+
  geom_text(aes(x= id, y= weight, label = weight_pct_label),
            position = position_stack(vjust = 0.5), color ="white")+
  geom_text(aes(x= id, y= 1.01, label = weight_id_label), color ="black")+
  scale_fill_manual(values=rand_pal)+
  facet_grid(cols= vars(pub_mo_facet)
             ,scales="free")+
  labs(
    # title = "May publication weight combos that give avg 0.90 f1"
  )


ldf_best_metrics_overall$gte2000 |>
  filter(
    pub_mo_label == "Apr",
         wt_id ==684 # if filtering to >20000
         ) |>
  left_join(
      ldf_conf_metrics$gte2000 |>
      filter(wt_id ==1) |>
      # select(-wt_id) |>
      rename(
        asi_only_f1 = .estimate
        ) |>
        select(pub_mo_label,adm1_name, asi_only_f1)
   , by = c("pub_mo_label","adm1_name")
  ) |>
  pivot_longer(cols = c(".estimate","asi_only_f1")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: April Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

ldf_best_metrics_overall$gte2000 |>
  filter(
    pub_mo_label == "May",
         wt_id ==524 # if filtering to >20000
         ) |>
  left_join(

      ldf_conf_metrics$gte2000 |>
      filter(wt_id ==1) |>
      # select(-wt_id) |>
      rename(
        asi_only_f1 = .estimate
        ) |>
        select(pub_mo_label,adm1_name, asi_only_f1)
   , by = c("pub_mo_label","adm1_name")
  ) |>
  pivot_longer(cols = c(".estimate","asi_only_f1")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: May Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )












df_best_wt_sets_w_metrics |>
  print(n=100) |>
  filter(pub_mo_label=="May") |>
  select(wt_set,avg_estimate) |>
  # pull(avg_estimate) |>
  unnest(wt_set) |>
  distinct() |>
  mutate(
    id = dense_rank(wt_id)
  ) |>
  ggplot(
    aes(x= id, y= weight,fill = parameter)
  )+
  geom_bar(
    stat= "identity"
  )+
  labs(
    title = "May publication weight combos that give avg 0.90 f1"
  )




top_wt_sets_per_mo <- df_conf_metrics  |>
  group_by(pub_mo_label, wt_id) |>
  summarise(
    avg_estimate = mean(.estimate)
  ) |>
  slice_max(
    order_by = avg_estimate,n= 1
  )
df_best_metrics_overall <- df_conf_metrics |>
  inner_join(
    top_wt_sets_per_mo, by = c("pub_mo_label","wt_id")
  )
df_best_wt_sets_w_metrics <- df_best_metrics_overall |>
  left_join(
    df_all_weights_flagged |>
      distinct(wt_id,wt_set)
    ) |>
  left_join(
      df_conf_metrics |>
      filter(wt_id ==1) |>
      # select(-wt_id) |>
      rename(
        asi_only_f1 = .estimate
        ) |>
        select(pub_mo_label,adm1_name, asi_only_f1)
   , by = c("pub_mo_label","adm1_name")
  )

df_best_wt_sets_w_metrics |>
  select(-wt_id) |>
  unnest(wt_set)


df_best_wt_sets_w_metrics |>
  filter(
    pub_mo_label == "Apr",
         wt_id ==346 # if filtering to >20000
         ) |>
  pivot_longer(cols = c(".estimate","asi_only_f1")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: April Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

df_wt_combos_long |>
  filter(wt_id  ==346)
```


Look at equal weigts

```{r}
df_conf_metrics |>
  filter(pub_mo_label == "Apr") |>
  filter(
    wt_id %in% c(1,equal_wts_id)
  ) |>
    mutate(
    name = ifelse(wt_id == equal_wts_id, "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= .estimate, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: April Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )
)

```

```{r}
df_all_weights_flagged
df_weighted_historical_sample <- df_all_weights_flagged |>
  filter(pub_mo_label == "Apr",adm1_name == "Takhar")

df_end_of_season_valid <- df_weighted_historical_sample |>
  select(yr_season,zscore_asi_Jun_flag) |>
  distinct() |>
  mutate(
    name = "ASI\nEnd of Season\n('truth')"
  ) |>
  rename(zscore_flag = zscore_asi_Jun_flag)

bind_rows(
  df_weighted_historical_sample ,
  df_end_of_season_valid
)

df_weighted_historical_sample |>
filter(
    wt_id %in% c(1,equal_wts_id,38)
  ) |>
  bind_rows(
    df_end_of_season_valid |>
      rename()
  ) |>
  mutate(
    yr = year(yr_season),
    name = case_when(
      wt_id == 1~"ASI only",
      wt_id == equal_wts_id~"Combined (equal)",
      wt_id == 38 ~"Combnied (optimal)",
      .default = name
      )
  ) |>
  ggplot()+
  geom_tile(
    aes(x = name,y= yr,  fill = zscore_flag), color = "white"
  ) +
  scale_fill_manual(
    values = c("TRUE"="red", "FALSE"="grey")
  )
```

```{r}

df_best_wt_sets_w_metrics |>
  filter(
    wt_id == 39
  ) |>
  select(-wt_id) |>
  unnest(wt_set)

df_best_wt_sets_w_metrics |>
  print(n=100) |>
  filter(pub_mo_label=="May") |>
  select(wt_set,avg_estimate) |>
  # pull(avg_estimate) |>
  unnest(wt_set) |>
  distinct() |>
  mutate(
    id = dense_rank(wt_id)
  ) |>
  ggplot(
    aes(x= id, y= weight,fill = parameter)
  )+
  geom_bar(
    stat= "identity"
  )+
  labs(
    title = "May publication weight combos that give avg 0.90 f1"
  )



df_conf_metrics |>
  filter(
    wt_id ==38 & pub_mo_label=="Apr"
  )




### Simlarity Metrics #######
xck_nest <- df_wt_combos_long |>
  group_by(wt_id) |>
  nest()

df_max_f1_apr  |>
  left_join(ck_nest) |>
  unnest(data)



ck <- df_max_f1_apr  |>
  left_join(ck_nest) |>
  unnest(data)



aligned_weights <- ck %>%
  group_by(adm1_name, wt_id, parameter) %>%
  summarise(weight = sum(weight), .groups = "drop") %>% # Aggregate weights if duplicates
  pivot_wider(names_from = parameter, values_from = weight, values_fill = 0) |>
  mutate(
    row_id = row_number()
  )

pairwise_comparisons <- aligned_weights %>%
  # mutate(row_id = row_number()) %>% # Add a unique ID to each row for pairing
  cross_join(aligned_weights, suffix = c("_x", "_y")) %>% # Generate all pairs
  filter(row_id_x < row_id_y) # Avoid duplicate comparisons and self-comparisons

pairwise_comparisons |>
  glimpse()
# Step 3: Compute similarity metrics
similarity_results |>
  glimpse()

similarity_results <- pairwise_comparisons %>%
  rowwise() %>%
  mutate(
    # Extract weight vectors using regex pattern
    weight_vector_x = list(c_across(matches("^(cumu.*_x|era5_land.*_x|NDSI.*_x|vhi_x|asi_x)$"))),
    weight_vector_y = list(c_across(matches("^(cumu.*_y|era5_land.*_y|NDSI.*_y|vhi_y|asi_y)$")))
  )

similarity_results <- similarity_results %>%
  rowwise() %>%
  mutate(
    # Calculate Euclidean distance
    manhattan_distance = sum(abs(unlist(weight_vector_x) - unlist(weight_vector_y)), na.rm = TRUE)
    # euclidean_distance = sqrt(sum((unlist(weight_vector_x) - unlist(weight_vector_y))^2, na.rm = TRUE)),
    # # Calculate Cosine similarity
    # cosine_similarity = sum(unlist(weight_vector_x) * unlist(weight_vector_y), na.rm = TRUE) /
    #   (sqrt(sum(unlist(weight_vector_x)^2, na.rm = TRUE)) * sqrt(sum(unlist(weight_vector_y)^2, na.rm = TRUE)))
  )

most_similar_weight_set <- similarity_results %>%
  group_by(wt_id_x) %>%
   summarise(
    # avg_jsd_distance = mean(jsd_distance, na.rm = TRUE),  # For JSD
    avg_manhattan_distance = mean(manhattan_distance, na.rm = TRUE)  # For Manhattan
  ) %>%
  # Rank by lowest JSD or Manhattan distance (lower is better)
  arrange(avg_manhattan_distance) %>%
  slice(1)

df_agg_all_weight_combos
df_wt_combos_long |>
  filter(wt_id == 883)

# Step 4: Select relevant columns for clarity
final_results <- similarity_results %>%
  select(
    adm1_name_x, wt_id_x,
    adm1_name_y, wt_id_y,
    euclidean_distance, cosine_similarity
  )

# View results
final_results |>
  ungroup() |>
  filter(adm1_name_x!= adm1_name_y)
############


id_simple_asi_model <- df_wt_combos_long |>
  filter(
    parameter == "asi",
    weight ==1
  ) |>
  pull(
    wt_id
  )

df_simple_asi_weight_set <- df_wt_combos_long |>
  filter(wt_id %in% id_simple_asi_model) |>
  group_by(wt_id) |>
  nest()




df_weight_sets_nested <- df_wt_combos_long |>
  group_by(wt_id) |>
  nest()



df_max_f1 |>
  group_by(
    pub_mo_label, wt_id
  ) |>
  mutate(
    num_wt_sets = n()
  ) |>
  filter(num_wt_sets>1)

df_max_f1_co_occur <- df_max_f1  |>
  group_by(
    pub_mo_label, wt_id
  ) |>
  mutate(
    num_wt_sets = n()
  ) |>
  ungroup() |>
  filter(num_wt_sets>1) |>
  left_join(df_weight_sets_nested, by ="wt_id") |>
  group_by(adm1_name,pub_mo_label) |>
  unnest(data)

# we have 3 clear options for June - def easiest to just go w/ ASI and/or VHI
df_max_f1_co_occur |>
  ungroup() |>
  filter(num_wt_sets>2) |>
  print(n=45) |>
  distinct(pub_mo_label,wt_id, parameter,weight)

df_max_f1_co_occur |>
  print(n=45) |>
  distinct(pub_mo_label,wt_id, parameter,weight)

df_max_f1_shared <- df_max_f1  |>
  # group_by(
  #   pub_mo_label, wt_id
  # ) |>
  # mutate(
  #   num_wt_sets = n()
  # ) |>
  # ungroup() |>
  # filter(num_wt_sets>1) |>
  left_join(df_weight_sets_nested, by ="wt_id") |>
  group_by(adm1_name,pub_mo_label) |>
  unnest(data) |>
  mutate(
    wt_id_dense = dense_rank(wt_id)
  ) |>
  ungroup() |>
  mutate(
    pub_mo_label = fct_relevel(pub_mo_label, "Apr","May","Jun")
  )


df_f1_labels <- df_max_f1_shared |>
  distinct(pub_mo_label, adm1_name,.estimate) |>
    left_join(
    df_conf_metrics |>
      filter(wt_id ==1) |>
      group_by(pub_mo_label,adm1_name) |>
      transmute(
        simple_estimate = scales::label_number(accuracy = 0.01)(.estimate)
  ), by = c("pub_mo_label","adm1_name")
  ) |>
      mutate(
    pub_mo_label = fct_relevel(pub_mo_label, "Apr","May","Jun")
  )


df_max_f1_shared |>
    mutate(
    pub_mo_label = fct_relevel(pub_mo_label, "Apr","May","Jun")
  ) |>
# filter(pub_mo_label %in% c("May","Jun"), adm1_name == "Sar-e-Pul") |>
  ggplot(
  )+
  geom_bar(stat="identity", aes(x= wt_id_dense, y= weight, fill = parameter))+
  # geom_text(size = 3, position = position_stack(vjust = 0.5))
  # geom_text(
  #   aes(
  #     label = scales::label_percent(accuracy=1)(weight)
  #     ),
  #   position = position_stack(vjust = 0.5), color ="white"
  # )+
  geom_label(
    data= df_f1_labels,
    x= 50, y= 0.5,color="red",alpha = 0.6,
    aes(
        label = glue(
          "Max F1: {scales::label_number(accuracy=0.01)(.estimate)}\nASI Only F1: {simple_estimate}"

          )
    )
  )+
  facet_grid(
    cols = vars(adm1_name),
    rows = vars(pub_mo_label),
    # labeller = labeller(adm1_name=~facet_label),
    scales = "free_x")+
  labs(
    title = "Weight combinations for optimal F1 scores by data publication date and province",
    subtitle = "Optimal F1 Score & F1 Score of simple model (ASI only shown)"
  )+
  theme(
    axis.title.x =element_blank(),
    axis.text.x =element_blank(),
    # strip.switch.pad.grid = unit(1, "in")
  )


dft_sel <- dft |>
  pivot_longer(
    cols = -c("yr_season","pub_mo_label","adm1_name","pub_mo_date"),
    names_to = "parameter"
  ) |>
  filter(
    parameter %in% c("era5_land_soil_moisture_1m","cumu_era5_land_total_precipitation_sum","NDSI_Snow_Cover_mean","vhi","asi")
  ) |>
  group_by(
    across(-c("value","yr_season","pub_mo_date"))
  )
dft_sel |>
  distinct(parameter)
scale(dft_sel$value )[,1]
dft_scaled <- dft_sel |>
  mutate(
    value_centered = scale(value, center= T,scale=F)[,1],
    zscore = scale(value,center=T,scale=T)[,1],
    across(
      c("value_centered","zscore"),~ifelse(parameter != "asi",.x*-1,.x)
    )
  )

dft_scaled |>
  # filter(parameter == "asi") |>
  arrange(parameter)

# unweighted
dft_z_agg_unweighted <- dft_scaled |>
  group_by(
   yr_season ,
   pub_mo_label,
   adm1_name,
   pub_mo_date
  ) |>
  summarise(
    across(
      .cols = c("zscore","value_centered"),
      .fns = \(x) mean(x,na.rm=T)
    ),.groups="drop"
  )


dft_unweighted_pred <- dft_z_agg_unweighted |>
  left_join(
    df_jun_asi_scaled_short
  ) |>
  threshold_var(
    var= "zscore",
    by = c("pub_mo_label","adm1_name"),
    rp_threshold = 3
  ) |>
  threshold_var(
    var= "zscore_asi_Jun",
    by = c("pub_mo_label","adm1_name"),
    rp_threshold = 3
  ) |>
  arrange(adm1_name, yr_season) |>
  mutate(
    across(ends_with("_flag"),~as_factor(.x))
  ) |>
  ungroup()

recall(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag,event_level ="second")
precision(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag,event_level ="second")
f_meas(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag,event_level ="second")
metrics(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag)

```

## Experimental

### Tidymodels

```{r}
# silge random forest
# box::use(tidymodels[...])
df_env_compare <- df_env_compare_ck


AOI_SUBSET = c("Takhar","Sar-e-Pul","Faryab")

df_env_model <- #df_env_filt |>
  df_env_compare |>
  filter(
    !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max","era5_land_runoff_max"),
    !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d")
    # pub_mo_date<="2024-04-01" # last date of CHIRPS for now -- at some points we can update
  )

df_env_model |> count(parameter)
# df_env_filt <- df_environmental_yr_adj2 |>
#   filter(
#     pub_mo_date<="2024-04-01" # last date of CHIRPS for now -- at some points we can update
#   )

df_env_long <- df_env_model |>
  select(
    date,
    yr_season,
    pub_mo_date,
    adm1_name,
    parameter,
    value
    ) |>
  mutate(
    pub_mo_label = month(pub_mo_date,abbr=T, label =T)
  ) |>
  # count(parameter)
  # group_by(
  # pub_mo_date,
  #   adm1_name,
  #   parameter
  # ) |>
  # filter(date == max(date)) |>
  # ungroup() |>
  filter(
    # already have these from ERA land up to 2024 rather than 2020
    !(parameter %in% c("total_precipitation","mean_2m_air_temperature")),
    adm1_name %in% AOI_SUBSET
  )

# df_env_long |> count(parameter)
# df_params_long <- df_env_long |>
#   bind_rows(df_seas5_ready |>
#               select(any_of(colnames(df_env_long))) |>
#               filter(adm1_name%in% AOI_SUBSET))
#
# df_params_long |>
#   count(parameter)

df_env_long |>
  filter(str_detect(parameter,"^asi|^vhi")) |>
  group_by(parameter, pub_mo_label) |>
  count() |>
  pivot_wider(names_from=pub_mo_label, values_from = n)
date_ranges <- df_env_long |>
  group_by(parameter) |>
  summarise(
    start = min(pub_mo_date),
    end = max(pub_mo_date)
  )

# date_ranges |>
#   print(n= 100)
# date_ranges2 <- df_params_long |>
#   group_by(parameter) |>
#   summarise(
#     start = min(yr_season),
#     end = max(yr_season)
#   )


df_env_long |> count(parameter)
df_lh <- df_env_long#df_params_long #|>
  # filter(
  #   yr_season>= date_ranges[date_ranges$parameter=="asi",]$start,
  #   yr_season<= date_ranges[date_ranges$parameter=="asi",]$end
  # )
df_jun_asi <- df_lh |>
  filter(parameter == "asi",month(pub_mo_date)==6) |>
  select(
    yr_season,adm1_name, value
  ) |>
  rename(
    asi_Jun = value
  )

ldf_lh <- split(df_lh,df_lh$pub_mo_label) |>
    map(
      \(dft){
        dft |>
          pivot_wider(
            id_cols = c("yr_season","pub_mo_label","adm1_name","pub_mo_date"),
            names_from = parameter,
            values_from = value
          ) |>
          left_join(
            df_jun_asi,by = c("adm1_name","yr_season")
          )
      }
    )
ldf_lh |>
  map(
    ~colnames(.x)
  )

# why 3 missing asi_Jun, chirps_precipitiation_sum
ldf_lh |>
  map(\(dft){
    dft |>
  summarise(
    across(
      everything(),\(x) sum(is.na(x))
    )
  ) |>
  t()
  })
ldf_lh$Mar$vhi
ldf_lh$Mar |>
  filter(is.na(vhi)) |>
  glimpse()


ldf_model_full <- ldf_lh |>
  map(
    \(dft){
      dft  |>
        select(-starts_with("NDSI"))
    }
  )

ldf_model_short <- ldf_lh |>
  map(
    \(dft){
      dft |>
        filter(
          year(yr_season)>=2000
          # if_all(starts_with("NDSI"),\(x){!is.na(x)})
        )
    }
  )

ldf_model_short |>
  map(
    ~colnames(.x)
  )
```

#### Pre-Processing

```{r}
# month_prediction <- "May"

# ranger_mode <-  "classification"

ranger_pred <-  function(
    list_df,
    pub_mo,
    mode,
    rp_threshold=3,
    empirical = TRUE
){
  dft <- list_df[[pub_mo]]

  if(mode=="regression"){
    dft <- dft |>
      mutate(
        asi_outcome = log10(asi_Jun+0.1)
      )  |>
      select(-asi_Jun)
  }
  if(mode== "classification"){
    dft <- dft |>
      utils$threshold_var(var = "asi_Jun", by = c("pub_mo_label"),rp_threshold=rp_threshold) |>
      mutate(
        asi_Jun_flag = ifelse(asi_Jun_flag,"flag","no_flag")
      ) |>
      rename(
        asi_outcome ="asi_Jun_flag"
      )
  }

  if(!empirical){
     ranger_spec <-
    rand_forest(
      mtry = tune(),
      min_n = tune(),
      trees = 1000) %>%
    set_mode(mode) %>%
    set_engine("ranger")

    df_split <-   initial_split(
      dft ,
      prop = 7/10,
      strata = asi_outcome
    )
    df_train <-   training(df_split)
    df_test <- testing(df_split)
    df_resamples <- bootstraps(df_train,strata= asi_outcome,times= 25)

    ranger_recipe <-
      recipe(
        formula = asi_outcome ~ .,
        data = df_train # this is where training data should go, but just messing around
      ) |>
      step_rm(
        yr_season,
        pub_mo_date,
        pub_mo_label,
        asi_Jun
      )
    ranger_spec <-
      rand_forest(
        mtry = tune(),
        min_n = tune(),
        trees = 1000) %>%
      set_mode(mode) %>%
      set_engine("ranger")
  }

  if(empirical){
    ranger_spec  <-
      rand_forest(trees= 10000,mtry=11) |>
      set_mode(mode) |>
      set_engine("ranger")

    ranger_recipe <-
      recipe(
        formula = asi_outcome ~  .,
        data = dft
      ) |>
      step_rm(
        yr_season,
        pub_mo_date,
        pub_mo_label,
        asi_Jun

      )
  }

  rf_model_workflow <-
    workflow() %>%
    add_recipe(ranger_recipe) %>%
    add_model(ranger_spec)
  prediction_pred <- ifelse(mode=="classification",".pred_class",".pred")

  if(!empirical){
    ranger_tune <-
      tune_grid(rf_model_workflow, resamples = df_resamples, grid = 11)

    final_rf <- rf_model_workflow |>
      finalize_workflow(select_best(ranger_tune))

    df_fit <- last_fit(final_rf,df_split)

    df_ret <- collect_predictions(df_fit) |>
      rename(
        .pred_rf = prediction_pred
      )

    df_ret |>
    ggplot(
        aes(x=.pred_rf, y= asi_outcome )
      ) +
      geom_point()+
      geom_abline()

    df_pred_pre_metrics <- metrics(df_ret, asi_outcome, .pred_rf)

    imp_spec <-  ranger_spec |>
      finalize_model(select_best(ranger_tune)) |>
      set_engine("ranger",importance ="permutation")

    vip_plot <- workflow() |>
      add_recipe(ranger_recipe) |>
      add_model(imp_spec) |>
      fit(df_train) |>
      pull_workflow_fit() |>
      vip(aesthetics = list(alpha=0.8))
  }
  if(empirical){
    rf_fit <- fit(rf_model_workflow,data = dft)
    prediction_pred <- ifelse(mode=="classification",".pred_class",".pred")

    df_ret <- dft |>
      ungroup() |>
      mutate(
        .pred_rf = predict(rf_fit,new_data= dft)[[prediction_pred]]
      )

    df_pred_pre_metrics <- metrics(
      df_ret |>
      transmute(
        asi_outcome= factor(asi_outcome),
        .pred_rf= factor(.pred_rf)
      ),
      asi_outcome, .pred_rf
      )

    imp_spec <-  ranger_spec |>
      set_engine("ranger",importance ="permutation")
    vip_plot <- workflow() |>
      add_recipe(ranger_recipe) |>
      add_model(imp_spec) |>
      fit(dft) |>
      pull_workflow_fit() |>
      vip(aesthetics = list(alpha=0.8),num_features = 20)
  }

  accuracy <- df_pred_pre_metrics|>
    filter(.metric == "accuracy") |>
    pull(.estimate)

  vip_title <- ifelse(mode=="classification","Best {pub_mo} predictors of end of season ASI {rp_threshold} year return period activation","Best {pub_mo} predictors of end of season ASI {rp_threshold} values")
  vip_subtitle <-  "Model Accuracy: {scales::label_percent(accuracy = 1)(accuracy)}"

  vip_plot <- vip_plot  +
    labs(
      title = glue(vip_title),
      subtitle = glue(vip_subtitle)
    )
  list(
    df= df_ret, plot = vip_plot
  )
}
# debugonce(ranger_pred)
months_iter <-  set_names(month.abb[2:5],month.abb[2:5])

mo_temp <- months_iter[1]
mod_res <- map(months_iter,\(mo_temp){
  ranger_pred(
  list_df = ldf_model_short,# ldf_model_short
  pub_mo = mo_temp,
  mode = 'classification',
  empirical = T
  )
}
)
mod_res$Feb$plot
mod_res$Mar$plot
mod_res$Apr$plot
mod_res$May$plot

set.seed(100)
mod_res_pred <- map(months_iter,\(mo_temp){
  ranger_pred(
  list_df = ldf_model_full,
  pub_mo = mo_temp,
  mode = 'classification',
  empirical = F
  )
}
)

mod_res_pred$Feb
mod_res_pred$Mar
mod_res_pred$Apr
mod_res_pred$May
```

#### Model Full dataset

```{r}
# trying on full data
df_temp |> count(adm1_name)
month_prediction <- "May"

df_temp <- ldf_model_full[[month_prediction]] |>
  mutate(
    asi_Jun_trans = log10(asi_Jun+0.1)
  )



# df_temp <- split(df_temp,df_temp$adm1_name)
# df_temp<- df_temp$Takhar


df_temp |> colnames()


dummy_recipe <-
  recipe(
    formula = asi_Jun_trans ~  .,
    data = df_temp
    ) |>
  step_rm(
    yr_season,
    pub_mo_date,
    asi_Jun
    )



# so without all the tuning because that requires the data folds
rf_model_spec  <-
  rand_forest(trees= 10000,mtry=1) |>
  set_mode("regression") |>
  set_engine("ranger")

dtree_model_spec  <-
  decision_tree() |>
  set_mode("regression") |>
  set_engine("rpart")

rf_model_workflow <-
  workflow() |>
  add_recipe(dummy_recipe) |>
  add_model(rf_model_spec)

dtree_model_workflow <-
  workflow() |>
  add_recipe(dummy_recipe) |>
  add_model(dtree_model_spec)

set.seed(123)

rf_fit <- fit(rf_model_workflow,data = df_temp)
dtree_fit <- fit(dtree_model_workflow,data = df_temp)

df_temp_w_pred <- df_temp |>
  mutate(
    .pred_rf = predict(rf_fit,new_data= df_temp)$.pred,
    .pred_dtree = predict(dtree_fit, new_data= df_temp)$.pred
  )

p_pred120 <- df_temp_w_pred |>
  select(asi_Jun_trans,starts_with(".pred")) |>
  pivot_longer(
    # id_cols = "asi_Jun_trans",
    cols = starts_with(".pred")
  ) |>
  ggplot(
    aes(asi_Jun_trans,value, color= name)
  )+
  geom_point()+
  geom_abline()


df_temp_w_pred |>
  select(asi_Jun_trans,starts_with(".pred")) |>
  ggplot(
    aes(asi_Jun_trans,.pred_rf)
  )+
  geom_point()+
  geom_abline()+
  labs(
    title = glue("RF: {month_prediction}")
  )




# if decisoin _tree
tree_fit <- fit_ck %>%
  pull_workflow_fit()

library(rpart.plot)
rpart.plot(tree_fit$fit, roundint= FALSE)
p_pred120
p_pred

imp_spec <-  rf_model_spec |>
  set_engine("ranger",importance ="permutation")

workflow() |>
  add_recipe(dummy_recipe) |>
  add_model(imp_spec) |>
  fit(df_temp) |>
  pull_workflow_fit() |>
  vip(aesthetics = list(alpha=0.8))+
  labs(title= glue("RF: {month_prediction}"))
```

```{r}

df_temp <- ldf_model_short$Jun

all(df_temp$asi == df_temp$asi_Jun)
#> [1] TRUE

dummy_recipe <-
  recipe(
    formula = asi_Jun ~  .,
    data = df_temp
    ) |>
  step_rm(
    yr_season,
    pub_mo_date
    )



# so without all the tuning because that requires the data folds
model_spec  <-
  rand_forest(trees= 10000,mtry=1) |>
  set_mode("regression") |>
  set_engine("ranger")


model_workflow <-
  workflow() |>
  add_recipe(dummy_recipe) |>
  add_model(model_spec)

set.seed(123)
fit_ck <- fit(model_workflow,df_temp)



df_temp_w_pred <- df_temp |>
  mutate(
    .pred = predict(fit_ck,new_data= df_temp)$.pred
  )

df_temp_w_pred |>
  select(asi_Jun_trans,.pred)




df_temp_w_pred |>
  ggplot(
    aes(.pred,asi_Jun_trans)
  )+
  geom_point()+
  geom_abline()

fit_ck_simple <- rand_forest(
  trees = 1,
  mtry = 1,
  min_n=1
  ) |>
  set_mode("regression") |>
  set_engine("ranger") |>
  fit(asi_Jun ~ asi, data = df_temp)

df_temp_w_pred_simple <- df_temp |>
  mutate(
    .pred = predict(fit_ck_simple, new_data = df_temp)$.pred
  )

df_temp_w_pred_simple |>
  select(asi_Jun, .pred) |>
  print(n=20)

# A tibble: 72  2
#    asi_Jun  .pred
#      <dbl>  <dbl>
#  1  10.6   10.6
#  2   6.31   8.66
#  3   0.902  0.875
#  4  99.9   92.4
#  5   0      0
#  6   0      0
#  7  30.9   33.3
#  8   0      0
#  9  72.3   72.3
# 10   0      0
# 11   0      0
# 12  18.6   18.0
# 13   0      0
# 14  52.2   52.2
# 15   0.037  0.037
# 16   0      0
# 17   3.09   3.09
# 18   0      0
# 19  33.3   33.3
# 20  62.6   62.6

ranger_model <- ranger::ranger(
  asi_Jun ~ asi,
  data = df_temp,
  num.trees = 1,
  mtry = 1,
  min.node.size = 1
)
tree_info <- ranger::treeInfo(ranger_model)
print(tree_info)

library(randomForest)

df_temp2 <- df_temp |>
  mutate(
    asi = asi_Jun
  )
rf_model <- randomForest(asi_Jun ~ asi, data = df_temp2, ntree = 1)
predict(rf_model, df_temp)

identical(predictions, df_temp2$asi_Jun)

library(rpart)
all(df_temp2$asi==df_temp2$asi_Jun)
# TRUE
tree_model <- rpart(asi_Jun ~ asi, data = df_temp2)
tree_predictions <- predict(tree_model, df_temp2)

identical(tree_predictions, df_temp2$asi_Jun)
#> FALSE

print(tree_model)  # Examine tree structure
# n= 72
#
# node), split, n, deviance, yval
#       * denotes terminal node
#
# 1) root 72 56472.0800 18.961030
#   2) asi< 32.0665 53  3000.9780  3.993943
#     4) asi< 11.7045 46   248.5315  1.297196 *
#     5) asi>=11.7045 7   219.5563 21.715430 *
#   3) asi>=32.0665 19  8479.7370 60.711320 *
all.equal(tree_predictions, df_temp2$asi_Jun, tolerance = 1e-8)
#> "names for target but not for current" "Mean relative difference: 0.3331601"


lm_model <- lm(asi_Jun ~ asi, data = df_temp2)
lm_predictions <- predict(lm_model, df_temp2)
identical(lm_predictions, df_temp2$asi_Jun)
#> FALSE
all.equal(lm_predictions, df_temp2$asi_Jun, tolerance = 1e-8)
#> "names for target but not for current"

```

```{r}

ldf_model_short$Feb |>
  threshold_var(var = "asi_Jun", rp_threshold=3) |>
  glimpse()

collect_predictions(df_fit) |>
  ggplot(
    aes(x=asi_Jun_trans, y= .pred )
  ) +
  geom_point()+
  geom_abline()

sum(ldf_model_full$Feb$asi_Jun==0)/length(ldf_model_full$Feb$asi_Jun)
log10(ldf_model_full$Feb$asi_Jun)


```

```{r}
df_wide_scaled |>
  filter(!is.nan(asi_Jun)) |> count(adm1_name)
df_model_clean <- df_wide_scaled |>
  ungroup() |>
   filter(!is.na(asi_Jun),
          adm1_name!="Badghis",
          adm1_name == "Faryab",
          pub_mo_date>= "1984-02-01",
          pub_mo_label != "Jan"
          ) |>
  ungroup()

df_model_clean_timestep <- df_model_clean |>
  filter(pub_mo_label=="Feb")

prob_vars <- which(df_model_clean_timestep |>
  summarise(
    across(everything(), ~sum(is.na(.)))
  ) |>
  t()!=0)

df_model_clean2 <- df_model_clean_timestep[,-prob_vars] |>
  mutate(
    # strata = paste0(adm1_name,"_",pub_mo_label)
    strata = adm1_name
  ) |>
  select(-pub_mo_date, -pub_mo_label, - adm1_name,-pub_yr_date)



df_split <-   initial_split(
  df_model_clean2,
  # strata= strata
  )
df_train <-   training(df_split)
df_test <- testing(df_split)
# df_folds <- vfold_cv(df_train, strata= strata,v= nrow(df_))
# df_folds <- vfold_cv(df_model_clean2, strata= strata,v= nrow(df_model_clean2))
df_folds <- bootstraps(df_train, strata= asi_Jun)

# usemodels::use_ranger(asi_Jun~., data= df_train)

ranger_recipe <-
  recipe(formula = asi_Jun ~ ., data = df_train) %>%
  step_string2factor(one_of(NA_character_))

ranger_spec <-
  rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>%
  set_mode("regression") %>%
  set_engine("ranger")

ranger_workflow <-
  workflow() %>%
  add_recipe(ranger_recipe) %>%
  add_model(ranger_spec)


# set.seed(1681)
ranger_tune <-
  tune_grid(
    ranger_workflow,
    resamples = df_folds,
    grid = 11
  )

show_best(ranger_tune, metric="rmse")
# show_best(ranger_tune, metric="rsq")
autoplot(ranger_tune)

final_rf <- ranger_workflow |>
  finalize_workflow(select_best(ranger_tune))
df_fit <- last_fit(final_rf,df_split)
df_fit <- last_fit(final_rf,initial_split(df_model_clean2,0.03))



collect_metrics(df_fit)
collect_predictions(df_fit) |>
  ggplot(
    aes(x=asi_Jun, y= .pred )
  ) +
  geom_point()+
  geom_abline()


df_fit <- final_rf %>%
  fit(data = df_model_clean2)

# feature importance

imp_spec <-  ranger_spec |>
  finalize_model(select_best(ranger_tune)) |>
  set_engine("ranger",importance ="permutation")

workflow() |>
  add_recipe(ranger_recipe) |>
  add_model(imp_spec) |>
  fit(df_train) |>
  pull_workflow_fit() |>
  vip(aesthetics = list(alpha=0.8))
```

```{r}
library(tidymodels)
library(vip)  # For variable importance plots

mode <-  c("regression","classification")[2]
df_model_clean <- df_wide_scaled |>
  ungroup() |>
   filter(!is.na(asi_Jun),
          adm1_name!="Badghis",
          # adm1_name == "Faryab",
          pub_mo_date>= "1984-02-01"
          # pub_mo_label != "Jan"
          ) |>
  ungroup()

df_model_clean_timestep <- df_model_clean |>
  filter(pub_mo_label=="Feb")

df_model_clean_timestep |>
  select(adm1_name,era5_land_snow_cover,asi_Jun) |>
  ggplot(aes(
    x= era5_land_snow_cover, y= asi_Jun
  ))+
  geom_point()+
  facet_wrap(~adm1_name)

df_model_clean_timestep |>
  mutate(
    asi_lgl = asi_Jun>=3
  ) |>
  ggplot(aes(
    x= asi_lgl, y= era5_land_snow_cover
  ))+
  geom_boxplot()+
  geom_jitter()+
  facet_wrap(~adm1_name)

df_model_clean_timestep |>
  select(pub_mo_date, adm1_name, asi_Jun, era5_land_snow_cover) |>
  group_by(adm1_name) |>
  mutate(
    r_asi_jun =rank(asi_Jun),
    r_snow = rank(era5_land_snow_cover)
  ) |>
  summarise(
    cor(r_asi_jun, r_snow, use="complete.obs")
  )
  arrange(adm1_name)
  ggplot(aes(
    x= asi_lgl, y= era5_land_snow_cover
  ))+
  geom_boxplot()+
  geom_jitter()+
  facet_wrap(~adm1_name)

df_model_clean_timestep |>
  group_by(adm1_name)  |>
  summarise(
    cor(era5_land_snow_cover, asi_Jun, use="complete.obs")
  )


prob_vars <- which(df_model_clean_timestep |>
  summarise(
    across(everything(), ~sum(is.na(.)))
  ) |>
  t()!=0)

df_model_clean2 <- df_model_clean_timestep[,-prob_vars] |>
  select(-pub_mo_date,
         -pub_mo_label,
         # - adm1_name,
         -pub_yr_date)



df_model_clean2$asi_Jun
if(mode == "classification"){
df_model_clean2 <- df_model_clean2 |>
  mutate(
    asi_Jun = ifelse(asi_Jun>=3,"high","low")
  )
df_folds <- mc_cv(df_model_clean2, prop = 7/10, times = 20,strata = asi_Jun)

}
df_model_clean2$asi_Jun

# Prepare the data

# Define the recipe (skip unnecessary steps)
ranger_recipe <-
  recipe(formula = asi_Jun ~ ., data = df_model_clean2)

# Define the model specification (fixed hyperparameters)
ranger_spec <-
  rand_forest(mtry = 3, min_n = 5, trees = 1000) %>%  # Adjust these as needed
  set_mode(
    mode
    ) %>%
  set_engine("ranger", importance = "permutation")  # Enable feature importance

ranger_spec |>
  fit_resamples(asi_Jun~.,resamples = df_folds)
# Define the workflow
ranger_workflow <-
  workflow() %>%
  add_recipe(ranger_recipe) %>%
  add_model(ranger_spec) #|>
  fit_resamples(resamples = df_folds)

# Fit the model on the full dataset
final_fit <- fit(ranger_workflow, data = df_model_clean2)

# Extract feature importance
final_fit %>%
  pull_workflow_fit() %>%
  vip(aesthetics = list(alpha = 0.8)) +
  labs(title = "Feature Importance")

# Evaluate model performance (optional, using predictions on the same data)
df_model_clean2 %>%
  mutate(
    .pred = predict(final_fit, new_data = df_model_clean2) %>% pull(.pred_class)
  ) %>%
  ggplot(aes(x = asi_Jun, y = .pred)) +
  geom_point() +
  geom_abline() +
  labs(title = "Model Predictions vs Actuals", x = "Actual asi_Jun", y = "Predicted asi_Jun")

bind_cols(
  df_model_clean2 |>
    select(yr_date,asi_Jun),predict(final_fit,new_data = df_model_clean2)
  ) |>
  print(n=40)
```

```{r}

df_model_clean <- df_wide_scaled |>
  ungroup() |>
   filter(!is.na(asi_Jun),
          adm1_name!="Badghis",
          # adm1_name == "Takhar",
          pub_mo_date>= "1984-02-01"
          # pub_mo_label != "Jan"
          ) |>
  ungroup()

df_model_clean_timestep <- df_model_clean |>
  filter(pub_mo_label=="Feb")

prob_vars <- which(df_model_clean_timestep |>
  summarise(
    across(everything(), ~sum(is.na(.)))
  ) |>
  t()!=0)

df_model_clean2 <- df_model_clean_timestep[,-prob_vars] |>
  # mutate(
  #   # strata = paste0(adm1_name,"_",pub_mo_label)
  #   strata = adm1_name
  # ) |>
  select(-pub_mo_date, -pub_mo_label,
         # - adm1_name,
         -pub_yr_date)



df_model_clean2$asi_Jun
if(mode == "classification"){
df_model_clean2 <- df_model_clean2 |>
  mutate(
    asi_Jun = as_factor(ifelse(asi_Jun>=3,"high","low"))
  )
# df_model_clean2 <- df_model_clean2 |>
#   mutate(
#     asi_Jun = ifelse(asi_Jun>=3,"high","low")
#   )
df_folds <- mc_cv(df_model_clean2, prop = 7/10, times = 20,strata = asi_Jun)

}

df_recipe <-
  recipe(
    formula = asi_Jun ~ .,
    # data = NULL# df_model_clean2
    data =  df_model_clean2
    ) #|>
  step_rm(adm1_name)



#building model
tree <- decision_tree() %>%
   set_engine("rpart") %>%
   set_mode("classification")

tree_wf <- workflow() %>%
   add_recipe(df_recipe) %>%
   add_model(tree) %>%
    fit(df_model_clean2) #



 tree_fit <- tree_wf %>%
  pull_workflow_fit()


rpart.plot(tree_fit$fit, roundint= FALSE)
#workflow

models_by_group <-
  df_model_clean2 %>%
  group_by(adm1_name) %>%
  nest() |>
  mutate(
    fits = map(data,\(dft){
      fit(tree_wf,data = dft)
      }
      )
  )
      # model_fit = map(data, ~ tree_wf %>% fit(data = .))
    # model_fit = map(data, ~ tree_wf %>% fit(.))






```

```{r}
# reate LOOCV folds
df_folds <- loo_cv(df_model_clean2)

# Define the recipe (unchanged)
ranger_recipe <-
  recipe(formula = asi_Jun ~ ., data = df_model_clean2) %>%
  step_string2factor(one_of(NA_character_))

# Define the model specification (with hyperparameters to tune)
ranger_spec <-
  rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>%
  set_mode("regression") %>%
  set_engine("ranger")

# Define the workflow
ranger_workflow <-
  workflow() %>%
  add_recipe(ranger_recipe) %>%
  add_model(ranger_spec)

# Tune with LOOCV
ranger_tune <-
  tune_grid(
    ranger_workflow,
    resamples = df_folds, # Use LOOCV folds
    grid = 11
  )
loocv_results <- map_df(1:nrow(df_model_clean2), function(i) {
  # Split the data: Leave one observation out
  test_data <- df_model_clean2[i, ]
  train_data <- df_model_clean2[-i, ]

  # Tune hyperparameters on the training data
  tune_res <- tune_grid(
    ranger_workflow,
    resamples = vfold_cv(train_data, v = 5),  # Use 5-fold CV for tuning within training set
    grid = 11
  )

  # Select the best parameters
  best_params <- select_best(tune_res, metric = "rmse")

  # Finalize the model
  final_workflow <- finalize_workflow(ranger_workflow, best_params)

  # Fit the finalized model to the training data
  final_model <- fit(final_workflow, data = train_data)

  # Predict on the test data
  preds <- predict(final_model, new_data = test_data) %>%
    bind_cols(test_data)

  return(preds)
})

loocv_results |>
  ggplot(aes(x = asi_Jun, y = .pred)) +
  geom_point() +
  geom_abline() +
  labs(title = "LOOCV Predictions vs. Actuals", x = "Actual", y = "Predicted")
# Review results
show_best(ranger_tune, metric = "rmse")
show_best(ranger_tune, metric = "rsq")
autoplot(ranger_tune)

# Finalize the model
final_rf <- ranger_workflow |>
  finalize_workflow(select_best(ranger_tune))

# Fit the final model on the full dataset
final_fit <- fit(final_rf, data = df_model_clean2)

# Evaluate predictions
predictions <- predict(final_fit, new_data = df_model_clean2) |>
  bind_cols(df_model_clean2)

# Visualize results
predictions |>
  ggplot(aes(x = asi_Jun, y = .pred)) +
  geom_point() +
  geom_abline()
```

```{r}
#' experimental
#'
df_env_compare$parameter |> unique()

df_environmental |>
  group_by(parameter) |>
  summarise(
    min = min(yr_date),
    max = max(yr_date)
  ) |>
  print(n=73)


df_env_scaled <- df_env_compare |>
  ungroup() |>
  mutate(
    mo_mon = extract_month(parameter),
    mo_pub = extract_month_pub(parameter)
  ) |>
  select(
    yr_date, yr_season, mo_mon,,mo_pub,everything()
  ) |>
  group_by(
    mo_pub,adm1_name,parameter
  ) |>
  mutate(
    value_scaled = scale(value, center= T)[,1],
    .after=value
  ) |>
  ungroup()

ldf_env_wide <- split(df_env_scaled,df_env_scaled$mo_pub) |>
  map(
    \(dft){
      dft |>
        pivot_wider(
          id_cols = c("adm1_name","yr_season","mo_pub"),
          names_from = "parameter",
          values_from = "value_scaled"
        ) |>
        select(yr_season,adm1_name,any_of(unique(df_env_scaled$parameter)))
    }
  )

lpca_modis <- split(ldf_env_wide$Mar,ldf_env_wide$Mar$adm1_name) |>
  map(
    \(dft){
      dft_num <- dft |>
        filter(
          year(yr_season)<=2021,
          year(yr_season)>=2000
          ) |>
        select(where(is.numeric),-matches("era5_[tc]"))

      # colSums(is.na(dft_num))

      princomp(dft_num)
    }
  )

install.packages("FactoMineR")
box::use(
  FactoMineR[...]
)
plot(lpca_modis$Badakhshan)
biplot(lpca_modis$Badakhshan)
biplot(lpca_modis$Faryab)
fv
summary(lpca_modis$Badakhshan)
colSums(is.na(ldf_env_wide$Mar |> filter(year(yr_season)>1984,year(yr_season)<=2021)))

ldf_env_wide$Mar |> select(where(is.numeric),-starts_with("era5"))

df_env_scaled |>
  filter(mo_pub=="Jun") |>
  pivot_wider(
        id_cols = c("adm1_name","yr_season","mo_pub"),
          names_from = "parameter",
          values_from = "value_scaled"

  )
  glimpse()

ldf_env_wide$Apr |>
  filter(is.na(era5_cumu_precip_Mar))
```

```{r}
#| eval: false

lp_tile <- aoi_adm1 |>
  map(
    \(adm_tmp){
      df_env_filt |>
        filter(adm1_name == adm_tmp) |>
      arrange(
        parameter,adm1_name,yr_season
      ) |>
        mutate (
          yr = year(yr_season),
          rp_bin = cut(rp_relevant_direction,
                       breaks = c(1, 2, 3, 4, 5, Inf),
                       labels = c("1-2", "2-3", "3-4", "4-5", ">=5"),
                       include.lowest = TRUE)
        ) |>
        ggplot(
          aes(x = yr_season, y = parameter, fill = rp_bin)
        )+
        geom_tile_interactive(aes(
          tooltip = glue(
            "year: {yr}
      parameter: {parameter}
      value: {value},
      RP: {rp_relevant_direction}
      "
          )
        ))+
        labs(title = adm_tmp)+
        scale_fill_manual(
          values = c(
            "1-2" = "#d4efdf",   # Light green (low severity)
            "2-3" = "#f7dc6f",   # Yellow (moderate severity)
            "3-4" = "#f5b041",   # Orange (high severity)
            "4-5" = "#e74c3c",   # Red (severe drought)
            ">=5" = "#943126"    # Dark red (extreme drought)
          )
        )+
        # facet_wrap(~!!sym(adm_tmp) ,ncol = 1)+
        theme(
          axis.text.y= element_text(size=5)
        )
    }
  )


girafe(ggobj = lp_tile[[2]])
```

## Suggested Next Steps:

## Appendix


one off snow checks
```{r}
#| eval: false


# quick gut check on era5 snow fall - can add to appendix during cleanup
df_era_snow <- df_environmental_yr_adj |>
  filter(
    yr_season >= df_validation_range[1],
    yr_season <= df_validation_range[2],
    adm1_name %in% aoi_adm1
  ) |>
  # this group_by() -> filter() step is just to grab the last dekad per month
  # for ASI & VHI, the rest of the data is monthly time-step so it has no
  # effect on them
  group_by(
    adm1_name, parameter, pub_mo_date
  ) |>
  filter(
    date == max(date)
  ) |>
  ungroup() |>
  filter(
    # FEB ERA SNOW
    month(pub_mo_date)==2
    # regex to detect snow_cover or snow melt
    &str_detect(parameter, "era.*snow(_cover|melt_sum)")|
      # ASI
      (str_detect(parameter,"asi")& month(pub_mo_date)==6)
  )

df_era_snow_wide <- df_era_snow |>
  pivot_wider(
    id_cols =c(yr_date, adm1_name),
    names_from = parameter,
    values_from = value
  )


df_era_snow_wide |>
  group_by(
    adm1_name
  ) |>
  summarise(
    cor(asi,era5_land_snow_cover)
  )

# snow melt correlations are negative... so feb pub (jan data) does negatively correlate to ASI
# more Jan snow melt less agg stress... could look more into ERA5 snowmelt var, but strange
df_era_snow_wide |>
  group_by(
    adm1_name
  ) |>
  summarise(
    cor(asi,era5_land_snowmelt_sum)
  )
df_env_filt_flag |> count(parameter) |> print(n=23)

df_era5_land_rps_wide <- df_env_filt_flag |>
  filter(
    # (month(pub_mo_date)==2 & str_detect(parameter, "era.*snow(_cover|melt_sum)"))
    (month(pub_mo_date)==2 & str_detect(parameter, "^era5_land"))|
      # ASI
      (str_detect(parameter,"asi")& month(pub_mo_date)==6)
  ) |>
   pivot_wider(
    id_cols =c(yr_date, adm1_name),
    names_from = parameter,
    values_from = rp_relevant_direction
  )

df_era5_land_rps_wide |>
ggplot(
    aes(x= era5_land_snow_cover, y= asi)
  )+
  geom_point()



thresh <- 5
df_era5_land_rps_wide |>
  filter(adm1_name %in% c("Takhar","Faryab","Sar-e-Pul")) |>
  mutate(
    perf_class = case_when(
      era5_land_surface_runoff_sum>=thresh & asi>=thresh~"TP",
      era5_land_surface_runoff_sum>=thresh & asi<thresh~"FP",
      era5_land_surface_runoff_sum<=thresh & asi>thresh~"FN",
      era5_land_surface_runoff_sum<=thresh & asi<=thresh~"TN"
    )
  ) |>
  glimpse() |>
  ggplot(
    aes(x= era5_land_surface_runoff_sum, y= asi, color = perf_class)
  )+
  geom_point()+
  scale_color_manual(
    values= c("TP"= hdx_hex("mint-hdx"),
              "TN"= hdx_hex("mint-light"),
              "FN"= hdx_hex("tomato-hdx"),
              "FP"=hdx_hex("tomato-light") )
  )+
  geom_vline(xintercept=thresh)+
  geom_hline(yintercept = thresh)+
  scale_x_log10()+
  scale_y_log10()+
  facet_wrap(~adm1_name)+
  theme(
    panel.border = element_rect(color="grey",fill=NA)
  )+
  labs(
    title= "Performance: Feb ERA5 Surface Runoff Sum Predicting End of Seasson ASI",
    caption = "High return periods runoff corresponds to deficit of runoff where as high RP ASI corresponds to higher vegetative stress (end of May)"
    )


# df_era_snow_esa |>
#   ggplot(aes(x= )
```


```{r}
#| eval: false

# dont really like where this plot was going -- probably delete
df_metrics_simple |>
  filter(
    pub_mo_label %in% c("Feb","Mar","Apr","May"),
  ) |>
  group_by(adm1_name, pub_mo_label) |>
  slice_max(f1, n= 3, with_ties = T) |>
  ggplot(
    aes(
      x= parameter_label,
      y= adm1_name,
      fill = f1
    )
  )+
  # scale_x_discrete(limits = c("Feb","Ma
  geom_tile() +
  facet_grid(
    cols = vars(pub_mo_label),scales="free"
    )+
  theme(
    axis.text.x = element_text(angle =90),
    panel.border = element_rect(fill = NA, color = "grey")
  )

```
### Point System

Point system - you need 4 points

Feb: SEAS5 : 4 points Mar: - Observational Rainfall (CHIRPS): 2 points - Observed Soil Moisture : 2 points - NDSI or SWE: 2 points April: - Observed Rainfall CHIRPS OR ERA5: 4 points - NDSI: 2 points - SWE: 2 points May: - ASI: 4 points - VHI: 2 points - Soil Moisture: 2 points

Jun: - ASI: 4 - VHI: 2 - Soil Moisture 2

```{r}

df_ps <- df_env_compare |>
  label_parameters() |>
  mutate(
    ps = case_when(
      str_detect(parameter,"NDSI|SWE") & pub_mo_label == "Mar"~"SNOW_COMBINED",
      str_detect(parameter,"era5_land_volumetric_soil")~"soil_moisture",
      str_detect(parameter,"NDSI") & pub_mo_label != "Mar"~"NDSI",
      # str_detect(ps,"chirps_cumu|era5_cumu_precip") ~"observational_chirps_era5",
      str_detect(parameter,"cumu_chirps|cumu_era5_land_total_precipitation") ~"cumu_observational_chirps_era5",
      .default = parameter
    )
  )




df_rps_composite <- df_ps |>
  group_by(
    pub_mo_label, yr_season,adm1_name,ps
  ) |>
  slice_max(
    rp_relevant_direction,n=1,with_ties = F
  ) |>

  summarise(
    rp = max(rp_relevant_direction,na.rm=T),
    # rp_flag = rp>=5,
  ) |>
  mutate(
      rp_flag = case_when(
      ps == "SEAS5-MAM" ~ rp >=5,
      ps == "soil_moisture" ~ rp>=10,
      ps == "SNOW_COMBINED" ~ rp>=10,
      ps == "NDSI" ~ rp>=10,
      ps == "cumu_observational_chirps_era5"~ rp>=3,
      ps == "asi"~ rp>=5,
      ps == "vhi"~ rp>=5,
      .default = NA_real_
    )
  ) |>
  ungroup()


df_pts_added  |>
  filter(pub_mo_label == "Mar") |>
  print(n=30)

df_pts_added <- df_rps_composite |>
  filter(pub_mo_label!="Jan") |>
  mutate(
    pts1 = case_when(

      # Feb - Just SEAS5
      pub_mo_label == "Feb" & ps=="SEAS5-MAM" & rp_flag~4,
      pub_mo_label == "Feb" & ps=="SEAS5-MAM" & !rp_flag~0,

      # March - soil moisture, snow, rainfall
      pub_mo_label == "Mar" & ps=="soil_moisture" & rp_flag~2,
      pub_mo_label == "Mar" & ps=="soil_moisture" & !rp_flag~0,
      pub_mo_label == "Mar" &  ps=="SNOW_COMBINED" & rp_flag~2,
      pub_mo_label == "Mar" &  ps=="SNOW_COMBINED" & !rp_flag~0,
      pub_mo_label == "Mar" &  ps=="cumu_observational_chirps_era5" & rp_flag~2,
      pub_mo_label == "Mar" &  ps=="cumu_observational_chirps_era5" & !rp_flag~0,

      # april Snow/ rainfall
      pub_mo_label == "Apr" & ps == "NDSI" & rp_flag~2,
      pub_mo_label == "Apr" & ps == "NDSI" & !rp_flag~0,
      pub_mo_label == "Apr" & ps == "SWE_inst" & rp_flag~2,
      pub_mo_label == "Apr" & ps == "SWE_inst" & !rp_flag~0,
      pub_mo_label == "Apr" & ps=="cumu_observational_chirps_era5" & rp_flag~4,
      pub_mo_label == "Apr" & ps=="cumu_observational_chirps_era5" & !rp_flag~0,

      # May
      pub_mo_label == "May" & ps == "asi" & rp_flag~4,
      pub_mo_label == "May" & ps == "asi" & !rp_flag~0,
      pub_mo_label == "May" & ps == "vhi" & rp_flag~2,
      pub_mo_label == "May" & ps == "vhi" & !rp_flag~0,
      pub_mo_label=="May" & ps == "soil_moisture" & rp_flag ~ 2,
      pub_mo_label=="May" & ps == "soil_moisture" & !rp_flag ~ 0,


      # june = ASI, VHI, soil moisture
      pub_mo_label == "Jun" & ps == "asi" & rp_flag~4,
      pub_mo_label == "Jun" & ps == "asi" & !rp_flag~0,
      pub_mo_label == "Jun" & ps == "vhi" & rp_flag~2,
      pub_mo_label == "Jun" & ps == "vhi" & !rp_flag~0,
      pub_mo_label=="Jun" & ps == "soil_moisture" & rp_flag ~ 2,
      pub_mo_label=="Jun" & ps == "soil_moisture" & !rp_flag ~ 0,
      # ps %in% ps_considered ~0,
      .default = NA)
    )






df_pts_added  |>
  filter(pub_mo_label %in% c("Feb","Mar","Apr","May","Jun")) |>
  group_by(pub_mo_label,adm1_name,yr_season ) |>
  summarise(
    pts1 = sum(pts1,na.rm=T),
    pts_flag = pts1>=4,.groups="drop_last"
  ) |>
  # count(pts1)
  summarise(
    ar = mean(pts_flag),
    rp = 1/ar
  ) |>
  print(
    n =25
  )

df_pts_added  |>
  filter(adm1_name!="Badghis") |>
  filter(pub_mo_label %in% c("Feb",
                       "Mar",
                       "Apr",
                       # "May",
                       "Jun"
                       )) |>
  group_by(pub_mo_label,yr_season,adm1_name ) |>
  summarise(
    pts1 = sum(pts1,na.rm=T),
    pts_flag = pts1>=4,.groups="drop_last"
  ) |>
  group_by(
    pub_mo_label,yr_season
  ) |>
  summarise(
    pts_flag = any(pts_flag)
  ) |>
  group_by(pub_mo_label) |>
  # count(pts1)
  summarise(
    ar = mean(pts_flag),
    rp = 1/ar
  ) |>
  print(
    n =25
  )

df_pts_added  |>
  filter(!adm1_name %in% c("Badghis","Badakhshan")) |>
  filter(pub_mo_label %in%
           c(
             # "Feb",
             # "Mar",
             # "Apr",
             "May",
             "Jun"
           )) |>
  group_by(pub_mo_label,yr_season,adm1_name ) |>
  summarise(
    pts1 = sum(pts1,na.rm=T),
    pts_flag = pts1>=4,.groups="drop_last"
  ) |>
  group_by(
    pub_mo_label,yr_season
  ) |>
  summarise(
    pts_flag = any(pts_flag)
  ) |>
  group_by(yr_season) |>
  summarise(
    pts_flag = any(pts_flag)
  ) |>
  # count(pts1)
  summarise(
    ar = mean(pts_flag),
    rp = 1/ar
  ) |>
  print(
    n =25
  )


df_pts_monitoring_historical <- df_pts_added  |>
  filter(
    pub_mo_label %in% c("Feb","Mar","Apr","May","Jun"), !is.na(pts1)
  )



df_pts_w_activation <- df_pts_monitoring_historical |>
   group_by(
    pub_mo_label, yr_season,adm1_name
  ) |>
  mutate(
    activate = sum(pts1)>=4,
    activate_color = ifelse(activate&pts1>0,"yes","no")
  )

mos_monitoring <- list(
  "Feb"= "Feb",
  "Mar"= "Mar",
  "Apr" = "Apr",
  # "May" = "May",
  "Jun" = "Jun"
)


lp_observational_tile <- map(
  mos_monitoring,
  \(mo_tmp){
  df_pts_w_activation |>
      filter(adm1_name!="Badghis",
             pub_mo_label==mo_tmp) |>
  ggplot(
    aes(x= yr_season,
        y= ps ,
        fill = pts1,
        color = activate_color)
  )+
  geom_tile( linewidth = 1)+
  scale_color_manual(
    values = c(
      "yes" = "red",
      "no"= NULL
    )
  )+
  scale_fill_gradient()+
  facet_grid(
    rows = vars(adm1_name),
    scales= "free")
  }
)

lp_observational_tile$Apr
lp_observational_tile$Mar

```

### Runoff

```{r}
# i need to have a little look at this runoff var

df_env_filt |>
  filter(
    str_detect(parameter, "runoff")
  ) |> count(parameter)

df_runoff <- df_env_filt |>
  filter(
    str_detect(parameter, "runoff_max")
    )

df_asi <- df_env_filt |>
  filter(str_detect(parameter, "asi"), pub_mo_label =="Jun") |>
  rename(
    asi = value
  ) |>
  select(
    adm1_name, pub_yr_date, asi,rp_asi = rp_relevant_direction
  )

df_runoff_asi <- df_runoff |>
  filter(
    # pub_mo_label == "Feb",
    adm1_name == "Takhar"
  ) |>
  left_join(
    df_asi, by = c("pub_yr_date","adm1_name")
  ) |>
  group_by(adm1_name, pub_mo_label) |>
  mutate(
    anom_runoff = value - mean(value, na.rm = T),
    anom_asi = asi - mean(asi, na.rm = T)
  ) |>
  select(
    adm1_name, date, pub_mo_date,pub_mo_label, anom_runoff,anom_asi, value, asi,
    rp_runoff = rp_relevant_direction,rp_asi
  )

df_runoff_asi |>
  ggplot(
    aes(x= anom_runoff, y= anom_asi, color = rp_runoff)
  )+
  geom_point_interactive(
    aes(tooltip = glue("date: {date}
                       value:{value}"))
  )+
  scale_color_continuous(trans = scales::log10_trans())+
  facet_wrap(~pub_mo_label, scales= "free")

runoff_rp_threshold <- 3
asi_rp_threshold <-  3
p_runoff_rp_scatter <- df_runoff_asi |>
  mutate(
    rp_class = case_when(
      rp_runoff>=runoff_rp_threshold &
      rp_asi >=asi_rp_threshold ~"TP",
      rp_runoff<runoff_rp_threshold &
      rp_asi <asi_rp_threshold ~"TN",

      rp_runoff>=runoff_rp_threshold &
      rp_asi <asi_rp_threshold ~"FP",

      rp_runoff<runoff_rp_threshold &
      rp_asi >=asi_rp_threshold ~"FN"

    )
  ) |>

  ggplot(
    aes(x= rp_runoff, y= rp_asi, color = rp_class)
  )+
  geom_vline(
    xintercept = runoff_rp_threshold
  )+
  geom_hline(
    yintercept = asi_rp_threshold
  )+
  geom_point_interactive(
    aes(tooltip = glue("date: {date}
                       runoff:{value}
                       asi: {asi}
                       RP ASI: {rp_asi}
                       RP runoff: {rp_runoff}"))
  )+
  # scale_color_continuous(trans = scales::log10_trans())+
  facet_wrap(~pub_mo_label, scales= "free")

girafe(ggobj =p_runoff_rp_scatter)


```
