---
title: Composite Indicator

format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  include: true
  echo: true
  warning: false
  message: false
  eval: true
  results: "asis"
  out.width: "100%"
  code-fold: true
editor: visual
project:
  execute-dir: project
---

## Intro

```{r}

aoi_adm1 <- c(
  "Takhar",
  # "Badakhshan",
  # "Badghis",
  "Sar-e-Pul" ,
  "Faryab"
  )

label_parameters <- function(df){
  df |>
    mutate(
     parameter_label = case_when(
      str_detect(parameter, "era5_land_volumetric_soil")~ "Soil Moisture (ERA5)",
      str_detect(parameter,"NDSI")~"NDSI",
      str_detect(parameter,"asi")~"ASI",
      str_detect(parameter,"vhi")~"VHI",

      str_detect(parameter,"cumu_chirps_precipitation_sum")~"Precip cumu (CHIRPS) ",
      str_detect(parameter,"chirps_precipitation_sum")~"Precip (CHIRPS)",
      str_detect(parameter,"cumu_era5_land_total_precipitation_sum")~"Precip cumu (ERA)",
      str_detect(parameter,"era5_land_total_precipitation_sum")~"Precip (ERA)",
      str_detect(parameter,"mean_2m_air_temperature")~"Temp (ERA)",
      str_detect(parameter,"era5_land_snow_depth_water_equivalent")~"SDWE (ERA5)",
      str_detect(parameter,"era5_land_snow_cover")~"Snow Cover (ERA5)",
      str_detect(parameter,"era5_land_snowmelt_sum")~"Snow Melt (ERA5)",

      str_detect(parameter,"runoff_max")~"Runoff max (ERA5)",
      str_detect(parameter,"runoff_sum")~"Runoff sum (ERA5)",
      str_detect(parameter,"SWE_inst")~"SWE (FLDAS)",


      .default = parameter
    )
    )
}



```

```{r}
SEASON_OF_INTEREST <- c(3,4,5)
# LATEST_SEASON_PREDICTABLE <-  c(3,4)

box::use(
  ../R/blob_connect,
  ../R/utils,
  seas5 = ../R/seas5_utils,
  loaders = ../R/load_funcs,
  dplyr[...],
  tidyr[...],
  stringr[...],
  glue[...],
  janitor[...],
  yardstick[...],
  ggplot2[...],
  gghdx[...],
  ggrepel[...],
  lubridate[...],
  sf[...],
  readr[...],
  purrr[...],
  patchwork[...],
  cumulus
)

gghdx()


```

## Heuristic Weighting

Based on the above let's select some indicators and make a weighting

```{r}
df_compiled_indicators <- cumulus$blob_read(
  container = "projects",
  name = "ds-aa-afg-drought/processed/vector/df_all_combined_indicators_v2.parquet"
)

df_env_model <-
  df_compiled_indicators |>
  # df_env_compare |>
    select(
    date,
    yr_season,
    pub_mo_date,
    pub_mo_label,
    adm1_name,
    parameter,
    value
    ) |>
  filter(
    # these parameters are just a bit of noise
    !parameter %in% c("NDSI_Snow_Cover_min","NDSI_Snow_Cover_max","era5_land_runoff_max"),
    # these were consolidated into the average of all 4 soil moisture layers
    !str_detect(parameter,"era5_land_volumetric_soil_water_layer_\\d"),
    # already have these from ERA land up to 2024 rather than 2020
    !(parameter %in% c("total_precipitation","mean_2m_air_temperature"))

  )

# just some manual inspectino stuff here
df_env_model |>
  filter(str_detect(parameter,"^asi|^vhi")) |>
  group_by(parameter, pub_mo_label) |>
  count() |>
  pivot_wider(names_from=pub_mo_label, values_from = n)

date_ranges <- df_env_model |>
  group_by(parameter) |>
  summarise(
    start = min(pub_mo_date),
    end = max(pub_mo_date)
  )
```

### Weighting Functons

```{r weightingFuncs}



#' create_weight_grid
#' @description
#' function to create valid weight sets in long data.frame that can be
#' iterated through
#' @param x
#' @param wt_vals
#'
#' @returns
#' @export
#'
#' @examples
#' params_set_gte_1984 <- c(
#'   "era5_land_soil_moisture_1m",
#'   "cumu_era5_land_total_precipitation_sum",
#'   "vhi",
#'   "era5_land_snow_cover", # toying w/ adding this or not
#'   "asi"
#'   )
#' weight_combos <- create_weight_grid(params_set_gte_1984,wt_vals = c(0,seq(0.1, 1, by = 0.1)))

create_weight_grid <- function(x,wt_vals){

  df <- map(x,\(xt){
    tibble(
      !!sym(xt) :=wt_vals
    )
  }
  )|>
    list_cbind()

  df_expanded <- expand_grid(!!!df)

  df_valid_combinations <- df_expanded[rowSums(df_expanded) == 1, ]

    df_valid_combinations |>
      mutate(
        wt_id = row_number()
      ) |>
      pivot_longer(
        -wt_id, names_to = "parameter", values_to ="weight"
      )
}



#' Title
#'
#' @param df
#' @param params_included
#' @param earliest_year
#'
#' @returns
#' @export
#'
#' @examples
#' df_env_model |>
#'   normalize_to_z(params_included = l_params$gte1984,earliest_year = 1984)

normalize_to_z <- function(df,params_included, earliest_year){
      df |>
        filter(
          year(yr_season)>=earliest_year
        ) |>
        mutate(
        pub_mo_label = as.character(pub_mo_label)
      ) |>
      filter(
        parameter %in% params_included,
        pub_mo_label %in% c(month.abb[4:6])
      ) |>
      group_by(
        pub_mo_label, adm1_name, parameter
      ) |>
      mutate(
        zscore = scale(value,center=T,scale=T)[,1],
        zscore = ifelse(parameter != "asi",zscore*-1,zscore)
      )  |>
      ungroup()
}

#' Title
#'
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
#' df_env_model |>
#'   normalize_to_z(params_included = l_params$gte2000,earliest_year = 2001) |>
#'   extract_truth_set()

extract_truth_set <- function(df){
  df |>
    filter(
      month(pub_mo_date)==6,
      parameter == "asi"
    ) |>
    select(
      yr_season, adm1_name, value,zscore_asi_Jun=zscore
    ) |>
    distinct()
}

#' Title
#'
#' @param df
#' @param params_included
#'
#' @returns
#' @export
#'
#' @examples
summarise_z <- function(
    df,
    params_included,
    weight_values
    ){

  df_weight_grid <- create_weight_grid(x = params_included, wt_vals = weight_values)

  df_filt <- df |>
    filter(
      parameter %in% params_included
    )
  split(df_filt,df_filt$pub_mo_label) |>
    map(\(dft){

      split(
        df_weight_grid,
        df_weight_grid$wt_id
      ) |>
        map(
          \(dft_w){

            dft_weighted <- dft |>
              left_join(
                dft_w, by = "parameter"
              )
            dft_summarised <- dft_weighted |>
              group_by(
                yr_season ,
                pub_mo_label,
                adm1_name,
                pub_mo_date,wt_id
              ) |>
              summarise(
                zscore = weighted.mean(zscore,w=weight,na.rm=T)
                ,.groups="drop"
              ) |>
              mutate(
                wt_set = list(dft_w)
              )
          }
        ) |>
        list_rbind()
    }
    ) |>
    list_rbind()

}



#' Title
#'
#' @param df
#' @param params_included
#' @param earliest_year
#' @param rp
#'
#' @returns
#' @export
#'
#' @examples
weighted_classify <- function(df,
                              params_included,
                              earliest_year,
                              rp=3,
                              weight_values ){
  df_normalized <- df |>
    normalize_to_z(
      params_included = params_included,
      earliest_year = earliest_year
    )

  df_truth <- df_normalized |>
    extract_truth_set()

  df_z_weighted <- df_normalized |>
    summarise_z(params_included = params_included,weight_values= weight_values)

   df_z_weighted |>
      left_join(
        df_truth |>
          select(-value)
      ) |>
      utils$threshold_var(
        var= "zscore",
        by = c("pub_mo_label","adm1_name","wt_id"),
        rp_threshold = rp
      ) |>
      utils$threshold_var(
        var= "zscore_asi_Jun",
        by = c("pub_mo_label","adm1_name","wt_id"),
        rp_threshold = rp
      ) |>
      arrange(adm1_name, yr_season,wt_id) |>
      ungroup()

}

single_indicator_performance <- function(df, parameter,earliest_year,rp){
  df_normalized <- df |>
    normalize_to_z(
      params_included = parameter,
      earliest_year = earliest_year
    )

  df_truth <- df_normalized |>
    extract_truth_set()



  df_classified <- df_normalized |>
    left_join(
      df_truth |>
        select(-value)
    ) |>
    utils$threshold_var(
      var= "zscore",
      by = c("pub_mo_label","adm1_name"),
      rp_threshold = rp
    ) |>
    utils$threshold_var(
      var= "zscore_asi_Jun",
      by = c("pub_mo_label","adm1_name"),
      rp_threshold = rp
    ) |>
    arrange(adm1_name, yr_season) |>
    ungroup()
  summarise_performance(df_classified,by = c("pub_mo_label","adm1_name"))
}
#' Title
#'
#' @param df
#' @param by
#'
#' @returns
#' @export
#'
#' @examples
summarise_performance <-  function(
    df,
    by=c("pub_mo_label","adm1_name","wt_id","wt_set"),
    parameter_subset = NULL
    ){
  if(!is.null(parameter_subset)){
    df |>
      filter(
        parameter %in% parameter_subset
      )
  }
  df |>
    mutate(
      across(ends_with("_flag"),\(x) factor(x,levels = c("TRUE","FALSE")))
    ) |>
    group_by(
      across({{by}})
    ) |>
    f_meas(zscore_asi_Jun_flag, zscore_flag, estimator= "binary",event_level = "first") |>
    ungroup()

}

#' Title
#'
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
top_performance_per_moment <- function(df,n=1){
  # df=  ldf_perf_all_models$gte1984,n=1
  df_max <- df |>
    group_by(pub_mo_label, wt_id) |>
    summarise(
      avg_estimate = mean(.estimate)
    ) |>
    slice_max(
      order_by = avg_estimate,
      n= n
    )|>
    ungroup()

  inner_join(
    df_max,
    select(df,
           any_of(c("adm1_name","pub_mo_label", "wt_id", "wt_set",".estimate","asi_f1"))
           ),
    by = c("pub_mo_label","wt_id")
  )
}

#' Title
#'
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
plot_optimal_compositions <-  function(df, label_plot = F, pal){

  df_labelled <- df|>
    select(
      pub_mo_label,adm1_name,wt_set,avg_estimate,.estimate
    ) |>
    unnest(wt_set) |>
    utils$label_parameters() |>
    mutate(
      p_label = glue(
      "id: {wt_id}
      f1: {scales::label_number(accuracy =0.001)(.estimate)}
      avg f1: {scales::label_number(accuracy =0.001)(avg_estimate)}"
      ),
      weight_pct_label = scales::percent(weight,accuracy = 1, trim = FALSE),
    )

  p <- df_labelled |>
    group_by(pub_mo_label, adm1_name) |>
    mutate(
      pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
      id = dense_rank(wt_id)
    ) |>
    ungroup() |>
    ggplot(
      aes(x= id, y= weight,fill = parameter_label)
    )+
    geom_bar(
      stat= "identity", color = "black"
    )+

    scale_fill_manual(values=pal)+
    facet_grid(cols= vars(pub_mo_facet),
               rows = vars(adm1_name)
               ,scales="free")+
    labs(
      x = "Different indicator weightings"

    )+
    scale_y_continuous(labels=scales::label_percent())+
    theme(
      legend.title = element_blank(),
      axis.text.x = element_blank()
    )
  if(label_plot){
    # df <- ldf_tops2$gte1984
    df |>
      mutate(
        p_label = glue("{wt_id}: {scales::label_number(accuracy =0.01)(avg_estimate)}")
      )
    p <- p +
      # composition plot
      geom_text(
        aes(x= id, y= weight, label = weight_pct_label),
        position = position_stack(vjust = 0.5), color ="black"
      )+
      # wt plot
      geom_label(aes(x= id, y= 1.1, label = p_label), color ="black",fill="beige",alpha=0.4)+
      # theme(
      #   panel.spacing.y = unit(4, "lines")
      # )+
      expand_limits(y = 1.2)
  }
  p
}



#' Title
#' helper func to find weight ids that are simple and optimal
#' not a perfect solution, but gets us 90% there.
#' @param df
#'
#' @returns
#' @export
#'
#' @examples
plot_low_var_optimal <-  function(df,
                                  pal,
                                  label_wt_id =T){


  dfp <- df|>
    select(
      pub_mo_label, wt_set
    ) |>
    unnest(wt_set) |>
    group_by(
      pub_mo_label,wt_id
    ) |>
    mutate(
      sd = sd(weight)

    ) |>
    group_by(pub_mo_label) |>
    slice_min(sd) |>
    group_by(pub_mo_label) |>
    mutate(
      pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
      id = dense_rank(wt_id)
    ) |>
    ungroup() |>
    utils$label_parameters()

  p <- dfp |>
    distinct() |>
    mutate(
      weight_pct_label = scales::percent(weight,accuracy = 1, trim = FALSE),
      weight_id_label = glue("weight id: {wt_id}")

    ) |>
    filter(pub_mo_label!= "Jun") |>
    ggplot(
      aes(
        x= id,
        y= weight,
        fill = parameter_label
      ),
      position = position_stack(vjust = 0.5)
    )+
    geom_bar(
      stat= "identity", color = "black"
    )+

    geom_text(aes(x= id, y= weight, label = weight_pct_label),
              position = position_stack(vjust = 0.5), color ="black")+
    scale_fill_manual(values=pal) +
    scale_y_continuous(labels =scales::label_percent())+
    facet_grid(cols= vars(pub_mo_facet)
               ,scales="free")+
    labs(
      # title = "May publication weight combos that give avg 0.90 f1"
    )
  if(label_wt_id){
    p <- p +
      geom_text(
        aes(
          x= id+.25,
          y= 0.5,
          label = weight_id_label
          ), color ="black"
      )
  }
  p
}


#' Title
#'
#' @param df
#' @param param_simple
#'
#' @returns
#' @export
#'
#' @examples
get_simple_weight_id <- function(df,param_simple){
    df |>
    select(wt_set) |>
    unnest(wt_set) |>
    distinct() |>
    filter(
      parameter == param_simple ,
      weight == 1
    ) |>
    pull(wt_id) |>
    unique()

}
#' Title
#' @description
#' After plotting different optimal weight combinations you might want to compare specific
#' weighting compositions to a simpler model (i.e ASI). So this allows you to plug in the
#' weight ids for both april and may as well as `param_simple` which has always been ASI up to
#' this point in analysis. The output data.frame is interesting for plotting.
#'
#' @param df
#' @param wt_id_apr
#' @param wt_id_may
#' @param param_simple
#'
#' @returns
#' @export
#'
#' @examples
compare_to_simple_model <- function(df,wt_id_apr, wt_id_may,param_simple){
  simple_model_id <-get_simple_weight_id(df= df, param_simple = param_simple)

    df_chosen <- df |>
      summarise_performance() |>
      filter(
        (pub_mo_label == "Apr" & wt_id==wt_id_apr)|
        (pub_mo_label == "May" & wt_id==wt_id_may)|
        wt_id == simple_model_id

      ) |>
      filter(pub_mo_label != "Jun")

    df_simple <- df_chosen |>
      filter(
        wt_id == simple_model_id
      )
    anti_join(df_chosen,df_simple) |>
      left_join(
        df_simple |>
          select(
            pub_mo_label,adm1_name,
            estimate_simple =.estimate),
        by = c("pub_mo_label","adm1_name")
      )

}
```

```{r decideParameters}
# setup as many sets of parameters as we want
# subequent code will iterate through each parameter set

params_set_gte_2001 <- c(
  "era5_land_soil_moisture_1m",
  "cumu_era5_land_total_precipitation_sum",
  "NDSI_Snow_Cover_mean",
  "vhi",
  "asi"
)

# and one that takes data going back to 1984
params_set_gte_1984 <- c(
  "era5_land_soil_moisture_1m",
  "cumu_era5_land_total_precipitation_sum",
  "vhi",
  "era5_land_snow_cover", # toying w/ adding this or not
  "asi"
)
params_set_gte_1984_no_snow <- c(
  "era5_land_soil_moisture_1m",
  "cumu_era5_land_total_precipitation_sum",
  "vhi",
    # "era5_land_snow_cover", # toying w/ adding this or not
  "asi"
)

params_set_gte_1984_w_mixed_forecast_obs <- c(
  "era5_land_soil_moisture_1m",
  "cumu_era5_land_total_precipitation_sum",
  "vhi",
  "mam_mixed_seas_observed",
    "era5_land_snow_cover", # toying w/ adding this or not
  "asi"
)

l_params <- list(
  "gte2001"= params_set_gte_2001,
  "gte1984" = params_set_gte_1984,
  "gte_1984_no_snow" = params_set_gte_1984_no_snow,
  "gte_1984_w_mixed_forecast_obs" = params_set_gte_1984_w_mixed_forecast_obs
)

# cram those into table!
df_parameter_configs <- tibble(
  param_set_id = names(l_params),
  parameter_sets = l_params,
  earliest_year =parse_number(param_set_id)
)
```

### Weight implementation

```{r}

# run weighting and aggregations

# For each paramter set (i.e gte1984, gte_1984_no_snow, gte2001) this returns
# 1. a historical data set per model composition
# 2. an assessment of f1 performance of each model set measuring against
# reference framework of just using ASI

# this takes approx 10 minutes to run on my computer. There are many millions
# of rows .. could save as a file, but it will take a long time even load
# from the blob.

system.time(
  ldf_perf <- pmap(
  list(
    df_parameter_configs$param_set_id,
    df_parameter_configs$parameter_sets,
    df_parameter_configs$earliest_year
    ),
  \(id,params, start_yr){

    # cat("reclassifying\n")
    df_all_weights_flagged <- weighted_classify(
      df = df_env_model,
      earliest_year = start_yr,
      params_included = unlist(params),
      weight_values = seq(0,1,by =0.05),
      rp=3
    )

    df_asi_perf <-  single_indicator_performance(
      df_env_model,
      parameter = "asi",
      earliest_year = start_yr,
      rp=3
    )

    df_perf <- df_all_weights_flagged |>
      summarise_performance() |>
       left_join(
        df_asi_perf |>
          rename(asi_f1 = .estimate)
        )

    l_ret <- list(
      ALL_MODELS = df_perf,
      HISTORICAL_WEIGHTED = df_all_weights_flagged
    )
    l_ret
  }
)
)

ldf_perf <- set_names(ldf_perf,df_parameter_configs$param_set_id)

# extract historical data.frame sets
ldf_historical <-  ldf_perf |>
  map("HISTORICAL_WEIGHTED")

ldf_perf_all_models <- ldf_perf |>
  map("ALL_MODELS")

ldf_top3 <- ldf_perf_all_models |>
  map(\ (dft){
    dft |>
      top_performance_per_moment(n=3)
  })


```

```{r}
#| eval: false

cumulus$blob_write(
  ldf_perf$gte1984$HISTORICAL_WEIGHTED ,
  container = "projects",
  name = "ds-aa-afg-drought/weight_parameter_set/gte1984/historical_weighted.parquet"
)
ck <- cumulus$blob_read(
    container = "projects",
    name = "ds-aa-afg-drought/weight_parameter_set/gte1984/historical_weighted.parquet"
)
```

```{r}
#| eval: false

cumulus$blob_write(
  ldf_perf$gte_1984_w_mixed_forecast_obs$HISTORICAL_WEIGHTED ,
  container = "projects",
  name = "ds-aa-afg-drought/weight_parameter_set/gte1984/historical_weighted_w_mixed_fcast_obsv.parquet"
)
```

### Weighting results

Here we look at the results when we use parameters going back to 1984 and include a snow cover variable. These plots were put in the google slides

```{r}

ldf_top3$gte1984 |>
  filter(pub_mo_label!="Jun") |>
  group_by(pub_mo_label, adm1_name) |>
  slice_max(avg_estimate,with_ties = F) |>
  # left_join(df_asi_performance |> rename(asi_f1 = .estimate)) |>
  pivot_longer (cols = c(".estimate","asi_f1")) |>
  mutate(
    indicator_label = ifelse(name == ".estimate", "Combined Indicator","Simple Indicator (ASI)" )
  ) |>
  ggplot(
    aes(x = adm1_name, y =value, fill = indicator_label)
  )+
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Indicator (ASI)"=hdx_hex("sapphire-light"))
  )+
  facet_wrap(~pub_mo_label)+
  scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: Composite Indicator vs ASI Only",
    subtitle= "Analysis 1984-present (including ERA5 snow)",
    caption= "Composite indicator contains optimal weighted aggregations of Snow Cover (ERA5), Soil Moisture (ERA5), ASI (FAO), VHI (FAO), and cumulative precipitation (ERA 5)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

pal_indicator_composition <-  c("ASI"="green4",
                                "VHI"="#CAB2D6",
                                "NDSI"= "#6A3D9A",
                                "Precip cumu (ERA)"= "skyblue2",
                                "era5_land_soil_moisture_1m"= "dodgerblue2",
                                "Snow Cover (ERA5)"= "white")


p_optimal_low_variance_weights <- plot_low_var_optimal(
  ldf_top3$gte1984,
  pal = pal_indicator_composition,label_wt_id = F
)+
  coord_flip()


 plot_optimal_compositions(
  ldf_top3$gte1984  |>
    filter(pub_mo_label != "Jun"),
  label_plot = T,
  pal = pal_indicator_composition
  ) +
  labs(
      title ="Indicator weighting options that result in max performance"
    )


df_compare <- compare_to_simple_model(
  ldf_historical$gte1984,
  wt_id_apr = 3316,#379 -- was optimal when did weight by 0.1 #14 is optimal
  wt_id_may = 3490, #142 & 84 are optimal for May, but dont include soil moisture
  param_simple = "asi"
  )

p_optimal_chosen_compared_to_asi <- df_compare |>
  pivot_longer(cols = c(".estimate","estimate_simple")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","Simple Model (ASI ONLY)")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge", color = "lightgrey") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Model (ASI ONLY)"=hdx_hex("sapphire-light"))
  )+
  labs(
    title = "F1 Score Comparison: April & May activation moments",
    subtitle= "Composite Indicator (one weight set per activation moment) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  facet_wrap(~pub_mo_label)+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

p_optimal_chosen_compared_to_asi +
  theme(
    plot.caption = element_blank(),legend.position = "top"
  )+
  p_optimal_low_variance_weights+
  theme(
    axis.title.y= element_blank(),
    axis.text.y= element_blank(),
    strip.text = element_blank(),
    legend.title = element_blank()
  )+
  plot_layout(nrow=2,heights = c(1,0.05))

```

### With mixed forecast observational

-   Now let's look at when we include the mixed observational forecast indicators
-   so here are the best performing modesl compared to f1

```{r}
ldf_top3$gte_1984_w_mixed_forecast_obs |>
  filter(pub_mo_label!="Jun") |>
  group_by(pub_mo_label, adm1_name) |>
  slice_max(avg_estimate,with_ties = F) |>
  # left_join(df_asi_performance |> rename(asi_f1 = .estimate)) |>
  pivot_longer (cols = c(".estimate","asi_f1")) |>
  mutate(
    indicator_label = ifelse(name == ".estimate", "Combined Indicator","Simple Indicator (ASI)" )
  ) |>
  ggplot(
    aes(x = adm1_name, y =value, fill = indicator_label)
  )+
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Indicator (ASI)"=hdx_hex("sapphire-light"))
  )+
  facet_wrap(~pub_mo_label)+
  scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: Composite Indicator vs ASI Only",
    subtitle= "Analysis 1984-present (including ERA5 snow)",
    caption= "Composite indicator contains optimal weighted aggregations of Snow Cover (ERA5), Soil Moisture (ERA5), ASI (FAO), VHI (FAO), and cumulative precipitation (ERA 5)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )
```

but let's compare the weights w/ less variance - `pseudo-regularization`

```{r}

# this is where i get the weight ids for ploting
plot_low_var_optimal(
  ldf_top3$gte_1984_w_mixed_forecast_obs,
  pal = pal_indicator_composition,label_wt_id = T
)+
  coord_flip()



# but in plot i dont need to show the weight id - so save this too
p_optimal_low_variance_weights <- plot_low_var_optimal(
  ldf_top3$gte_1984_w_mixed_forecast_obs,
  pal = pal_indicator_composition,label_wt_id = F
)+
  coord_flip()


# there are too many ties -- this takes forecer to render and is unreadable :-)
 # plot_optimal_compositions(
 #  ldf_top3$gte_1984_w_mixed_forecast_obs  |>
 #    filter(pub_mo_label != "Jun"),
 #  label_plot = T,
 #  pal = pal_indicator_composition
 #  ) +
 #  labs(
 #      title ="Indicator weighting options that result in max performance"
 #    )


# this takes a minute

df_compare_gte_1984_w_mixed_forecast_obs <- compare_to_simple_model(
  ldf_historical$gte_1984_w_mixed_forecast_obs,
  wt_id_apr = 9834,
  wt_id_may = 12297, #142 & 84 are optimal for May, but dont include soil moisture
  param_simple = "asi"
  )


pal_indicator_composition <-  c("ASI"="green4",
                                "VHI"="#CAB2D6",
                                "NDSI"= "#6A3D9A",
                                "Precip cumu (ERA)"= "skyblue2",
                                "era5_land_soil_moisture_1m"= "dodgerblue2",
                                "mam_mixed_seas_observed" = "cyan",
                                "Snow Cover (ERA5)"= "white")



p_comparison_1984_w_mixed_forecast_obs <- df_compare_gte_1984_w_mixed_forecast_obs |>
  pivot_longer(cols = c(".estimate","estimate_simple")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","Simple Model (ASI ONLY)")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge", color = "lightgrey") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Model (ASI ONLY)"=hdx_hex("sapphire-light"))
  )+
  labs(
    title = "F1 Score Comparison: April & May activation moments",
    subtitle= "Composite Indicator (one weight set per activation moment) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  facet_wrap(~pub_mo_label)+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

p_comparison_1984_w_mixed_forecast_obs +
  theme(
    plot.caption = element_blank(),legend.position = "top"
  )+
  p_optimal_low_variance_weights+
  theme(
    axis.title.y= element_blank(),
    axis.text.y= element_blank(),
    strip.text = element_blank(),
    legend.title = element_blank()
  )+
  plot_layout(nrow=2,heights = c(1,0.15))
```

## Historical timeline

Make tile plot showing when simple indicator - composite indicator - and truth (ASI EOS) would have activated.

```{r}
simple_model_id <- get_simple_weight_id(
  df= ldf_historical$gte1984,
  param_simple = "asi"
  )


optimal_model_wt_id <-  c(14,379,3316)[3]
df_simple_historical <- ldf_historical$gte1984 |>
  select(
    yr_season,wt_id,wt_set, pub_mo_label, adm1_name,ends_with("flag")

  ) |>
  pivot_longer(cols = ends_with("flag")) |>
  filter(wt_id ==simple_model_id,name== "zscore_flag") |>
  mutate(
    weight_type = "Simple (ASI only)"
  )
df_simple_historical |> count(value)

df_historical_weights_sel <- ldf_historical$gte1984 |>
  select(
    yr_season,wt_id,wt_set, pub_mo_label, adm1_name,ends_with("flag")

  ) |>
  pivot_longer(cols = ends_with("flag")) |>
  filter(
    wt_id == optimal_model_wt_id & pub_mo_label=="Apr"
  ) |>
  mutate(
    weight_type =ifelse(name == "zscore_flag","Combined","EOS ASI\n'Truth'")
  )


bind_rows(
  df_historical_weights_sel,
  df_simple_historical
) |>
  filter(adm1_name == "Takhar",pub_mo_label == "Apr") |>
  mutate(
    yr = year(yr_season),
    weight_type = factor(weight_type, c("Simple (ASI only)","Combined","EOS ASI\n'Truth'"))
  ) |>
  ggplot()+
  geom_tile(
    aes(x = weight_type,y= yr,  fill = value), color = "white",linewidth = 0.7
  ) +
  scale_fill_manual(
    values = c("TRUE"=hdx_hex("tomato-hdx"), "FALSE"="grey")
  )+
  scale_y_reverse(breaks= 1984:2024,
                  expand= c(0,0)
                  # limits = c(2024,1984)
                  )+
  scale_x_discrete(expand =c(0,0))+
  labs(
    title = "Historical analyses of indicators",
    subtitle = "April - Takhar Province",
    caption = glue("optimal model id: {optimal_model_wt_id}")
  )+
  facet_wrap(~adm1_name)+
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  )
```

```{r}
ldf_top3$gte2001 |>
  filter(pub_mo_label!="Jun") |>
    group_by(pub_mo_label, adm1_name) |>
  slice_max(avg_estimate,with_ties = F) |>
  # left_join(df_asi_performance |> rename(asi_f1 = .estimate)) |>
  pivot_longer (cols = c(".estimate","asi_f1")) |>
  mutate(
    indicator_label = ifelse(name == ".estimate", "Combined Indicator","Simple Indicator (ASI)" )
  ) |>
  ggplot(
    aes(x = adm1_name, y =value, fill = indicator_label)
  )+
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Indicator (ASI)"=hdx_hex("sapphire-light"))
  )+
  facet_wrap(~pub_mo_label)+
  scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: Composite Indicator vs ASI Only",
    subtitle= "Analysis 2001-present (including MODIS Snow -NDSI)",
    caption= "Composite indicator contains optimal weighted aggregations of NDSI Snow Cover (MODIS), Soil Moisture (ERA5), ASI (FAO), VHI (FAO), and cumulative precipitation (ERA 5)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

ldf_top3$gte_1984_no_snow |>
  filter(pub_mo_label!="Jun") |>
    group_by(pub_mo_label, adm1_name) |>
  slice_max(avg_estimate,with_ties = F) |>
  # left_join(df_asi_performance |> rename(asi_f1 = .estimate)) |>
  pivot_longer (cols = c(".estimate","asi_f1")) |>
  mutate(
    indicator_label = ifelse(name == ".estimate", "Combined Indicator","Simple Indicator (ASI)" )
  ) |>
  ggplot(
    aes(x = adm1_name, y =value, fill = indicator_label)
  )+
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(
    values = c("Combined Indicator"=hdx_hex("tomato-light"), "Simple Indicator (ASI)"=hdx_hex("sapphire-light"))
  )+
  facet_wrap(~pub_mo_label)+
  scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: Composite Indicator vs ASI Only",
    subtitle= "Analysis 1984-Present (No Snow included)",
    caption= "Composite indicator contains optimal weighted aggregations of  Soil Moisture, ASI, VHI, and cumulative precipitation",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )
```

# Check point

some of below probably not running or correct anymore.

double check RP'S

```{r}
ldf_historical$gte1984 |>
  group_by(
    adm1_name,
    #yr_season,
    pub_mo_label,
    wt_id,
    wt_set,
  ) |>
  summarise(
    activation_rate_wt = mean(zscore_flag),
    activation_rate_truth = mean(zscore_asi_Jun_flag)
  )
ldf_historical$gte2001 |>
  group_by(
    adm1_name,
    #yr_season,
    pub_mo_label,
    wt_id,
    wt_set,
  ) |>
  summarise(
    activation_rate_wt = mean(zscore_flag),
    activation_rate_truth = mean(zscore_asi_Jun_flag)
  )

ldf_historical$gte_1984_no_snow |>
  group_by(
    adm1_name,
    #yr_season,
    pub_mo_label,
    wt_id,
    wt_set,
  ) |>
  summarise(
    activation_rate_wt = mean(zscore_flag),
    activation_rate_truth = mean(zscore_asi_Jun_flag)
  )

```

```{r}
df_truth_joint <- df_all_weights_flagged |>
  distinct(
    yr_season,adm1_name, zscore_asi_Jun_flag
  ) |>
  group_by(
    yr = year(yr_season)
  ) |>
  summarise(
    flag = any(zscore_asi_Jun_flag)
  ) |>
  mutate(
    indicator_type = "Truth"
  )

 df_apr_joint <- df_all_weights_flagged |>
  filter(
    (pub_mo_label == "Apr" &
    wt_id %in% c(1,684)
    )) |>
    mutate(
      yr= year(yr_season),
    indicator_type = case_when(
      wt_id == 1 ~ "Simple (ASI only)",
      # name== "zscore_asi_Jun_flag"~"Truth",
      .default= "Composite",
      # .default = "combined"

  )
    ) |>
   select(yr,indicator_type, pub_mo_label, adm1_name,zscore_flag) |>
   group_by(
     yr, indicator_type, pub_mo_label
   ) |>
   summarise(
     flag = any(zscore_flag)
   )
df_apr_joint |>
  bind_rows(
    df_truth_joint
  ) |>
  group_by(
    indicator_type
  ) |>
  summarise(
    flag = mean(flag)
  )
df_apr_joint |>
  bind_rows(
    df_truth_joint
  ) |>
  ggplot()+
  geom_tile(
    aes(x = indicator_type,y= yr,  fill = flag), color = "white"
  ) +
  scale_fill_manual(
    values = c("TRUE"=hdx_hex("tomato-hdx"), "FALSE"="grey")
  )+
  labs(
    title = "April Joint"
  )




```

```{r}
box::use(forcats[...])
df_historical_rethresh <- ldf_historical$gte1984 |>
    utils$threshold_var(
      var= "zscore",
      by = c("pub_mo_label","adm1_name","wt_id"),
      rp_threshold = 4
    ) |>
    utils$threshold_var(
      var= "zscore_asi_Jun",
      by = c("pub_mo_label","adm1_name","wt_id"),
      rp_threshold = 4
    ) |>
  mutate(
    across(ends_with("_flag"),\(x){fct_relevel(as_factor(x),"TRUE","FALSE")})
  )


df_rethresh_f1 <- df_historical_rethresh |>
  group_by(
    across(c("pub_mo_label","adm1_name","wt_id"))
  ) |>
  f_meas(truth = zscore_asi_Jun_flag,estimate=zscore_flag)

df_rethresh_f1 |>
  group_by(pub_mo_label,wt_id) |>
  mutate(
    avg_f1 = mean(.estimate)
  ) |>
  group_by(pub_mo_label,adm1_name) |>

  slice_max(
    order_by = avg_f1,
    n=1 , with_ties = F
  )
```

```{r}
# okay -- above we get best weights per adm and month.
# can we instead test each set of weights against all provinces and get the highest f1 score
# that works per weight

l_top_wt_sets_per_mo <- ldf_conf_metrics  |>
  map(
    \(dft){
      dft |>
        group_by(pub_mo_label, wt_id) |>
        summarise(
          avg_estimate = mean(.estimate)
        ) |>
        slice_max(
          order_by = avg_estimate,
          n= 1
        )

    }

  )

ldf_best_metrics_overall<- map2(ldf_conf_metrics,l_top_wt_sets_per_mo,
  \(dft_conf,dft_top){
          dft_conf |>
        inner_join(
          dft_top, by = c("pub_mo_label","wt_id")
          )
  }
)
ldf_best_metrics_overall<- map(named_weight_sets,
  \(nmt){
    dft_conf <- ldf_conf_metrics[[nmt]]
    dft_top <- l_top_wt_sets_per_mo[[nmt]]
    dft_conf |>
        inner_join(
          dft_top, by = c("pub_mo_label","wt_id")
          )
  }
)


ldf_top_weights_nested <-  map2(l_top_wt_sets_per_mo,l_weight_combos,
     \(df_top,df_wt_set){
     df_top_wt_vals <- df_top |>
         inner_join(
           df_wt_set |>
             group_by(wt_id) |>
             nest()
           # by = "wt_id"
           # wt_id %in% df_top$wt_id
         )
         # utils$label_parameters()
     }
 )


ldf_best_metrics_overall$gte1984 |> select(wt_set) |> unnest(wt_set) |> count(parameter)



lp_optimal_weight_composition <- map(ldf_best_metrics_overall,
     \(dft){

       dft <- dft|>
         select(
           pub_mo_label,adm1_name,wt_set,avg_estimate,.estimate
         ) |>
         unnest(wt_set) |>
         utils$label_parameters()

       dft |>
         group_by(pub_mo_label, adm1_name) |>
         mutate(
           pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
           id = dense_rank(wt_id)
         ) |>
         ungroup() |>
         ggplot(
           aes(x= id, y= weight,fill = parameter_label)
         )+
         geom_bar(
           stat= "identity", color = "grey"
         )+

         scale_fill_manual(values=rand_pal)+
         facet_grid(cols= vars(pub_mo_facet),
                    rows = vars(adm1_name)
                    ,scales="free")+
         labs(
           # title = "May publication weight combos that give avg 0.90 f1"
         )


     }
)
lp_optimal_weight_composition$gte2000
lp_optimal_weight_composition$gte1984

lp_weight_performance_per_adm <- map(
  ldf_best_metrics_overall,\(dft){

         dft <- dft |>
           mutate(
             pub_mo_facet = factor(pub_mo_label, levels = c("Apr","May","Jun"))
           )

         dft |>
           ggplot(
             aes(
               x= adm1_name,
               y= .estimate,
               group = wt_id
               # color =as_factor(wt_id)
             )
           )+
           geom_point(alpha=0.3,size=4)+
           geom_line(alpha=0.3)+
           facet_wrap(~pub_mo_facet)+
           geom_hline(
             data = dft |>
               select(
                 pub_mo_facet,
                 adm1_name,
                 avg_estimate
                 ) |> distinct(),
             aes(yintercept = avg_estimate),
             linetype = "dashed",
             color = hdx_hex("tomato-hdx")
           )+
           labs(
             y= "F1 Performance"
           )+
           theme(
             panel.border = element_rect(fill=NA, color = "grey")
           )

     }
)


lp_weight_performance_per_adm$gte2000
lp_weight_performance_per_adm$gte1984



# i like these, but problem is faryab actually performs worse
dft_low_sd_f1 <- ldf_best_metrics_overall$gte2000 |>
  select(
    pub_mo_label, wt_set
  ) |>
  unnest(wt_set) |>
  group_by(
    pub_mo_label,wt_id
  ) |>
  mutate(
    sd = sd(weight)
  ) |>
  group_by(pub_mo_label) |>
  slice_min(sd) |>
  group_by(pub_mo_label) |>
  mutate(
    pub_mo_facet = factor(pub_mo_label,levels= c("Apr","May","Jun")),
    id = dense_rank(wt_id)
  ) |>
  ungroup() |>
  utils$label_parameters()

dft_low_sd_f1 |>
  distinct() |>
  mutate(
    weight_pct_label = scales::percent(weight,accuracy = 1, trim = FALSE),
    weight_id_label = glue("weight id: {wt_id}")

  ) |>
  filter(pub_mo_label!= "Jun") |>
  ggplot(
    aes(x= id, y= weight,fill = parameter_label), position = position_stack(vjust = 0.5)
  )+
  geom_bar(
    stat= "identity"
  )+
  geom_text(aes(x= id, y= weight, label = weight_pct_label),
            position = position_stack(vjust = 0.5), color ="white")+
  geom_text(aes(x= id, y= 1.01, label = weight_id_label), color ="black")+
  scale_fill_manual(values=rand_pal)+
  facet_grid(cols= vars(pub_mo_facet)
             ,scales="free")+
  labs(
    # title = "May publication weight combos that give avg 0.90 f1"
  )


ldf_best_metrics_overall$gte2000 |>
  filter(
    pub_mo_label == "Apr",
         wt_id ==684 # if filtering to >20000
         ) |>
  left_join(
      ldf_conf_metrics$gte2000 |>
      filter(wt_id ==1) |>
      # select(-wt_id) |>
      rename(
        asi_only_f1 = .estimate
        ) |>
        select(pub_mo_label,adm1_name, asi_only_f1)
   , by = c("pub_mo_label","adm1_name")
  ) |>
  pivot_longer(cols = c(".estimate","asi_only_f1")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: April Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

ldf_best_metrics_overall$gte2000 |>
  filter(
    pub_mo_label == "May",
         wt_id ==524 # if filtering to >20000
         ) |>
  left_join(

      ldf_conf_metrics$gte2000 |>
      filter(wt_id ==1) |>
      # select(-wt_id) |>
      rename(
        asi_only_f1 = .estimate
        ) |>
        select(pub_mo_label,adm1_name, asi_only_f1)
   , by = c("pub_mo_label","adm1_name")
  ) |>
  pivot_longer(cols = c(".estimate","asi_only_f1")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: May Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )












df_best_wt_sets_w_metrics |>
  print(n=100) |>
  filter(pub_mo_label=="May") |>
  select(wt_set,avg_estimate) |>
  # pull(avg_estimate) |>
  unnest(wt_set) |>
  distinct() |>
  mutate(
    id = dense_rank(wt_id)
  ) |>
  ggplot(
    aes(x= id, y= weight,fill = parameter)
  )+
  geom_bar(
    stat= "identity"
  )+
  labs(
    title = "May publication weight combos that give avg 0.90 f1"
  )




top_wt_sets_per_mo <- df_conf_metrics  |>
  group_by(pub_mo_label, wt_id) |>
  summarise(
    avg_estimate = mean(.estimate)
  ) |>
  slice_max(
    order_by = avg_estimate,n= 1
  )
df_best_metrics_overall <- df_conf_metrics |>
  inner_join(
    top_wt_sets_per_mo, by = c("pub_mo_label","wt_id")
  )
df_best_wt_sets_w_metrics <- df_best_metrics_overall |>
  left_join(
    df_all_weights_flagged |>
      distinct(wt_id,wt_set)
    ) |>
  left_join(
      df_conf_metrics |>
      filter(wt_id ==1) |>
      # select(-wt_id) |>
      rename(
        asi_only_f1 = .estimate
        ) |>
        select(pub_mo_label,adm1_name, asi_only_f1)
   , by = c("pub_mo_label","adm1_name")
  )

df_best_wt_sets_w_metrics |>
  select(-wt_id) |>
  unnest(wt_set)


df_best_wt_sets_w_metrics |>
  filter(
    pub_mo_label == "Apr",
         wt_id ==346 # if filtering to >20000
         ) |>
  pivot_longer(cols = c(".estimate","asi_only_f1")) |>
  mutate(
    name = ifelse(str_detect(name,".estimate"), "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= value, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: April Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )

df_wt_combos_long |>
  filter(wt_id  ==346)
```

Look at equal weigts

```{r}
df_conf_metrics |>
  filter(pub_mo_label == "Apr") |>
  filter(
    wt_id %in% c(1,equal_wts_id)
  ) |>
    mutate(
    name = ifelse(wt_id == equal_wts_id, "Combined Indicator","ASI only")
  ) |>
  ggplot(
    aes(x= adm1_name,y= .estimate, fill = name )
  ) +
  geom_bar(stat= "identity", position = "dodge") +
    scale_y_continuous(
    breaks = seq(0,1,by = 0.05)
  )+
  labs(
    title = "F1 Score Comparison: April Publication Month",
    subtitle= "Composite Indicator (one optimal weight set applied) vs ASI Only",
    caption= "Composite indicator created via weighted aggregation of z-scores of cumulative rainfall (ERA5), snow cover (NDSI), Soil Moisture (ERA5), VHI (FAO), ASI (FAO)",
    y= "F1 Score"
  )+
  theme(
    legend.title=element_blank(),
    axis.title.x = element_blank()
  )
)

```

```{r}
df_all_weights_flagged
df_weighted_historical_sample <- df_all_weights_flagged |>
  filter(pub_mo_label == "Apr",adm1_name == "Takhar")

df_end_of_season_valid <- df_weighted_historical_sample |>
  select(yr_season,zscore_asi_Jun_flag) |>
  distinct() |>
  mutate(
    name = "ASI\nEnd of Season\n('truth')"
  ) |>
  rename(zscore_flag = zscore_asi_Jun_flag)

bind_rows(
  df_weighted_historical_sample ,
  df_end_of_season_valid
)

df_weighted_historical_sample |>
filter(
    wt_id %in% c(1,equal_wts_id,38)
  ) |>
  bind_rows(
    df_end_of_season_valid |>
      rename()
  ) |>
  mutate(
    yr = year(yr_season),
    name = case_when(
      wt_id == 1~"ASI only",
      wt_id == equal_wts_id~"Combined (equal)",
      wt_id == 38 ~"Combnied (optimal)",
      .default = name
      )
  ) |>
  ggplot()+
  geom_tile(
    aes(x = name,y= yr,  fill = zscore_flag), color = "white"
  ) +
  scale_fill_manual(
    values = c("TRUE"="red", "FALSE"="grey")
  )
```

```{r}

df_best_wt_sets_w_metrics |>
  filter(
    wt_id == 39
  ) |>
  select(-wt_id) |>
  unnest(wt_set)

df_best_wt_sets_w_metrics |>
  print(n=100) |>
  filter(pub_mo_label=="May") |>
  select(wt_set,avg_estimate) |>
  # pull(avg_estimate) |>
  unnest(wt_set) |>
  distinct() |>
  mutate(
    id = dense_rank(wt_id)
  ) |>
  ggplot(
    aes(x= id, y= weight,fill = parameter)
  )+
  geom_bar(
    stat= "identity"
  )+
  labs(
    title = "May publication weight combos that give avg 0.90 f1"
  )



df_conf_metrics |>
  filter(
    wt_id ==38 & pub_mo_label=="Apr"
  )




### Simlarity Metrics #######
xck_nest <- df_wt_combos_long |>
  group_by(wt_id) |>
  nest()

df_max_f1_apr  |>
  left_join(ck_nest) |>
  unnest(data)



ck <- df_max_f1_apr  |>
  left_join(ck_nest) |>
  unnest(data)



aligned_weights <- ck %>%
  group_by(adm1_name, wt_id, parameter) %>%
  summarise(weight = sum(weight), .groups = "drop") %>% # Aggregate weights if duplicates
  pivot_wider(names_from = parameter, values_from = weight, values_fill = 0) |>
  mutate(
    row_id = row_number()
  )

pairwise_comparisons <- aligned_weights %>%
  # mutate(row_id = row_number()) %>% # Add a unique ID to each row for pairing
  cross_join(aligned_weights, suffix = c("_x", "_y")) %>% # Generate all pairs
  filter(row_id_x < row_id_y) # Avoid duplicate comparisons and self-comparisons

pairwise_comparisons |>
  glimpse()
# Step 3: Compute similarity metrics
similarity_results |>
  glimpse()

similarity_results <- pairwise_comparisons %>%
  rowwise() %>%
  mutate(
    # Extract weight vectors using regex pattern
    weight_vector_x = list(c_across(matches("^(cumu.*_x|era5_land.*_x|NDSI.*_x|vhi_x|asi_x)$"))),
    weight_vector_y = list(c_across(matches("^(cumu.*_y|era5_land.*_y|NDSI.*_y|vhi_y|asi_y)$")))
  )

similarity_results <- similarity_results %>%
  rowwise() %>%
  mutate(
    # Calculate Euclidean distance
    manhattan_distance = sum(abs(unlist(weight_vector_x) - unlist(weight_vector_y)), na.rm = TRUE)
    # euclidean_distance = sqrt(sum((unlist(weight_vector_x) - unlist(weight_vector_y))^2, na.rm = TRUE)),
    # # Calculate Cosine similarity
    # cosine_similarity = sum(unlist(weight_vector_x) * unlist(weight_vector_y), na.rm = TRUE) /
    #   (sqrt(sum(unlist(weight_vector_x)^2, na.rm = TRUE)) * sqrt(sum(unlist(weight_vector_y)^2, na.rm = TRUE)))
  )

most_similar_weight_set <- similarity_results %>%
  group_by(wt_id_x) %>%
   summarise(
    # avg_jsd_distance = mean(jsd_distance, na.rm = TRUE),  # For JSD
    avg_manhattan_distance = mean(manhattan_distance, na.rm = TRUE)  # For Manhattan
  ) %>%
  # Rank by lowest JSD or Manhattan distance (lower is better)
  arrange(avg_manhattan_distance) %>%
  slice(1)

df_agg_all_weight_combos
df_wt_combos_long |>
  filter(wt_id == 883)

# Step 4: Select relevant columns for clarity
final_results <- similarity_results %>%
  select(
    adm1_name_x, wt_id_x,
    adm1_name_y, wt_id_y,
    euclidean_distance, cosine_similarity
  )

# View results
final_results |>
  ungroup() |>
  filter(adm1_name_x!= adm1_name_y)
############


id_simple_asi_model <- df_wt_combos_long |>
  filter(
    parameter == "asi",
    weight ==1
  ) |>
  pull(
    wt_id
  )

df_simple_asi_weight_set <- df_wt_combos_long |>
  filter(wt_id %in% id_simple_asi_model) |>
  group_by(wt_id) |>
  nest()




df_weight_sets_nested <- df_wt_combos_long |>
  group_by(wt_id) |>
  nest()



df_max_f1 |>
  group_by(
    pub_mo_label, wt_id
  ) |>
  mutate(
    num_wt_sets = n()
  ) |>
  filter(num_wt_sets>1)

df_max_f1_co_occur <- df_max_f1  |>
  group_by(
    pub_mo_label, wt_id
  ) |>
  mutate(
    num_wt_sets = n()
  ) |>
  ungroup() |>
  filter(num_wt_sets>1) |>
  left_join(df_weight_sets_nested, by ="wt_id") |>
  group_by(adm1_name,pub_mo_label) |>
  unnest(data)

# we have 3 clear options for June - def easiest to just go w/ ASI and/or VHI
df_max_f1_co_occur |>
  ungroup() |>
  filter(num_wt_sets>2) |>
  print(n=45) |>
  distinct(pub_mo_label,wt_id, parameter,weight)

df_max_f1_co_occur |>
  print(n=45) |>
  distinct(pub_mo_label,wt_id, parameter,weight)

df_max_f1_shared <- df_max_f1  |>
  # group_by(
  #   pub_mo_label, wt_id
  # ) |>
  # mutate(
  #   num_wt_sets = n()
  # ) |>
  # ungroup() |>
  # filter(num_wt_sets>1) |>
  left_join(df_weight_sets_nested, by ="wt_id") |>
  group_by(adm1_name,pub_mo_label) |>
  unnest(data) |>
  mutate(
    wt_id_dense = dense_rank(wt_id)
  ) |>
  ungroup() |>
  mutate(
    pub_mo_label = fct_relevel(pub_mo_label, "Apr","May","Jun")
  )


df_f1_labels <- df_max_f1_shared |>
  distinct(pub_mo_label, adm1_name,.estimate) |>
    left_join(
    df_conf_metrics |>
      filter(wt_id ==1) |>
      group_by(pub_mo_label,adm1_name) |>
      transmute(
        simple_estimate = scales::label_number(accuracy = 0.01)(.estimate)
  ), by = c("pub_mo_label","adm1_name")
  ) |>
      mutate(
    pub_mo_label = fct_relevel(pub_mo_label, "Apr","May","Jun")
  )


df_max_f1_shared |>
    mutate(
    pub_mo_label = fct_relevel(pub_mo_label, "Apr","May","Jun")
  ) |>
# filter(pub_mo_label %in% c("May","Jun"), adm1_name == "Sar-e-Pul") |>
  ggplot(
  )+
  geom_bar(stat="identity", aes(x= wt_id_dense, y= weight, fill = parameter))+
  # geom_text(size = 3, position = position_stack(vjust = 0.5))
  # geom_text(
  #   aes(
  #     label = scales::label_percent(accuracy=1)(weight)
  #     ),
  #   position = position_stack(vjust = 0.5), color ="white"
  # )+
  geom_label(
    data= df_f1_labels,
    x= 50, y= 0.5,color="red",alpha = 0.6,
    aes(
        label = glue(
          "Max F1: {scales::label_number(accuracy=0.01)(.estimate)}\nASI Only F1: {simple_estimate}"

          )
    )
  )+
  facet_grid(
    cols = vars(adm1_name),
    rows = vars(pub_mo_label),
    # labeller = labeller(adm1_name=~facet_label),
    scales = "free_x")+
  labs(
    title = "Weight combinations for optimal F1 scores by data publication date and province",
    subtitle = "Optimal F1 Score & F1 Score of simple model (ASI only shown)"
  )+
  theme(
    axis.title.x =element_blank(),
    axis.text.x =element_blank(),
    # strip.switch.pad.grid = unit(1, "in")
  )


dft_sel <- dft |>
  pivot_longer(
    cols = -c("yr_season","pub_mo_label","adm1_name","pub_mo_date"),
    names_to = "parameter"
  ) |>
  filter(
    parameter %in% c("era5_land_soil_moisture_1m","cumu_era5_land_total_precipitation_sum","NDSI_Snow_Cover_mean","vhi","asi")
  ) |>
  group_by(
    across(-c("value","yr_season","pub_mo_date"))
  )
dft_sel |>
  distinct(parameter)
scale(dft_sel$value )[,1]
dft_scaled <- dft_sel |>
  mutate(
    value_centered = scale(value, center= T,scale=F)[,1],
    zscore = scale(value,center=T,scale=T)[,1],
    across(
      c("value_centered","zscore"),~ifelse(parameter != "asi",.x*-1,.x)
    )
  )

dft_scaled |>
  # filter(parameter == "asi") |>
  arrange(parameter)

# unweighted
dft_z_agg_unweighted <- dft_scaled |>
  group_by(
   yr_season ,
   pub_mo_label,
   adm1_name,
   pub_mo_date
  ) |>
  summarise(
    across(
      .cols = c("zscore","value_centered"),
      .fns = \(x) mean(x,na.rm=T)
    ),.groups="drop"
  )


dft_unweighted_pred <- dft_z_agg_unweighted |>
  left_join(
    df_jun_asi_scaled_short
  ) |>
  threshold_var(
    var= "zscore",
    by = c("pub_mo_label","adm1_name"),
    rp_threshold = 3
  ) |>
  threshold_var(
    var= "zscore_asi_Jun",
    by = c("pub_mo_label","adm1_name"),
    rp_threshold = 3
  ) |>
  arrange(adm1_name, yr_season) |>
  mutate(
    across(ends_with("_flag"),~as_factor(.x))
  ) |>
  ungroup()

recall(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag,event_level ="second")
precision(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag,event_level ="second")
f_meas(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag,event_level ="second")
metrics(dft_unweighted_pred,truth = zscore_asi_Jun_flag,estimate = zscore_flag)

```
