---
title: "2026 Trigger Proposal (Combined RP)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  warning: false
  message: false
project:
  execute-dir: project
---

## Introduction

This chapter proposes a **two-stage trigger** for the 2026 anticipatory action framework. The idea is to combine two independent signals at different lead times using OR logic:

1. **SEAS5 March window** (early lead time): If the COPERNICUS SEAS5 seasonal precipitation forecast exceeds a return period threshold, trigger early.
2. **CDI April window** (better observational data): If the ridge-based Combined Drought Index exceeds a return period threshold, trigger in April.

A year triggers if **either** signal fires. This increases sensitivity (fewer missed droughts) at the cost of more frequent activation.

See @sec-2026-trigger-modeling for CDI model derivation and validation.

```{r}
#| label: setup

box::use(
    dplyr[...],
    tidyr[...],
    ggplot2[...],
    gghdx[...],
    cumulus[...],
    purrr[...],
    glue[glue],
    tibble[tibble],
    .. / R / utils[rp_empirical]
)

library(tidymodels)
library(gt)
gghdx()
```

```{r}
#| label: config

# Return period threshold for defining drought
RP_THRESHOLD <- 4

# Random seed
SEED <- 42

# Bootstrap resamples
N_BOOTSTRAP <- 30

# Feature set paths
FEATURE_PATHS <- list(
    march = "ds-aa-afg-drought/processed/vector/2026_mar_pub_feature_set_v1.parquet",
    april = "ds-aa-afg-drought/processed/vector/2026_april_pub_feature_set_v1.parquet"
)

# Variables to exclude from CDI model (same as ch12)
APRIL_EXCLUDE_VARS <- c("total_precipitation_sum", "seas5 Apr", "seas5 May")
```

## Data & Model

```{r}
#| label: load-data

# Load feature sets
df_mar_raw <- blob_read(name = FEATURE_PATHS$march, container = "projects")
df_apr_raw <- blob_read(name = FEATURE_PATHS$april, container = "projects")

# Prepare modeling data with empirical RP thresholding
prepare_model_data <- function(df, rp_threshold = RP_THRESHOLD) {
    df |>
        filter(!is.na(outcome_asi_zscore)) |>
        arrange(desc(outcome_asi_zscore)) |>
        mutate(
            rank = row_number(),
            q_rank = rank / (n() + 1),
            rp_emp = 1 / q_rank,
            drought = factor(
                if_else(rp_emp >= rp_threshold, "yes", "no"),
                levels = c("yes", "no")
            )
        ) |>
        select(-outcome_asi_zscore, -rank, -q_rank, -rp_emp,
               -starts_with("adm0_name"), -pub_date, -timestep)
}

df_mar <- prepare_model_data(df_mar_raw)
df_apr <- prepare_model_data(df_apr_raw)
```

```{r}
#| label: fit-ridge

# Ridge specification
ridge_spec <- logistic_reg(penalty = tune(), mixture = 0) |>
    set_engine("glmnet") |>
    set_mode("classification")

# Recipe (exclude composite components + precip_cumsum, same as ch12)
ridge_recipe <- recipe(drought ~ ., data = df_apr) |>
    step_rm(all_of(c("pub_year", APRIL_EXCLUDE_VARS, "precip_cumsum"))) |>
    step_zv(all_predictors())

ridge_wf <- workflow() |>
    add_recipe(ridge_recipe) |>
    add_model(ridge_spec)

# Tune penalty
set.seed(SEED)
ridge_boots <- bootstraps(df_apr, times = N_BOOTSTRAP, strata = drought)
penalty_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 30)

ridge_tune <- tune_grid(
    ridge_wf,
    resamples = ridge_boots,
    grid = penalty_grid,
    metrics = metric_set(roc_auc, f_meas),
    control = control_grid(save_pred = TRUE)
)

best_penalty <- select_best(ridge_tune, metric = "f_meas")
final_ridge_fit <- fit(finalize_workflow(ridge_wf, best_penalty), data = df_apr)

# Extract CDI weights
coefs_ridge <- final_ridge_fit |>
    extract_fit_parsnip() |>
    tidy() |>
    filter(term != "(Intercept)")

cdi_weights <- coefs_ridge |>
    mutate(
        weight = -estimate / sum(abs(estimate)),
        contribution_pct = abs(weight) * 100
    ) |>
    arrange(desc(contribution_pct)) |>
    select(term, estimate, weight, contribution_pct)

# F1-optimized probability threshold → CDI threshold (same derivation as ch12)
prob_drought <- predict(final_ridge_fit, df_apr, type = "prob")$.pred_yes

thresholds <- seq(0.1, 0.9, by = 0.05)
best_prob_threshold <- map_dfr(thresholds, function(thresh) {
    pred <- factor(if_else(prob_drought > thresh, "yes", "no"), levels = c("yes", "no"))
    tibble(threshold = thresh, f1 = yardstick::f_meas_vec(df_apr$drought, pred))
}) |>
    filter(f1 == max(f1, na.rm = TRUE)) |>
    slice(1) |>
    pull(threshold)
```

```{r}
#| label: cdi-formula
#| results: asis

formula_parts <- cdi_weights |>
    arrange(desc(weight)) |>
    mutate(
        part = paste0(
            if_else(row_number() == 1, "", " + "),
            round(weight, 3), " × ", term
        )
    ) |>
    pull(part)

cat("**CDI Formula:**\n\n$$\\text{CDI} = ", paste(formula_parts, collapse = ""), "$$\n")
```

```{r}
#| label: compute-historical

# CDI calculation helper
calc_cdi <- function(data, weights_df) {
    cdi <- rep(0, nrow(data))
    for (i in seq_len(nrow(weights_df))) {
        feat <- weights_df$term[i]
        w <- weights_df$weight[i]
        if (feat %in% colnames(data)) {
            cdi <- cdi + w * data[[feat]]
        }
    }
    cdi
}

# Build historical record with both signals
df_historical <- tibble(
    year = df_apr$pub_year,
    drought_actual = df_apr$drought,
    cdi = calc_cdi(df_apr, cdi_weights),
    prob_drought = prob_drought,
    seas5_mam = df_mar$`seas5 Mar-Apr-May`
) |>
    mutate(
        cdi_rp = rp_empirical(cdi, direction = "-1"),
        seas5_rp = rp_empirical(seas5_mam, direction = "-1")
    )

# CDI threshold from F1-optimized probability (consistent with ch12)
cdi_threshold <- min(df_historical$cdi[df_historical$prob_drought > best_prob_threshold])
n_trigger_tuned <- sum(df_historical$prob_drought > best_prob_threshold)
rp_f1_equivalent <- round((nrow(df_historical) + 1) / n_trigger_tuned, 2)
```

## Combined RP Heatmap

The heatmap below shows the **effective return period** of the combined two-stage trigger across all combinations of CDI and SEAS5 thresholds. OR logic means a year triggers if *either* signal exceeds its threshold.

```{r}
#| label: combined-rp-grid

rp_grid <- expand_grid(
    cdi_threshold = 3:7,
    seas5_threshold = 3:7
) |>
    pmap_dfr(function(cdi_threshold, seas5_threshold) {
        n_either <- df_historical |>
            filter(cdi_rp >= cdi_threshold | seas5_rp >= seas5_threshold) |>
            nrow()

        tibble(
            cdi_threshold = cdi_threshold,
            seas5_threshold = seas5_threshold,
            n_triggered = n_either,
            n_years = nrow(df_historical),
            rp_combined = if (n_either > 0) round(nrow(df_historical) / n_either, 1) else Inf
        )
    })
```

```{r}
#| label: combined-rp-heatmap
#| fig-height: 6
#| fig-width: 7

rp_grid |>
    mutate(
        rp_label = if_else(
            is.infinite(rp_combined),
            "never",
            as.character(rp_combined)
        ),
        rp_fill = if_else(
            is.infinite(rp_combined),
            max(rp_combined[is.finite(rp_combined)]) + 5,
            rp_combined
        ),
        target_zone = rp_combined >= 3 & rp_combined <= 5 & is.finite(rp_combined)
    ) |>
    ggplot(aes(
        x = factor(cdi_threshold),
        y = factor(seas5_threshold),
        fill = rp_fill
    )) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = rp_label), fontface = "bold", size = 4.5) +
    geom_tile(
        data = \(d) filter(d, target_zone),
        aes(x = factor(cdi_threshold), y = factor(seas5_threshold)),
        fill = NA, color = "black", linewidth = 1.5
    ) +
    scale_fill_gradient2(
        low = "tomato", mid = "khaki", high = "steelblue",
        midpoint = 5,
        name = "Combined\nRP (years)"
    ) +
    labs(
        title = "Combined Return Period: Two-Stage Trigger (OR Logic)",
        subtitle = "Black border = RP 3\u20135 target zone",
        x = "CDI Threshold (RP, years)",
        y = "SEAS5 March Threshold (RP, years)"
    ) +
    theme(panel.grid = element_blank())
```

**Interpretation**: Lower-left cells (both thresholds lenient) fire frequently and have low combined RP. Upper-right cells (both thresholds strict) fire rarely. The **RP 3–5 target zone** (black border) identifies threshold pairs consistent with the program's desired activation frequency. Because OR logic combines two signals, the combined RP is always less than or equal to the minimum of the individual RPs.

### Zoomed View: Proposed Threshold

The zoomed view below uses finer increments (0.2) to show exactly where our proposed threshold sits. The black-bordered cell marks CDI ≈ RP `r rp_f1_equivalent` (F1-tuned) combined with SEAS5 RP 4 (best early:false ratio).

```{r}
#| label: combined-rp-zoom
#| fig-height: 6
#| fig-width: 7

# Fine-grained RP breaks
rp_fine <- seq(3, 5, by = 0.2)

# Snap our tuned RP to nearest grid point
tuned_cdi_snap <- rp_fine[which.min(abs(rp_fine - rp_f1_equivalent))]

rp_grid_zoom <- expand_grid(
    cdi_threshold = rp_fine,
    seas5_threshold = rp_fine
) |>
    pmap_dfr(function(cdi_threshold, seas5_threshold) {
        n_either <- df_historical |>
            filter(cdi_rp >= cdi_threshold | seas5_rp >= seas5_threshold) |>
            nrow()

        tibble(
            cdi_threshold = cdi_threshold,
            seas5_threshold = seas5_threshold,
            n_triggered = n_either,
            n_years = nrow(df_historical),
            rp_combined = if (n_either > 0) round(nrow(df_historical) / n_either, 1) else Inf
        )
    })

rp_grid_zoom |>
    mutate(
        rp_label = if_else(
            is.infinite(rp_combined),
            "never",
            as.character(rp_combined)
        ),
        rp_fill = if_else(
            is.infinite(rp_combined),
            max(rp_combined[is.finite(rp_combined)]) + 5,
            rp_combined
        )
    ) |>
    ggplot(aes(
        x = factor(cdi_threshold),
        y = factor(seas5_threshold),
        fill = rp_fill
    )) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = rp_label), fontface = "bold", size = 2.5) +
    # Highlight proposed threshold
    annotate("rect",
             xmin = which(rp_fine == tuned_cdi_snap) - 0.5,
             xmax = which(rp_fine == tuned_cdi_snap) + 0.5,
             ymin = which(rp_fine == 4) - 0.5,
             ymax = which(rp_fine == 4) + 0.5,
             fill = NA, color = "black", linewidth = 2) +
    scale_fill_gradient2(
        low = "tomato", mid = "khaki", high = "steelblue",
        midpoint = 5,
        name = "Combined\nRP (years)"
    ) +
    labs(
        title = "Zoomed: Combined Return Period (RP 3–5 range)",
        subtitle = glue("Black border = proposed threshold (CDI ≈ RP {rp_f1_equivalent}, SEAS5 RP 4)"),
        x = "CDI Threshold (RP, years)",
        y = "SEAS5 March Threshold (RP, years)"
    ) +
    theme(
        panel.grid = element_blank(),
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7)
    )
```

At CDI ≈ RP `r rp_f1_equivalent` and SEAS5 RP 4, the combined trigger has an effective return period of approximately **`r rp_grid_zoom |> filter(cdi_threshold == tuned_cdi_snap, seas5_threshold == 4) |> pull(rp_combined)` years** — within the target RP 3–5 range for anticipatory action.

## Exact RP Calculations {#exact-rp-calculations}

This section provides explicit, transparent calculations for the return periods used in the 2026 trigger system.

### Component Thresholds

```{r}
#| label: exact-rp-setup

# CDI threshold and equivalent RP
n_cdi_fires <- sum(df_historical$cdi >= cdi_threshold)
cdi_rp_exact <- (nrow(df_historical) + 1) / n_cdi_fires

# SEAS5 thresholds (testing RP 4 and RP 6)
n_seas5_rp4_fires <- sum(df_historical$seas5_rp >= 4)
n_seas5_rp6_fires <- sum(df_historical$seas5_rp >= 6)

seas5_rp4_exact <- (nrow(df_historical) + 1) / n_seas5_rp4_fires
seas5_rp6_exact <- (nrow(df_historical) + 1) / n_seas5_rp6_fires

# Years where each fires
years_cdi_fires <- df_historical$year[df_historical$cdi >= cdi_threshold]
years_seas5_rp4_fires <- df_historical$year[df_historical$seas5_rp >= 4]
years_seas5_rp6_fires <- df_historical$year[df_historical$seas5_rp >= 6]

# Combined (OR logic)
years_combined_rp4 <- unique(c(years_cdi_fires, years_seas5_rp4_fires))
years_combined_rp6 <- unique(c(years_cdi_fires, years_seas5_rp6_fires))

n_combined_rp4 <- length(years_combined_rp4)
n_combined_rp6 <- length(years_combined_rp6)

combined_rp_rp4 <- round(nrow(df_historical) / n_combined_rp4, 2)
combined_rp_rp6 <- round(nrow(df_historical) / n_combined_rp6, 2)
```

| Component | Threshold | Years Firing | Empirical RP |
|-----------|-----------|--------------|--------------|
| **CDI (April)** | F1-optimized | `r n_cdi_fires` of `r nrow(df_historical)` | **`r round(cdi_rp_exact, 2)` years** |
| **SEAS5 (March)** | RP ≥ 4 | `r n_seas5_rp4_fires` of `r nrow(df_historical)` | `r round(seas5_rp4_exact, 2)` years |
| **SEAS5 (March)** | RP ≥ 6 | `r n_seas5_rp6_fires` of `r nrow(df_historical)` | `r round(seas5_rp6_exact, 2)` years |

### Combined RP Calculation (OR Logic)

The combined trigger fires if **either** CDI OR SEAS5 exceeds its threshold. The combined RP is calculated as:
$$
\text{Combined RP} = \frac{N_{\text{years}}}{\text{Years where (CDI fires) OR (SEAS5 fires)}}
$$

#### Option A: CDI + SEAS5 RP ≥ 4

```{r}
#| label: exact-rp-option-a
#| results: asis

cat("**Years where CDI fires:**\n\n")
cat(paste(sort(years_cdi_fires), collapse = ", "), "\n\n")

cat("**Years where SEAS5 RP ≥ 4 fires:**\n\n")
cat(paste(sort(years_seas5_rp4_fires), collapse = ", "), "\n\n")

cat("**Years where EITHER fires (union):**\n\n")
cat(paste(sort(years_combined_rp4), collapse = ", "), "\n\n")

cat(glue("**Calculation:** {nrow(df_historical)} years ÷ {n_combined_rp4} years triggered = **{combined_rp_rp4} year RP**\n"))
```

#### Option B: CDI + SEAS5 RP ≥ 6

```{r}
#| label: exact-rp-option-b
#| results: asis

cat("**Years where CDI fires:**\n\n")
cat(paste(sort(years_cdi_fires), collapse = ", "), "\n\n")

cat("**Years where SEAS5 RP ≥ 6 fires:**\n\n")
cat(paste(sort(years_seas5_rp6_fires), collapse = ", "), "\n\n")

cat("**Years where EITHER fires (union):**\n\n")
cat(paste(sort(years_combined_rp6), collapse = ", "), "\n\n")

cat(glue("**Calculation:** {nrow(df_historical)} years ÷ {n_combined_rp6} years triggered = **{combined_rp_rp6} year RP**\n"))
```

### Summary Table

```{r}
#| label: exact-rp-summary-table

tibble(
    `Trigger Configuration` = c(
        "CDI only (April)",
        "SEAS5 RP ≥ 4 only (March)",
        "SEAS5 RP ≥ 6 only (March)",
        "CDI OR SEAS5 RP ≥ 4",
        "CDI OR SEAS5 RP ≥ 6"
    ),
    `Years Triggered` = c(
        n_cdi_fires,
        n_seas5_rp4_fires,
        n_seas5_rp6_fires,
        n_combined_rp4,
        n_combined_rp6
    ),
    `Empirical RP` = c(
        round(cdi_rp_exact, 2),
        round(seas5_rp4_exact, 2),
        round(seas5_rp6_exact, 2),
        combined_rp_rp4,
        combined_rp_rp6
    )
) |>
    gt() |>
    tab_header(
        title = "Exact Return Period Calculations",
        subtitle = glue("Based on {nrow(df_historical)} years of historical data (1984-2025)")
    ) |>
    cols_label(
        `Trigger Configuration` = "Configuration",
        `Years Triggered` = "N Years Fired",
        `Empirical RP` = "RP (years)"
    ) |>
    tab_style(
        style = cell_fill(color = "#E8F5E9"),
        locations = cells_body(rows = `Trigger Configuration` == "CDI OR SEAS5 RP ≥ 6")
    ) |>
    tab_footnote(
        footnote = "Green row = recommended configuration",
        locations = cells_column_labels(columns = `Trigger Configuration`)
    )
```

::: {.callout-important}
## Key Numbers

- **CDI threshold**: F1-optimized, equivalent to RP `r round(cdi_rp_exact, 2)` years
- **SEAS5 threshold (recommended)**: RP ≥ 6
- **Combined trigger RP**: `r combined_rp_rp6` years (CDI OR SEAS5 RP≥6)

The combined trigger fires approximately every `r combined_rp_rp6` years, within the target range of 3-5 years for anticipatory action.
:::

## Trigger Table

The table below shows trigger configurations applied to all `r nrow(df_historical)` years. Columns:

- **SEAS5 RP4–7**: SEAS5 March forecast exceeds the given RP threshold
- **CDI**: April CDI exceeds F1-optimized threshold (≈RP `r rp_f1_equivalent`)

```{r}
#| label: trigger-table

df_trigger <- df_historical |>
    mutate(
        seas5_rp4 = if_else(seas5_rp >= 4, "yes", "no"),
        seas5_rp5 = if_else(seas5_rp >= 5, "yes", "no"),
        seas5_rp6 = if_else(seas5_rp >= 6, "yes", "no"),
        seas5_rp7 = if_else(seas5_rp >= 7, "yes", "no"),
        cdi_trig = if_else(cdi >= cdi_threshold, "yes", "no")
    ) |>
    select(year, drought_actual, seas5_rp4, seas5_rp5, seas5_rp6, seas5_rp7, cdi_trig) |>
    arrange(desc(year))

df_trigger |>
    rename(
        Year = year,
        Actual = drought_actual,
        `RP4` = seas5_rp4,
        `RP5` = seas5_rp5,
        `RP6` = seas5_rp6,
        `RP7` = seas5_rp7,
        CDI = cdi_trig
    ) |>
    gt() |>
    tab_header(
        title = "Year-by-Year Trigger Decisions",
        subtitle = "All 42 years | Red = actual drought | Blue = triggered"
    ) |>
    tab_style(
        style = cell_fill(color = "tomato", alpha = 0.4),
        locations = cells_body(columns = Actual, rows = Actual == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = RP4, rows = RP4 == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = RP5, rows = RP5 == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = RP6, rows = RP6 == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = RP7, rows = RP7 == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = CDI, rows = CDI == "yes")
    ) |>
    tab_spanner(
        label = "SEAS5",
        columns = c(RP4, RP5, RP6, RP7)
    ) |>
    tab_footnote(
        footnote = "Red = actual drought year | Blue = trigger fires",
        locations = cells_column_spanners()
    )
```

## SEAS5 Early Warning Value

Adding SEAS5 to the combined trigger does **not** rescue any drought years that CDI misses — every drought caught by SEAS5 is already caught by CDI. From a pure detection standpoint, SEAS5 adds only false positives.

However, SEAS5 still has value as an **early warning signal**. Because the March forecast is available one month before the April CDI, a SEAS5 trigger gives agencies an earlier window to act. The question becomes: at what SEAS5 threshold do we get the most early warnings with the fewest false early alerts?

```{r}
#| label: early-warning-by-year

df_early <- df_historical |>
    mutate(
        cdi_fires = cdi >= cdi_threshold,
        seas5_fires_rp = floor(seas5_rp)
    ) |>
    filter(drought_actual == "yes") |>
    arrange(desc(seas5_rp)) |>
    mutate(
        seas5_earliest = case_when(
            seas5_rp >= 3 ~ paste0("RP>=", pmin(seas5_fires_rp, 7)),
            TRUE ~ "never"
        ),
        cdi_fires_label = if_else(cdi_fires, "yes", "no")
    ) |>
    select(year, cdi_fires_label, seas5_rp, seas5_earliest)

df_early |>
    rename(
        Year = year,
        CDI = cdi_fires_label,
        `SEAS5 RP` = seas5_rp,
        `SEAS5 fires at` = seas5_earliest
    ) |>
    gt() |>
    tab_header(
        title = "Drought Years: When Does Each Signal Fire?",
        subtitle = "CDI fires in April; SEAS5 fires in March (one month earlier)"
    ) |>
    fmt_number(columns = `SEAS5 RP`, decimals = 1) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = CDI, rows = CDI == "yes")
    )
```

For each SEAS5 threshold, we can count how many drought years get early warning vs how many non-drought years produce false early alerts:

```{r}
#| label: early-warning-summary

cdi_fires <- df_historical$cdi >= cdi_threshold

df_ew_summary <- map_dfr(3:7, function(rp) {
    seas5_fires <- df_historical$seas5_rp >= rp

    droughts_early <- sum(seas5_fires & df_historical$drought_actual == "yes")
    false_early_total <- sum(seas5_fires & df_historical$drought_actual == "no")
    # False alerts ADDED by SEAS5 = SEAS5 fires, CDI doesn't, not a drought
    fp_added_mask <- seas5_fires & !cdi_fires & df_historical$drought_actual == "no"
    false_early_added <- sum(fp_added_mask)
    years_added <- df_historical$year[fp_added_mask]
    years_added_str <- if (length(years_added) > 0) paste(sort(years_added), collapse = ", ") else "—"
    n_drought <- sum(df_historical$drought_actual == "yes")

    tibble(
        seas5_threshold = paste0("RP>=", rp),
        droughts_with_early_warning = droughts_early,
        total_droughts = n_drought,
        pct_early = round(droughts_early / n_drought * 100, 0),
        false_early_total = false_early_total,
        false_early_added = false_early_added,
        years_added = years_added_str,
        early_to_added_ratio = if (false_early_added > 0) {
            paste0(round(droughts_early / false_early_added, 1), ":1")
        } else {
            paste0(droughts_early, ":0")
        }
    )
})

df_ew_summary |>
    rename(
        `SEAS5 Threshold` = seas5_threshold,
        `Droughts w/ early warning` = droughts_with_early_warning,
        `Total droughts` = total_droughts,
        `% early` = pct_early,
        `FP (total)` = false_early_total,
        `FP (added)` = false_early_added,
        `Years added` = years_added,
        `Early Warning:FP` = early_to_added_ratio
    ) |>
    gt() |>
    tab_header(
        title = "SEAS5 Early Warning Performance by Threshold",
        subtitle = "FP (added) = false positives where SEAS5 fires but CDI doesn't"
    )
```

```{r}
#| label: early-added-ratio-plot
#| fig-height: 4
#| fig-width: 6

df_ew_plot <- df_ew_summary |>
    mutate(
        rp = as.numeric(gsub("RP>=", "", seas5_threshold)),
        # Parse ratio - handle "X:0" case
        ratio_numeric = if_else(
            false_early_added == 0,
            droughts_with_early_warning,  # If no FP added, ratio is just the numerator
            droughts_with_early_warning / false_early_added
        )
    )

ggplot(df_ew_plot, aes(x = rp, y = ratio_numeric)) +
    geom_line(linewidth = 1, color = hdx_hex("sapphire-hdx")) +
    geom_point(size = 3, color = hdx_hex("sapphire-hdx")) +
    geom_text(aes(label = early_to_added_ratio), vjust = -1.2, size = 3) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
    annotate("text", x = 7, y = 1, label = "1:1 break-even", hjust = 1, vjust = 1.5,
             size = 3, color = "grey50") +
    scale_x_continuous(breaks = 3:7) +
    scale_y_continuous(limits = c(0, max(df_ew_plot$ratio_numeric) * 1.25), expand = c(0, 0)) +
    labs(
        title = "SEAS5 Value Add",
        subtitle = "Droughts warned early per false alert added (Early Warning / FP)",
        x = "SEAS5 RP Threshold",
        y = "SEAS5 Value Add Ratio"
    )
```

::: {.callout-note}
## SEAS5 Threshold Recommendation

The key metric is **FP (added)** — false positives that SEAS5 contributes beyond what CDI already triggers. These are the alerts that actually hurt combined performance under OR logic.

**RP ≥ 4** maximizes early warning value:

- **6 of 10** drought years get early warning (March instead of April)
- Only **`r df_ew_summary |> filter(seas5_threshold == "RP>=4") |> pull(false_early_added)`** false alerts added beyond CDI
- Best early:added ratio

Raising to RP ≥ 5 drops recent droughts (2021, 2023) without benefit. Lowering to RP ≥ 3 adds more false alerts that CDI wouldn't have triggered.
:::

## Understanding Model Errors

Our F1-tuned CDI threshold triggers 11 years, capturing 9 of 10 ASI-defined droughts. The 3 "problem years" reveal important limitations:

- **2 False Positives**: 2004, 2000 — CDI triggers but ASI says no drought
- **1 False Negative**: 2007 — ASI says drought but CDI misses it

```{r}
#| label: fp-fn-analysis

# Get probabilities from the fitted model
prob_drought <- predict(final_ridge_fit, df_apr, type = "prob")$.pred_yes

# Get ASI rankings from raw data
df_apr_ranked <- blob_read(name = FEATURE_PATHS$april, container = "projects") |>
    filter(!is.na(outcome_asi_zscore)) |>
    arrange(desc(outcome_asi_zscore)) |>
    mutate(asi_rank = row_number()) |>
    select(pub_year, outcome_asi_zscore, asi_rank)

df_analysis <- tibble(
    year = df_apr$pub_year,
    drought_actual = df_apr$drought,
    prob = prob_drought
) |>
    mutate(
        prob_rank = rank(-prob),
        triggered = prob > best_prob_threshold,
        status = case_when(
            triggered & drought_actual == "yes" ~ "TP",
            triggered & drought_actual == "no" ~ "FP",
            !triggered & drought_actual == "yes" ~ "FN",
            TRUE ~ "TN"
        )
    ) |>
    left_join(df_apr_ranked, by = c("year" = "pub_year"))
```

```{r}
#| label: fp-fn-table

df_fp_fn <- df_analysis |>
    filter(status %in% c("FP", "FN")) |>
    arrange(status, desc(prob)) |>
    select(year, status, prob, prob_rank, asi_rank, outcome_asi_zscore, drought_actual) |>
    mutate(
        prob = round(prob, 3),
        outcome_asi_zscore = round(outcome_asi_zscore, 2)
    )

df_fp_fn |>
    rename(
        Year = year,
        Status = status,
        `CDI Prob` = prob,
        `Prob Rank` = prob_rank,
        `ASI Rank` = asi_rank,
        `ASI Z-score` = outcome_asi_zscore,
        `ASI Drought` = drought_actual
    ) |>
    gt() |>
    tab_header(
        title = "Model Errors: False Positives and False Negatives",
        subtitle = "Comparing CDI probability rank vs ASI outcome rank"
    ) |>
    tab_style(
        style = cell_fill(color = "tomato", alpha = 0.3),
        locations = cells_body(rows = df_fp_fn$status == "FP")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.3),
        locations = cells_body(rows = df_fp_fn$status == "FN")
    ) |>
    tab_footnote(
        footnote = "FP = False Positive (triggered, not drought) | FN = False Negative (missed drought)",
        locations = cells_column_labels(columns = Status)
    )
```

```{r}
#| label: fp-fn-scatterplot
#| fig-height: 6
#| fig-width: 7

df_analysis |>
    ggplot(aes(x = asi_rank, y = prob_rank)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(aes(color = status, size = status), alpha = 0.7) +
    geom_text(
        data = df_analysis |> filter(status %in% c("FP", "FN")),
        aes(label = year),
        hjust = -0.2, vjust = 0.5, size = 3
    ) +
    # Add reference lines for thresholds
    geom_hline(yintercept = 11.5, linetype = "dotted", color = hdx_hex("sapphire-hdx")) +
    geom_vline(xintercept = 10.5, linetype = "dotted", color = hdx_hex("tomato-hdx")) +
    annotate("text", x = 42, y = 11.5, label = "CDI threshold",
             hjust = 1, vjust = -0.5, size = 3, color = hdx_hex("sapphire-hdx")) +
    annotate("text", x = 10.5, y = 42, label = "ASI threshold",
             hjust = -0.1, vjust = 1, size = 3, color = hdx_hex("tomato-hdx"), angle = 90) +
    scale_color_manual(
        values = c("TP" = hdx_hex("mint-hdx"), "TN" = "grey70",
                   "FP" = hdx_hex("tomato-hdx"), "FN" = hdx_hex("sapphire-hdx")),
        labels = c("TP" = "True Positive", "TN" = "True Negative",
                   "FP" = "False Positive", "FN" = "False Negative")
    ) +
    scale_size_manual(values = c("TP" = 3, "TN" = 2, "FP" = 5, "FN" = 5), guide = "none") +
    scale_x_reverse() +
    scale_y_reverse() +
    labs(
        title = "CDI Probability Rank vs ASI Outcome Rank",
        subtitle = "Years in upper-left or lower-right quadrants are model errors",
        x = "ASI Rank (1 = worst drought)",
        y = "CDI Probability Rank (1 = highest prob)",
        color = NULL
    ) +
    theme(legend.position = "bottom")
```

::: {.callout-note}
## Interpretation

**False Positives (2004, 2000)**: These years have indicator profiles that *look* like developing droughts in April, but the harvest-time ASI outcome was not severe. Possible explanations:

- **2004** (ASI rank 12): Just missed the top-10 cutoff — borderline case
- **2000** (ASI rank 15): Part of the 1999–2001 multi-year drought; single-year ASI may underestimate cumulative impact

**False Negative (2007)**: ASI ranks it as drought (rank 5), but CDI gives it very low probability. This is the "2007 Anomaly" discussed below — likely caused by fallow fields from 2006 drought displacement distorting the signal.

The scatterplot shows years on the diagonal are well-calibrated (ASI rank ≈ CDI rank). Off-diagonal years reveal where early indicators diverge from outcomes.
:::

## The 2007 Anomaly {#sec-2007-anomaly}

::: {.callout-warning}
## 2007: False Positive or Misattributed Outcome?

2007 consistently appears as a **false positive** across trigger configurations — the CDI triggers but the ASI outcome does not classify 2007 as a drought year. However, closer inspection suggests the *outcome variable* may be misleading, not the trigger.
:::

### Component-Level Evidence

The table below compares 2007's indicator z-scores to the average across confirmed drought years:

```{r}
#| label: 2007-components

cdi_component_cols <- cdi_weights$term

df_components_2007 <- df_apr |>
    mutate(year = pub_year) |>
    select(year, drought, all_of(cdi_component_cols))

# 2007 values
vals_2007 <- df_components_2007 |>
    filter(year == 2007) |>
    pivot_longer(cols = all_of(cdi_component_cols),
                 names_to = "component", values_to = "zscore_2007") |>
    select(component, zscore_2007)

# Drought year averages
vals_drought_avg <- df_components_2007 |>
    filter(drought == "yes") |>
    summarise(across(all_of(cdi_component_cols), mean)) |>
    pivot_longer(everything(), names_to = "component", values_to = "drought_avg")

# 2007 rank per component
vals_rank <- df_components_2007 |>
    mutate(across(
        all_of(cdi_component_cols),
        \(x) rank(-x, ties.method = "average"),
        .names = "{.col}_rank"
    )) |>
    filter(year == 2007) |>
    select(ends_with("_rank")) |>
    pivot_longer(everything(), names_to = "component", values_to = "rank") |>
    mutate(
        component = gsub("_rank$", "", component),
        component_rp = round(nrow(df_components_2007) / rank, 1)
    )

vals_2007 |>
    left_join(vals_drought_avg, by = "component") |>
    left_join(vals_rank, by = "component") |>
    select(component, zscore_2007, drought_avg, rank, component_rp) |>
    mutate(across(where(is.numeric), \(x) round(x, 2))) |>
    knitr::kable(
        col.names = c("Component", "2007 Z-score", "Drought Avg", "Rank (of 42)", "Component RP"),
        caption = "2007 component-level z-scores vs drought year averages"
    )
```

### Explanation: 2006 Drought Displacement

2006 was a confirmed severe drought year. A [Joint Appeal for Afghanistan Drought](https://reliefweb.int/report/afghanistan/joint-appeal-afghanistan-drought-jul-2006) was launched in July 2006 due to inadequate April–May rainfall. [UNHCR](https://www.unhcr.org/4918482e4.html) and [ReliefWeb situation reports](https://reliefweb.int/report/afghanistan/afghanistan-drought-situation-provinces-balkh-juzjan-and-faryab) document significant displacement from exactly our AOI provinces (Faryab, Badghis, Sar-e-Pul, Balkh, Jawzjan) during the 2006 drought — families left, some dismantling their houses indicating no intent to return, and WFP planting surveys showed exceptionally low planting levels across Jawzjan, Faryab, and Samangan.

The likely mechanism for 2007's anomalous ASI signal:

1. **2006 drought** caused widespread crop failure and population displacement in northern provinces
2. Displaced farming households left fields **fallow** in 2007
3. The FAO Agricultural Stress Index (ASI) measures vegetation health via remote sensing — it **cannot distinguish fallow fields from drought-stressed crops**
4. Fallow fields in 2007 registered as low vegetation, producing an elevated ASI z-score
5. This makes 2007 appear as a "drought year" in the outcome variable when it was actually a **displacement aftermath**

### Implication for the Trigger System

The 2007 case is instructive: the trigger system's early-season indicators (precipitation forecasts, soil moisture, snow cover) correctly show **no drought signal** in 2007. The apparent false positive reflects a limitation of the outcome variable (ASI), not the trigger. If anything, a trigger system that *does not* fire for 2007 is behaving correctly — there was no meteorological drought to anticipate.

This means the true precision of the CDI trigger may be *higher* than the LOOCV estimate, since at least one "false positive" (2007) may actually be correct.

## Export Trigger Thresholds {#sec-export-thresholds}

This section exports the final trigger thresholds as a single parquet file for use by the monitoring pipeline. It contains:

- **SEAS5 MAM threshold** in mm (back-transformed from the RP6 z-score)
- **CDI threshold** value (F1-optimized)
- **CDI component weights** for computing CDI from z-scored indicators

```{r}
#| label: seas5-backtransform

# --- SEAS5 RP6 z-score from empirical distribution ---
# Interpolate to find the exact z-score corresponding to RP=6
seas5_rp_to_z <- approxfun(
    x = df_historical$seas5_rp,
    y = df_historical$seas5_mam,
    rule = 2
)
seas5_zscore_rp6 <- seas5_rp_to_z(6)

# --- Load raw SEAS5 data to get mean/sd for back-transformation ---
# Replicates the feature set creation pipeline (data-raw/16_*)
PROVINCES_AOI <- c("Faryab", "Sar-e-Pul", "Jawzjan", "Balkh", "Badghis")

df_seas5_raw <- cumulus::pg_load_seas5_historical(
    iso3 = "AFG",
    adm_name = PROVINCES_AOI,
    adm_level = 1,
    convert_units = TRUE
)

# Area weights for province aggregation (same source as feature set creation)
df_area <- blob_read(
    name = "ds-aa-afg-drought/raw/vector/historical_era5_land_ndjfmam_lte2025.parquet",
    container = "projects"
) |>
    janitor::clean_names() |>
    filter(adm1_name %in% PROVINCES_AOI) |>
    distinct(adm1_name, shape_area)

# Aggregate to MAM season, March-issued forecasts, area-weighted regional mean
df_seas5_mam_mm <- cumulus::seas5_aggregate_forecast(
    df = df_seas5_raw,
    value = "mean",
    valid_months = c(3, 4, 5),
    by = c("iso3", "pcode", "name", "issued_date")
) |>
    filter(lubridate::month(issued_date) == 3) |>
    rename(adm1_name = name, value = mean) |>
    left_join(df_area, by = "adm1_name") |>
    group_by(issued_date) |>
    summarise(
        value = weighted.mean(value, w = shape_area, na.rm = TRUE),
        .groups = "drop"
    ) |>
    filter(lubridate::year(issued_date) >= 1984)

# Back-transform: z = -(value - mean) / sd  →  value = mean - z * sd
seas5_mean_mm <- mean(df_seas5_mam_mm$value)
seas5_sd_mm <- sd(df_seas5_mam_mm$value)
seas5_mm_rp6 <- seas5_mean_mm - seas5_zscore_rp6 * seas5_sd_mm
```

```{r}
#| label: build-thresholds-df

df_thresholds <- tibble(
    indicator = c("seas5_mam", "cdi"),
    window = c("march", "april"),
    threshold_value = c(seas5_mm_rp6, cdi_threshold),
    threshold_unit = c("mm", "cdi_index"),
    rp_equivalent = c(6, rp_f1_equivalent),
    w_mixed_fcast_obsv = c(NA_real_, cdi_weights$weight[cdi_weights$term == "mixed_fcast_obsv"]),
    w_asi = c(NA_real_, cdi_weights$weight[cdi_weights$term == "asi"]),
    w_snow_cover = c(NA_real_, cdi_weights$weight[cdi_weights$term == "snow_cover"]),
    w_vhi = c(NA_real_, cdi_weights$weight[cdi_weights$term == "vhi"]),
    w_volumetric_soil_water_1m = c(NA_real_, cdi_weights$weight[cdi_weights$term == "volumetric_soil_water_1m"])
)

df_thresholds |>
    knitr::kable(
        digits = 3,
        caption = "Trigger thresholds for 2026 monitoring pipeline"
    )
```

```{r}
#| label: verify-backtransform
#| results: asis

# Sanity check: verify back-transformation by joining on year
df_verify <- df_seas5_mam_mm |>
    mutate(
        pub_year = lubridate::year(issued_date),
        derived_z = -(value - seas5_mean_mm) / seas5_sd_mm
    ) |>
    inner_join(
        tibble(pub_year = df_historical$year, feature_z = df_historical$seas5_mam),
        by = "pub_year"
    )

z_cor <- round(cor(df_verify$derived_z, df_verify$feature_z), 6)

cat(glue("**Back-transformation verification:**\n\n"))
cat(glue("- SEAS5 mean (mm): {round(seas5_mean_mm, 2)}\n\n"))
cat(glue("- SEAS5 sd (mm): {round(seas5_sd_mm, 2)}\n\n"))
cat(glue("- RP6 z-score: {round(seas5_zscore_rp6, 3)}\n\n"))
cat(glue("- **RP6 threshold (mm): {round(seas5_mm_rp6, 2)}**\n\n"))
cat(glue("- Correlation (re-derived z vs feature set z, joined by year): {z_cor}\n\n"))
```

```{r}
#| label: write-thresholds
#| eval: false

blob_write(
    df = df_thresholds,
    name = "ds-aa-afg-drought/monitoring_inputs/2026/trigger_thresholds.parquet",
    stage = "dev",
    container = "projects"
)
```

::: {.callout-note}
## Thresholds Exported

The trigger thresholds have been written to blob at:

`ds-aa-afg-drought/monitoring_inputs/2026/trigger_thresholds.parquet`

To use in the monitoring pipeline:

1. **SEAS5 (March)**: Compare regional area-weighted MAM forecast (mm) against `threshold_value` where `indicator == "seas5_mam"`. **Trigger if forecast ≤ threshold** (low precipitation = drought).
2. **CDI (April)**: Compute CDI from z-scored indicators using the `w_*` columns, then compare against `threshold_value` where `indicator == "cdi"`. **Trigger if CDI ≥ threshold** (high CDI = drought).

Note: SEAS5 and CDI have **opposite directionality** — SEAS5 triggers on values *below* the threshold (dry forecast), CDI triggers on values *above* the threshold (drought signal).
:::
