---
title: "CERF Allocation Alignment"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  warning: false
  message: false
project:
  execute-dir: project
---

## Introduction

This chapter evaluates how well the proposed trigger signals align with **historical CERF Rapid Response drought allocations** for Afghanistan. CERF Rapid Response allocations represent an independent, external validation of drought severity — they reflect the humanitarian community's real-time assessment that conditions warranted emergency funding.

We filter to **Rapid Response** allocations only (excluding Underfunded Emergencies), as these represent the acute drought response most analogous to anticipatory action triggers. We treat CERF Rapid Response allocation as the outcome variable and ask: if we had applied our CDI or SEAS5 triggers historically, how often would they have correctly predicted a CERF allocation year?

```{r}
#| label: setup

box::use(
    dplyr[...],
    tidyr[...],
    ggplot2[...],
    gghdx[...],
    cumulus[...],
    purrr[...],
    glue[glue],
    tibble[tibble],
    janitor[clean_names],
    lubridate[year, as_date],
    .. / R / utils[rp_empirical]
)

library(tidymodels)
library(gt)
gghdx()
```

```{r}
#| label: config

RP_THRESHOLD <- 4
SEED <- 42
N_BOOTSTRAP <- 30

FEATURE_PATHS <- list(
    march = "ds-aa-afg-drought/processed/vector/2026_mar_pub_feature_set_v1.parquet",
    april = "ds-aa-afg-drought/processed/vector/2026_april_pub_feature_set_v1.parquet"
)

APRIL_EXCLUDE_VARS <- c("total_precipitation_sum", "seas5 Apr", "seas5 May")
```

## Data

```{r}
#| label: load-data

# Load feature sets
df_mar_raw <- blob_read(name = FEATURE_PATHS$march, container = "projects")
df_apr_raw <- blob_read(name = FEATURE_PATHS$april, container = "projects")

# Prepare modeling data
prepare_model_data <- function(df, rp_threshold = RP_THRESHOLD) {
    df |>
        filter(!is.na(outcome_asi_zscore)) |>
        arrange(desc(outcome_asi_zscore)) |>
        mutate(
            rank = row_number(),
            q_rank = rank / (n() + 1),
            rp_emp = 1 / q_rank,
            drought = factor(
                if_else(rp_emp >= rp_threshold, "yes", "no"),
                levels = c("yes", "no")
            )
        ) |>
        select(-outcome_asi_zscore, -rank, -q_rank, -rp_emp,
               -starts_with("adm0_name"), -pub_date, -timestep)
}

df_mar <- prepare_model_data(df_mar_raw)
df_apr <- prepare_model_data(df_apr_raw)
```

```{r}
#| label: fit-ridge

ridge_spec <- logistic_reg(penalty = tune(), mixture = 0) |>
    set_engine("glmnet") |>
    set_mode("classification")

ridge_recipe <- recipe(drought ~ ., data = df_apr) |>
    step_rm(all_of(c("pub_year", APRIL_EXCLUDE_VARS, "precip_cumsum"))) |>
    step_zv(all_predictors())

ridge_wf <- workflow() |>
    add_recipe(ridge_recipe) |>
    add_model(ridge_spec)

set.seed(SEED)
ridge_boots <- bootstraps(df_apr, times = N_BOOTSTRAP, strata = drought)
penalty_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 30)

ridge_tune <- tune_grid(
    ridge_wf,
    resamples = ridge_boots,
    grid = penalty_grid,
    metrics = metric_set(roc_auc, f_meas),
    control = control_grid(save_pred = TRUE)
)

best_penalty <- select_best(ridge_tune, metric = "f_meas")
final_ridge_fit <- fit(finalize_workflow(ridge_wf, best_penalty), data = df_apr)

coefs_ridge <- final_ridge_fit |>
    extract_fit_parsnip() |>
    tidy() |>
    filter(term != "(Intercept)")

cdi_weights <- coefs_ridge |>
    mutate(
        weight = -estimate / sum(abs(estimate)),
        contribution_pct = abs(weight) * 100
    ) |>
    arrange(desc(contribution_pct)) |>
    select(term, estimate, weight, contribution_pct)

# F1-optimized probability threshold → CDI threshold (same derivation as ch12)
prob_drought <- predict(final_ridge_fit, df_apr, type = "prob")$.pred_yes

thresholds <- seq(0.1, 0.9, by = 0.05)
best_prob_threshold <- map_dfr(thresholds, function(thresh) {
    pred <- factor(if_else(prob_drought > thresh, "yes", "no"), levels = c("yes", "no"))
    tibble(threshold = thresh, f1 = yardstick::f_meas_vec(df_apr$drought, pred))
}) |>
    filter(f1 == max(f1, na.rm = TRUE)) |>
    slice(1) |>
    pull(threshold)
```

```{r}
#| label: load-cerf

df_cerf <- blob_read("ds-aa-afg-drought/raw/vector/afg_drought_cerf_allocations.xlsx") |>
    clean_names() |>
    mutate(allocation_date = as_date(allocation_date)) |>
    filter(window == "Rapid Response")

df_cerf_yr <- df_cerf |>
    group_by(yr = year(allocation_date)) |>
    summarise(n_allocations = n()) |>
    # 2025 CERF Rapid Response allocation occurred but is not yet in the dataset
    bind_rows(tibble(yr = 2025, n_allocations = 1)) |>
    complete(yr = 2006:2025, fill = list(n_allocations = 0)) |>
    mutate(cerf = factor(if_else(n_allocations > 0, "yes", "no"), levels = c("yes", "no")))
```

```{r}
#| label: build-historical

calc_cdi <- function(data, weights_df) {
    cdi <- rep(0, nrow(data))
    for (i in seq_len(nrow(weights_df))) {
        feat <- weights_df$term[i]
        w <- weights_df$weight[i]
        if (feat %in% colnames(data)) {
            cdi <- cdi + w * data[[feat]]
        }
    }
    cdi
}

df_historical <- tibble(
    year = df_apr$pub_year,
    drought_actual = df_apr$drought,
    cdi = calc_cdi(df_apr, cdi_weights),
    prob_drought = prob_drought,
    seas5_mam = df_mar$`seas5 Mar-Apr-May`
) |>
    mutate(
        cdi_rp = rp_empirical(cdi, direction = "-1"),
        seas5_rp = rp_empirical(seas5_mam, direction = "-1")
    )

# CDI threshold from F1-optimized probability (consistent with ch12)
cdi_threshold <- min(df_historical$cdi[df_historical$prob_drought > best_prob_threshold])

# Join with CERF data — restrict to CERF coverage period (2006–2024)
df_cerf_analysis <- df_historical |>
    inner_join(df_cerf_yr, by = c("year" = "yr"))
```

**CERF allocation years**: `r paste(df_cerf_yr |> filter(cerf == "yes") |> pull(yr), collapse = ", ")`.

Analysis period: `r min(df_cerf_analysis$year)`–`r max(df_cerf_analysis$year)` (`r nrow(df_cerf_analysis)` years), of which `r sum(df_cerf_analysis$cerf == "yes")` had CERF drought allocations.

## ROC-AUC: Continuous Signal Quality

Before examining thresholds, we evaluate how well the **continuous** CDI and SEAS5 scores discriminate CERF allocation years from non-allocation years. ROC-AUC is threshold-independent and measures overall ranking quality.

```{r}
#| label: roc-auc

roc_cdi <- yardstick::roc_auc_vec(
    df_cerf_analysis$cerf,
    df_cerf_analysis$cdi
)

roc_seas5 <- yardstick::roc_auc_vec(
    df_cerf_analysis$cerf,
    df_cerf_analysis$seas5_mam
)

tibble(
    Signal = c("CDI (April)", "SEAS5 MAM (March)"),
    `ROC-AUC` = c(roc_cdi, roc_seas5)
) |>
    gt() |>
    tab_header(
        title = "ROC-AUC: Continuous Signal vs CERF Allocation",
        subtitle = glue("{min(df_cerf_analysis$year)}\u2013{max(df_cerf_analysis$year)}")
    ) |>
    fmt_number(columns = `ROC-AUC`, decimals = 3)
```

```{r}
#| label: roc-curves
#| fig-height: 5
#| fig-width: 7

roc_data_cdi <- yardstick::roc_curve(
    df_cerf_analysis |> mutate(cerf = factor(cerf, levels = c("yes", "no"))),
    cerf,
    cdi
) |> mutate(signal = "CDI (April)")

roc_data_seas5 <- yardstick::roc_curve(
    df_cerf_analysis |> mutate(cerf = factor(cerf, levels = c("yes", "no"))),
    cerf,
    seas5_mam
) |> mutate(signal = "SEAS5 MAM (March)")

bind_rows(roc_data_cdi, roc_data_seas5) |>
    ggplot(aes(x = 1 - specificity, y = sensitivity, color = signal)) +
    geom_path(linewidth = 1.2) +
    geom_abline(linetype = "dashed", color = "grey50") +
    scale_color_manual(values = c(
        "CDI (April)" = hdx_hex("sapphire-hdx"),
        "SEAS5 MAM (March)" = hdx_hex("tomato-hdx")
    )) +
    labs(
        title = "ROC Curves: Trigger Signals vs CERF Allocation",
        subtitle = glue("CDI AUC = {round(roc_cdi, 3)} | SEAS5 AUC = {round(roc_seas5, 3)}"),
        x = "False Positive Rate (1 - Specificity)",
        y = "True Positive Rate (Sensitivity)",
        color = NULL
    )
```

## F1 by RP Threshold

For each RP threshold, we flag years where CDI or SEAS5 exceeds the threshold and compute classification metrics against CERF allocation.

```{r}
#| label: f1-by-threshold

df_f1_by_rp <- map_dfr(seq(1, 7, by = 0.1), function(rp) {
    cdi_flag <- factor(
        if_else(df_cerf_analysis$cdi_rp >= rp, "yes", "no"),
        levels = c("yes", "no")
    )
    seas5_flag <- factor(
        if_else(df_cerf_analysis$seas5_rp >= rp, "yes", "no"),
        levels = c("yes", "no")
    )
    combined_flag <- factor(
        if_else(df_cerf_analysis$cdi_rp >= rp | df_cerf_analysis$seas5_rp >= rp, "yes", "no"),
        levels = c("yes", "no")
    )

    cerf_truth <- df_cerf_analysis$cerf

    calc_metrics <- function(pred, label) {
        tibble(
            signal = label,
            rp = rp,
            f1 = yardstick::f_meas_vec(cerf_truth, pred),
            precision = yardstick::precision_vec(cerf_truth, pred),
            recall = yardstick::recall_vec(cerf_truth, pred)
        )
    }

    bind_rows(
        calc_metrics(cdi_flag, "CDI (April)"),
        calc_metrics(seas5_flag, "SEAS5 (March)"),
        calc_metrics(combined_flag, "Combined OR")
    )
})
```

```{r}
#| label: f1-by-threshold-plot
#| fig-height: 5
#| fig-width: 8

df_f1_by_rp |>
    ggplot(aes(x = rp, y = f1, color = signal)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1) +
    scale_x_continuous(breaks = 1:7) +
    scale_color_manual(values = c(
        "CDI (April)" = hdx_hex("sapphire-hdx"),
        "SEAS5 (March)" = hdx_hex("tomato-hdx"),
        "Combined OR" = hdx_hex("mint-hdx")
    )) +
    labs(
        title = "How Well Do Different RP Thresholds Predict CERF Allocation?",
        subtitle = glue("Merged 5-province area | {min(df_cerf_analysis$year)}\u2013{max(df_cerf_analysis$year)}"),
        x = "RP Threshold",
        y = "F1 Score",
        color = NULL
    )
```

```{r}
#| label: f1-threshold-table

df_f1_by_rp |>
    filter(rp %in% 2:6) |>
    mutate(across(c(f1, precision, recall), \(x) round(x, 3))) |>
    pivot_wider(
        names_from = signal,
        values_from = c(f1, precision, recall),
        names_glue = "{signal}_{.value}"
    ) |>
    select(rp, starts_with("CDI"), starts_with("SEAS5"), starts_with("Combined")) |>
    gt() |>
    tab_header(
        title = "Classification Metrics by RP Threshold",
        subtitle = "CERF allocation as outcome"
    ) |>
    tab_spanner(label = "CDI (April)", columns = starts_with("CDI")) |>
    tab_spanner(label = "SEAS5 (March)", columns = starts_with("SEAS5")) |>
    tab_spanner(label = "Combined OR", columns = starts_with("Combined")) |>
    cols_label(
        rp = "RP",
        `CDI (April)_f1` = "F1",
        `CDI (April)_precision` = "Prec",
        `CDI (April)_recall` = "Rec",
        `SEAS5 (March)_f1` = "F1",
        `SEAS5 (March)_precision` = "Prec",
        `SEAS5 (March)_recall` = "Rec",
        `Combined OR_f1` = "F1",
        `Combined OR_precision` = "Prec",
        `Combined OR_recall` = "Rec"
    )
```

## Asymmetric RP Threshold Heatmap

The plot above shows performance when both signals use the same RP threshold. But CDI and SEAS5 may have different optimal thresholds. The heatmap below explores all combinations, with marginal bar plots showing each signal's standalone performance.

```{r}
#| label: asymmetric-rp-grid

# Compute F1 for all CDI × SEAS5 RP combinations
df_asymmetric <- expand_grid(
    cdi_rp_thresh = 2:7,
    seas5_rp_thresh = 2:7
) |>
    pmap_dfr(function(cdi_rp_thresh, seas5_rp_thresh) {
        combined_flag <- factor(
            if_else(
                df_cerf_analysis$cdi_rp >= cdi_rp_thresh |
                df_cerf_analysis$seas5_rp >= seas5_rp_thresh,
                "yes", "no"
            ),
            levels = c("yes", "no")
        )

        f1 <- yardstick::f_meas_vec(df_cerf_analysis$cerf, combined_flag)

        tibble(
            cdi_rp_thresh = cdi_rp_thresh,
            seas5_rp_thresh = seas5_rp_thresh,
            f1 = if_else(is.na(f1), 0, f1)
        )
    })

# Marginal F1 for CDI alone
df_cdi_marginal <- map_dfr(2:7, function(rp) {
    flag <- factor(
        if_else(df_cerf_analysis$cdi_rp >= rp, "yes", "no"),
        levels = c("yes", "no")
    )
    f1 <- yardstick::f_meas_vec(df_cerf_analysis$cerf, flag)
    tibble(rp = rp, f1 = if_else(is.na(f1), 0, f1))
})

# Marginal F1 for SEAS5 alone
df_seas5_marginal <- map_dfr(2:7, function(rp) {
    flag <- factor(
        if_else(df_cerf_analysis$seas5_rp >= rp, "yes", "no"),
        levels = c("yes", "no")
    )
    f1 <- yardstick::f_meas_vec(df_cerf_analysis$cerf, flag)
    tibble(rp = rp, f1 = if_else(is.na(f1), 0, f1))
})
```

```{r}
#| label: asymmetric-heatmap-plot
#| fig-height: 7
#| fig-width: 8

library(patchwork)

# Main heatmap
p_heat <- df_asymmetric |>
    ggplot(aes(x = factor(cdi_rp_thresh), y = factor(seas5_rp_thresh), fill = f1)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = round(f1, 2)), size = 3.5, fontface = "bold") +
    scale_fill_gradient2(
        low = "white", mid = "khaki", high = hdx_hex("mint-hdx"),
        midpoint = 0.4,
        limits = c(0, max(df_asymmetric$f1)),
        name = "F1"
    ) +
    labs(x = "CDI RP Threshold", y = "SEAS5 RP Threshold") +
    theme(
        panel.grid = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 5, 5)
    )

# Top marginal bar (CDI)
p_top <- df_cdi_marginal |>
    ggplot(aes(x = factor(rp), y = f1)) +
    geom_col(fill = hdx_hex("sapphire-hdx"), width = 0.7) +
    geom_text(aes(label = round(f1, 2)), vjust = -0.3, size = 3) +
    scale_y_continuous(limits = c(0, max(df_cdi_marginal$f1) * 1.15), expand = c(0, 0)) +
    labs(x = NULL, y = NULL, title = "CDI alone") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        plot.margin = margin(0, 0, -10, 0)
    )

# Right marginal bar (SEAS5)
p_right <- df_seas5_marginal |>
    ggplot(aes(x = f1, y = factor(rp))) +
    geom_col(fill = hdx_hex("tomato-hdx"), width = 0.7) +
    geom_text(aes(label = round(f1, 2)), hjust = -0.2, size = 3) +
    scale_x_continuous(limits = c(0, max(df_seas5_marginal$f1) * 1.15), expand = c(0, 0)) +
    labs(x = NULL, y = NULL, title = "SEAS5 alone") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        plot.margin = margin(0, 0, 0, -10)
    )

# Empty corner
p_empty <- ggplot() + theme_void()

# Combine with patchwork
(p_top + p_empty + p_heat + p_right) +
    plot_layout(
        widths = c(4, 1),
        heights = c(1, 4)
    ) +
    plot_annotation(
        title = "Combined Trigger F1 (OR Logic) by RP Threshold",
        subtitle = "Heatmap: combined | Top: CDI alone | Right: SEAS5 alone"
    )
```

### Zoomed View: RP 3–5 Range

The overview heatmap uses integer RP thresholds. The zoomed view below uses 0.1 increments to show exactly where our F1-tuned threshold (~RP 3.9) sits within the optimal range.

```{r}
#| label: zoomed-heatmap
#| fig-height: 7
#| fig-width: 8

# Fine-grained RP breaks for zoom
rp_fine <- seq(3, 5, by = 0.2)

# Compute our tuned threshold's equivalent RP
n_trigger_tuned <- sum(df_historical$prob_drought > best_prob_threshold)
tuned_rp <- round((nrow(df_historical) + 1) / n_trigger_tuned, 1)

df_zoom <- expand_grid(
    cdi_rp_thresh = rp_fine,
    seas5_rp_thresh = rp_fine
) |>
    pmap_dfr(function(cdi_rp_thresh, seas5_rp_thresh) {
        combined_flag <- factor(
            if_else(
                df_cerf_analysis$cdi_rp >= cdi_rp_thresh |
                df_cerf_analysis$seas5_rp >= seas5_rp_thresh,
                "yes", "no"
            ),
            levels = c("yes", "no")
        )
        f1 <- yardstick::f_meas_vec(df_cerf_analysis$cerf, combined_flag)
        tibble(
            cdi_rp_thresh = cdi_rp_thresh,
            seas5_rp_thresh = seas5_rp_thresh,
            f1 = if_else(is.na(f1), 0, f1)
        )
    })

# Marginal F1 for CDI alone (zoomed range)
df_cdi_zoom <- map_dfr(rp_fine, function(rp) {
    flag <- factor(
        if_else(df_cerf_analysis$cdi_rp >= rp, "yes", "no"),
        levels = c("yes", "no")
    )
    f1 <- yardstick::f_meas_vec(df_cerf_analysis$cerf, flag)
    tibble(rp = rp, f1 = if_else(is.na(f1), 0, f1))
})

# Marginal F1 for SEAS5 alone (zoomed range)
df_seas5_zoom <- map_dfr(rp_fine, function(rp) {
    flag <- factor(
        if_else(df_cerf_analysis$seas5_rp >= rp, "yes", "no"),
        levels = c("yes", "no")
    )
    f1 <- yardstick::f_meas_vec(df_cerf_analysis$cerf, flag)
    tibble(rp = rp, f1 = if_else(is.na(f1), 0, f1))
})

# Find position of our tuned threshold (snap to nearest grid point)
tuned_cdi_snap <- rp_fine[which.min(abs(rp_fine - tuned_rp))]

# Main heatmap
p_heat_zoom <- df_zoom |>
    ggplot(aes(x = factor(cdi_rp_thresh), y = factor(seas5_rp_thresh), fill = f1)) +
    geom_tile(color = "white", linewidth = 0.3) +
    geom_text(aes(label = round(f1, 2)), size = 2.2) +
    # Highlight our tuned threshold
    annotate("rect",
             xmin = which(rp_fine == tuned_cdi_snap) - 0.5,
             xmax = which(rp_fine == tuned_cdi_snap) + 0.5,
             ymin = which(rp_fine == 4) - 0.5,
             ymax = which(rp_fine == 4) + 0.5,
             fill = NA, color = "black", linewidth = 1.5) +
    scale_fill_gradient2(
        low = "white", mid = "khaki", high = hdx_hex("mint-hdx"),
        midpoint = 0.4,
        limits = c(0, max(df_zoom$f1)),
        name = "F1"
    ) +
    labs(x = "CDI RP Threshold", y = "SEAS5 RP Threshold") +
    theme(
        panel.grid = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 5, 5),
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7)
    )

# Top marginal bar (CDI)
p_top_zoom <- df_cdi_zoom |>
    ggplot(aes(x = factor(rp), y = f1)) +
    geom_col(fill = hdx_hex("sapphire-hdx"), width = 0.7) +
    geom_text(aes(label = round(f1, 2)), vjust = -0.3, size = 2) +
    scale_y_continuous(limits = c(0, max(df_cdi_zoom$f1) * 1.15), expand = c(0, 0)) +
    labs(x = NULL, y = NULL, title = "CDI alone") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        plot.margin = margin(0, 0, -10, 0)
    )

# Right marginal bar (SEAS5)
p_right_zoom <- df_seas5_zoom |>
    ggplot(aes(x = f1, y = factor(rp))) +
    geom_col(fill = hdx_hex("tomato-hdx"), width = 0.7) +
    geom_text(aes(label = round(f1, 2)), hjust = -0.2, size = 2) +
    scale_x_continuous(limits = c(0, max(df_seas5_zoom$f1) * 1.15), expand = c(0, 0)) +
    labs(x = NULL, y = NULL, title = "SEAS5 alone") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        plot.margin = margin(0, 0, 0, -10)
    )

# Empty corner
p_empty_zoom <- ggplot() + theme_void()

# Combine with patchwork
(p_top_zoom + p_empty_zoom + p_heat_zoom + p_right_zoom) +
    plot_layout(
        widths = c(4, 1),
        heights = c(1, 4)
    ) +
    plot_annotation(
        title = "Zoomed: Combined Trigger F1 (RP 3–5 range)",
        subtitle = glue("Black border = proposed threshold (CDI ≈ RP {tuned_rp}, SEAS5 RP 4)")
    )
```

::: {.callout-note}
## Interpretation

CERF alignment shows a **plateau** in the RP 3–5 range — lowering the threshold doesn't improve performance (the same years trigger within the CERF period), and raising it only marginally affects F1 until you start missing CERF years at RP ≥ 5. This means CERF performance alone doesn't determine the optimal threshold.

The **early:false ratio** (see previous chapter) breaks the tie: within the CERF-optimal plateau, SEAS5 RP ≥ 4 maximizes early warning value (6/10 droughts get March warning) while minimizing false early alerts (4 over 42 years, ratio 1.5:1). Lowering to RP 3 adds false alerts without improving CERF alignment; raising to RP 5 loses recent droughts (2021, 2023) without removing false positives.

**Bottom line**: RP ~4 sits in the sweet spot — optimal for CERF alignment AND best early:false ratio for anticipatory action.
:::

## Year-by-Year Alignment

```{r}
#| label: cerf-alignment-table

df_cerf_analysis |>
    mutate(
        cdi_trig = if_else(cdi >= cdi_threshold, "yes", "no"),
        seas5_rp4 = if_else(seas5_rp >= 4, "yes", "no"),
        combined = if_else(cdi >= cdi_threshold | seas5_rp >= 4, "yes", "no")
    ) |>
    select(year, cerf, drought_actual, cdi_trig, seas5_rp4, combined) |>
    arrange(desc(year)) |>
    rename(
        Year = year,
        CERF = cerf,
        `ASI Drought` = drought_actual,
        CDI = cdi_trig,
        `SEAS5 RP4` = seas5_rp4,
        `Combined` = combined
    ) |>
    gt() |>
    tab_header(
        title = "Year-by-Year: Trigger Decisions vs CERF Allocations",
        subtitle = glue("{min(df_cerf_analysis$year)}\u2013{max(df_cerf_analysis$year)}")
    ) |>
    tab_style(
        style = cell_fill(color = "tomato", alpha = 0.4),
        locations = cells_body(columns = CERF, rows = CERF == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "gold", alpha = 0.3),
        locations = cells_body(columns = `ASI Drought`, rows = `ASI Drought` == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = CDI, rows = CDI == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = `SEAS5 RP4`, rows = `SEAS5 RP4` == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "steelblue", alpha = 0.4),
        locations = cells_body(columns = Combined, rows = Combined == "yes")
    ) |>
    tab_spanner(label = "Trigger", columns = c(CDI, `SEAS5 RP4`, `Combined`)) |>
    tab_footnote(
        footnote = "Red = CERF allocation | Gold = ASI drought | Blue = trigger fires",
        locations = cells_column_spanners()
    )
```

::: {.callout-note}
## Interpretation

CERF Rapid Response allocations and ASI-defined droughts overlap substantially but are not identical — CERF reflects humanitarian decision-making which incorporates factors beyond meteorological drought (e.g., conflict, food prices, displacement). Years where the trigger aligns with CERF but not ASI (or vice versa) highlight where the two framings diverge. Underfunded Emergencies allocations (2008, 2010) are excluded as they represent a different funding mechanism with different triggering criteria.
:::
