---
title: "Sensitivity Analysis: SEAS5 Climatology Baseline"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  warning: false
  message: false
project:
  execute-dir: project
---

## Introduction

ECMWF uses 1993-2016 as their climatology reference period for SEAS5, citing pre-1990 satellite and observation network quality issues. This raises the question: **does our SEAS5 early warning component perform differently when we exclude potentially unreliable pre-1991 forecasts from the baseline?**

This analysis keeps the CDI unchanged (1984-2025 baseline) and tests only the SEAS5 early warning component under two baseline periods:

| Component | Baseline | Rationale |
|-----------|----------|-----------|
| CDI (April trigger) | 1984-2025 | ASI/ERA5 observations are reliable; maximize sample |
| SEAS5 (March early warning) | Compare 1984-2025 vs 1991-2025 | Test sensitivity to pre-1991 forecast quality |

```{r}
#| label: setup

box::use(
    dplyr[...],
    tidyr[...],
    ggplot2[...],
    gghdx[...],
    cumulus[...],
    purrr[...],
    glue[glue],
    tibble[tibble],
    lubridate[...],
    janitor[clean_names],
    stringr[str_detect],
    .. / R / utils[rp_empirical]
)

library(tidymodels)
library(gt)
library(patchwork)
gghdx()
```

```{r}
#| label: config

# Return period threshold for defining drought
RP_THRESHOLD <- 4

# Bootstrap resamples for cross-validation
N_BOOTSTRAP <- 30

# Random seed
SEED <- 42

# Provinces of interest
PROVINCES_AOI <- c("Faryab", "Sar-e-Pul", "Jawzjan", "Balkh", "Badghis")

# Feature set paths (uses existing 1984 baseline)
FEATURE_PATHS <- list(
    march = "ds-aa-afg-drought/processed/vector/2026_mar_pub_feature_set_v1.parquet",
    april = "ds-aa-afg-drought/processed/vector/2026_april_pub_feature_set_v1.parquet"
)

# Variables to exclude from CDI model
APRIL_EXCLUDE_VARS <- c("total_precipitation_sum", "seas5 Apr", "seas5 May")

# SEAS5 baseline periods to compare
SEAS5_BASELINES <- list(
    "1984-2025" = 1984,
    "1991-2025" = 1991
)
```

## Load Data & Fit CDI Model

We use the existing feature sets (1984-2025 baseline) for CDI construction, identical to the main analysis.

```{r}
#| label: load-data

# Load feature sets
df_mar_raw <- blob_read(name = FEATURE_PATHS$march, container = "projects")
df_apr_raw <- blob_read(name = FEATURE_PATHS$april, container = "projects")

# Prepare modeling data with empirical RP thresholding
prepare_model_data <- function(df, rp_threshold = RP_THRESHOLD) {
    df |>
        filter(!is.na(outcome_asi_zscore)) |>
        arrange(desc(outcome_asi_zscore)) |>
        mutate(
            rank = row_number(),
            q_rank = rank / (n() + 1),
            rp_emp = 1 / q_rank,
            drought = factor(
                if_else(rp_emp >= rp_threshold, "yes", "no"),
                levels = c("yes", "no")
            )
        ) |>
        select(-outcome_asi_zscore, -rank, -q_rank, -rp_emp,
               -starts_with("adm0_name"), -pub_date, -timestep)
}

df_mar <- prepare_model_data(df_mar_raw)
df_apr <- prepare_model_data(df_apr_raw)
```

```{r}
#| label: fit-cdi-model

# Fit ridge model (identical to main analysis)
ridge_spec <- logistic_reg(penalty = tune(), mixture = 0) |>
    set_engine("glmnet") |>
    set_mode("classification")

ridge_recipe <- recipe(drought ~ ., data = df_apr) |>
    step_rm(all_of(c("pub_year", APRIL_EXCLUDE_VARS, "precip_cumsum"))) |>
    step_zv(all_predictors())

ridge_wf <- workflow() |>
    add_recipe(ridge_recipe) |>
    add_model(ridge_spec)

set.seed(SEED)
ridge_boots <- bootstraps(df_apr, times = N_BOOTSTRAP, strata = drought)
penalty_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 30)

ridge_tune <- tune_grid(
    ridge_wf,
    resamples = ridge_boots,
    grid = penalty_grid,
    metrics = metric_set(roc_auc, f_meas),
    control = control_grid(save_pred = TRUE)
)

best_penalty <- select_best(ridge_tune, metric = "f_meas")
final_ridge_fit <- fit(finalize_workflow(ridge_wf, best_penalty), data = df_apr)

# Extract CDI weights
coefs_ridge <- final_ridge_fit |>
    extract_fit_parsnip() |>
    tidy() |>
    filter(term != "(Intercept)")

cdi_weights <- coefs_ridge |>
    mutate(
        weight = -estimate / sum(abs(estimate)),
        contribution_pct = abs(weight) * 100
    ) |>
    arrange(desc(contribution_pct)) |>
    select(term, estimate, weight, contribution_pct)

# CDI calculation helper
calc_cdi <- function(data, weights_df) {
    cdi <- rep(0, nrow(data))
    for (i in seq_len(nrow(weights_df))) {
        feat <- weights_df$term[i]
        w <- weights_df$weight[i]
        if (feat %in% colnames(data)) {
            cdi <- cdi + w * data[[feat]]
        }
    }
    cdi
}

# Get F1-optimized threshold
prob_drought <- predict(final_ridge_fit, df_apr, type = "prob")$.pred_yes

thresholds <- seq(0.1, 0.9, by = 0.05)
best_prob_threshold <- map_dfr(thresholds, function(thresh) {
    pred <- factor(if_else(prob_drought > thresh, "yes", "no"), levels = c("yes", "no"))
    tibble(threshold = thresh, f1 = yardstick::f_meas_vec(df_apr$drought, pred))
}) |>
    filter(f1 == max(f1, na.rm = TRUE)) |>
    slice(1) |>
    pull(threshold)

# Build CDI record
df_cdi <- tibble(
    year = df_apr$pub_year,
    drought_actual = df_apr$drought,
    cdi = calc_cdi(df_apr, cdi_weights),
    prob_drought = prob_drought
) |>
    mutate(cdi_rp = rp_empirical(cdi, direction = "-1"))

cdi_threshold <- min(df_cdi$cdi[df_cdi$prob_drought > best_prob_threshold])
cdi_fires <- df_cdi$cdi >= cdi_threshold
```

## SEAS5 Baseline Comparison

Now we load raw SEAS5 data and calculate z-scores under both baseline periods.

```{r}
#| label: load-seas5-raw

# Load raw SEAS5 data
df_seas5_raw <- cumulus::pg_load_seas5_historical(
    iso3 = "AFG",
    adm_name = PROVINCES_AOI,
    adm_level = 1,
    convert_units = TRUE
)

# Get area weights for province aggregation
df_area <- blob_read(
    name = "ds-aa-afg-drought/raw/vector/historical_era5_land_ndjfmam_lte2025.parquet",
    stage = "dev",
    container_name = "projects"
) |>
    clean_names() |>
    filter(adm1_name %in% PROVINCES_AOI) |>
    distinct(adm1_name, shape_area)

# Aggregate SEAS5 MAM forecast and merge provinces
df_seas5_mam <- cumulus::seas5_aggregate_forecast(
    df = df_seas5_raw,
    value = "mean",
    valid_months = c(3, 4, 5),
    by = c("iso3", "pcode", "name", "issued_date")
) |>
    filter(month(issued_date) == 3) |>
    rename(adm1_name = name, value = mean) |>
    left_join(df_area, by = "adm1_name") |>
    group_by(issued_date) |>
    summarise(
        value = weighted.mean(value, w = shape_area, na.rm = TRUE),
        .groups = "drop"
    ) |>
    mutate(year = year(issued_date))
```

```{r}
#| label: calc-seas5-zscores

# Calculate z-scores under both baselines for comparison
# For 1984 baseline: use 1984-2025 mean/SD
# For 1991 baseline: use 1991-2025 mean/SD

# 1984 baseline statistics
mean_1984 <- mean(df_seas5_mam$value[df_seas5_mam$year >= 1984], na.rm = TRUE)
sd_1984 <- sd(df_seas5_mam$value[df_seas5_mam$year >= 1984], na.rm = TRUE)

# 1991 baseline statistics
mean_1991 <- mean(df_seas5_mam$value[df_seas5_mam$year >= 1991], na.rm = TRUE)
sd_1991 <- sd(df_seas5_mam$value[df_seas5_mam$year >= 1991], na.rm = TRUE)

# Create comparison dataframe with both z-scores
df_seas5_compare <- df_seas5_mam |>
    filter(year >= 1984, year <= 2025) |>
    mutate(
        # Z-scores under each baseline (inverted so positive = drought)
        zscore_1984 = -((value - mean_1984) / sd_1984),
        zscore_1991 = -((value - mean_1991) / sd_1991)
    )

# Add empirical RPs
# For 1984 baseline: RP calculated on all 1984-2025 years
df_seas5_compare <- df_seas5_compare |>
    mutate(seas5_rp_1984 = rp_empirical(zscore_1984, direction = "-1"))

# For 1991 baseline: RP calculated only on 1991+ years
df_1991_rp <- df_seas5_compare |>
    filter(year >= 1991) |>
    mutate(seas5_rp_1991 = rp_empirical(zscore_1991, direction = "-1")) |>
    select(year, seas5_rp_1991)

df_seas5_compare <- df_seas5_compare |>
    left_join(df_1991_rp, by = "year")
```

### Z-Score Distribution Comparison

```{r}
#| label: zscore-distribution
#| fig-height: 5
#| fig-width: 10

# Prepare data for density plot (long format)
df_zscore_long <- df_seas5_compare |>
    filter(year >= 1991) |>  # Only years where both baselines valid
    select(year, zscore_1984, zscore_1991) |>
    pivot_longer(
        cols = c(zscore_1984, zscore_1991),
        names_to = "baseline",
        values_to = "zscore"
    ) |>
    mutate(baseline = if_else(baseline == "zscore_1984", "1984-2025", "1991-2025"))

p1 <- df_zscore_long |>
    ggplot(aes(x = zscore, fill = baseline)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = c("1984-2025" = hdx_hex("sapphire-hdx"),
                                  "1991-2025" = hdx_hex("tomato-hdx"))) +
    labs(
        title = "SEAS5 MAM Z-Score Distribution",
        subtitle = "Comparing baseline periods (1991-2025 years only)",
        x = "Z-score (positive = drought)",
        y = "Density",
        fill = "Baseline"
    )

p2 <- df_seas5_compare |>
    filter(year >= 1991) |>
    ggplot(aes(x = zscore_1984, y = zscore_1991)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(size = 2.5, color = hdx_hex("sapphire-hdx")) +
    geom_smooth(method = "lm", se = FALSE, color = hdx_hex("mint-hdx")) +
    labs(
        title = "Z-Score Comparison: 1984 vs 1991 Baseline",
        subtitle = "Points above diagonal = higher z-score under 1991 baseline",
        x = "Z-score (1984-2025 baseline)",
        y = "Z-score (1991-2025 baseline)"
    )

p1 + p2
```

```{r}
#| label: zscore-correlation

# Correlation between z-scores (on overlapping years only)
cor_zscores <- cor(
    df_seas5_compare$zscore_1984[df_seas5_compare$year >= 1991],
    df_seas5_compare$zscore_1991[df_seas5_compare$year >= 1991],
    use = "complete.obs"
)

# Mean/SD comparison for each baseline period
baseline_stats <- tibble(
    baseline = c("1984-2025", "1991-2025"),
    n_years = c(
        sum(df_seas5_mam$year >= 1984 & df_seas5_mam$year <= 2025),
        sum(df_seas5_mam$year >= 1991 & df_seas5_mam$year <= 2025)
    ),
    mean_raw = c(
        mean(df_seas5_mam$value[df_seas5_mam$year >= 1984], na.rm = TRUE),
        mean(df_seas5_mam$value[df_seas5_mam$year >= 1991], na.rm = TRUE)
    ),
    sd_raw = c(
        sd(df_seas5_mam$value[df_seas5_mam$year >= 1984], na.rm = TRUE),
        sd(df_seas5_mam$value[df_seas5_mam$year >= 1991], na.rm = TRUE)
    )
)

baseline_stats |>
    knitr::kable(
        col.names = c("Baseline", "N Years", "Mean (mm)", "SD (mm)"),
        caption = "SEAS5 MAM climatology statistics by baseline period",
        digits = 1
    )
```

### What Do RP Thresholds Mean in mm?

To make the thresholds tangible, we can back-calculate from z-scores to actual precipitation values. For each RP threshold, we find the corresponding z-score percentile, then convert to mm using the baseline's mean and SD.

```{r}
#| label: mm-thresholds
#| fig-height: 5
#| fig-width: 10

# For each baseline, find the z-score threshold for each RP
# RP = (n+1) / rank, so for RP=4 with n=35, rank = (35+1)/4 = 9
# This means the 9th highest z-score (or ~75th percentile of drought severity)

# Get z-scores for each baseline period
zscores_1984 <- df_seas5_compare$zscore_1984[df_seas5_compare$year >= 1984]
zscores_1991 <- df_seas5_compare$zscore_1991[df_seas5_compare$year >= 1991]

# Function to find z-score threshold for a given RP
get_zscore_threshold <- function(zscores, rp) {
    n <- length(zscores)
    rank <- (n + 1) / rp
    # Sort descending (higher z = more severe drought)
    sorted <- sort(zscores, decreasing = TRUE)
    # Interpolate to get the threshold
    if (rank <= 1) return(sorted[1])
    if (rank >= n) return(sorted[n])
    lower_idx <- floor(rank)
    upper_idx <- ceiling(rank)
    frac <- rank - lower_idx
    sorted[lower_idx] * (1 - frac) + sorted[upper_idx] * frac
}

# Calculate thresholds for RPs 3-7 for both baselines
rp_thresholds <- map_dfr(3:7, function(rp) {
    z_1984 <- get_zscore_threshold(zscores_1984, rp)
    z_1991 <- get_zscore_threshold(zscores_1991, rp)

    # Convert z-score back to mm: value = mean - zscore * sd
    # (since zscore = -(value - mean)/sd for drought)
    mm_1984 <- mean_1984 - z_1984 * sd_1984
    mm_1991 <- mean_1991 - z_1991 * sd_1991

    tibble(
        rp = rp,
        z_1984 = z_1984,
        z_1991 = z_1991,
        mm_1984 = mm_1984,
        mm_1991 = mm_1991
    )
})

# Plot: mm thresholds by RP for each baseline
p_mm <- rp_thresholds |>
    select(rp, mm_1984, mm_1991) |>
    pivot_longer(cols = c(mm_1984, mm_1991), names_to = "baseline", values_to = "mm") |>
    mutate(baseline = if_else(baseline == "mm_1984", "1984-2025", "1991-2025")) |>
    ggplot(aes(x = rp, y = mm, color = baseline, group = baseline)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 3) +
    geom_hline(aes(yintercept = mean_1984), linetype = "dashed", color = hdx_hex("sapphire-hdx"), alpha = 0.5) +
    geom_hline(aes(yintercept = mean_1991), linetype = "dashed", color = hdx_hex("tomato-hdx"), alpha = 0.5) +
    annotate("text", x = 7.2, y = mean_1984, label = paste0("Mean (1984): ", round(mean_1984, 0), " mm"),
             hjust = 0, size = 3, color = hdx_hex("sapphire-hdx")) +
    annotate("text", x = 7.2, y = mean_1991, label = paste0("Mean (1991): ", round(mean_1991, 0), " mm"),
             hjust = 0, size = 3, color = hdx_hex("tomato-hdx")) +
    scale_color_manual(values = c("1984-2025" = hdx_hex("sapphire-hdx"),
                                   "1991-2025" = hdx_hex("tomato-hdx"))) +
    scale_x_continuous(breaks = 3:7, limits = c(3, 8.5)) +
    labs(
        title = "SEAS5 Precipitation Threshold by Return Period",
        subtitle = "Lower precipitation = more severe drought signal",
        x = "Return Period (years)",
        y = "SEAS5 MAM Forecast (mm)",
        color = "Baseline"
    ) +
    theme(legend.position = "bottom")

p_mm
```

```{r}
#| label: mm-threshold-table

rp_thresholds |>
    mutate(
        across(starts_with("z_"), ~ round(.x, 2)),
        across(starts_with("mm_"), ~ round(.x, 0)),
        mm_diff = mm_1991 - mm_1984
    ) |>
    select(rp, z_1984, mm_1984, z_1991, mm_1991, mm_diff) |>
    gt() |>
    tab_header(
        title = "SEAS5 Trigger Thresholds in mm",
        subtitle = "Forecast precipitation below these values triggers early warning"
    ) |>
    tab_spanner(label = "1984-2025 Baseline", columns = c(z_1984, mm_1984)) |>
    tab_spanner(label = "1991-2025 Baseline", columns = c(z_1991, mm_1991)) |>
    cols_label(
        rp = "RP",
        z_1984 = "Z-score",
        mm_1984 = "mm",
        z_1991 = "Z-score",
        mm_1991 = "mm",
        mm_diff = "Δ mm"
    ) |>
    tab_style(
        style = cell_fill(color = "#FFF3E0"),
        locations = cells_body(rows = rp == 4)
    ) |>
    tab_style(
        style = cell_fill(color = "#F3E5F5"),
        locations = cells_body(rows = rp == 6)
    ) |>
    tab_footnote(
        footnote = "Orange = RP4 threshold; Purple = RP6 threshold",
        locations = cells_column_labels(columns = rp)
    )
```

::: {.callout-note}
## Interpreting the Thresholds

```{r}
#| label: threshold-interpretation
#| results: asis

mm_rp4_1991 <- round(rp_thresholds$mm_1991[rp_thresholds$rp == 4], 0)
mm_rp6_1991 <- round(rp_thresholds$mm_1991[rp_thresholds$rp == 6], 0)
mm_rp4_1984 <- round(rp_thresholds$mm_1984[rp_thresholds$rp == 4], 0)
mm_rp6_1984 <- round(rp_thresholds$mm_1984[rp_thresholds$rp == 6], 0)

cat(paste0("**Using 1991-2025 baseline:**\n\n"))
cat(paste0("- RP ≥ 4 triggers when SEAS5 forecasts **< ", mm_rp4_1991, " mm** for MAM\n"))
cat(paste0("- RP ≥ 6 triggers when SEAS5 forecasts **< ", mm_rp6_1991, " mm** for MAM\n\n"))

cat(paste0("**Using 1984-2025 baseline:**\n\n"))
cat(paste0("- RP ≥ 4 triggers when SEAS5 forecasts **< ", mm_rp4_1984, " mm** for MAM\n"))
cat(paste0("- RP ≥ 6 triggers when SEAS5 forecasts **< ", mm_rp6_1984, " mm** for MAM\n\n"))

cat("The 1991 baseline has a slightly different climatological mean, which shifts the mm thresholds accordingly.")
```
:::

**Z-score correlation**: r = `r round(cor_zscores, 3)` — on the overlapping years (1991-2025), the two baselines produce highly correlated z-scores, but the 1991 baseline has a different mean/SD reference point.

### Year-by-Year Comparison

For a fair comparison, we evaluate both SEAS5 baselines on the **overlapping years** (1991-2025) where both can be assessed. This means:

- **1984-2025 baseline**: z-scores calculated using 1984-2025 mean/SD, RP calculated on 1984-2025
- **1991-2025 baseline**: z-scores calculated using 1991-2025 mean/SD, RP calculated on 1991-2025

```{r}
#| label: year-comparison-table

# Combine with CDI and drought status
df_comparison <- df_seas5_compare |>
    left_join(
        df_cdi |> select(year, drought_actual, cdi, cdi_rp),
        by = "year"
    ) |>
    filter(!is.na(drought_actual)) |>
    mutate(
        cdi_fires = cdi >= cdi_threshold,
        seas5_fires_1984 = seas5_rp_1984 >= 4,
        seas5_fires_1991 = seas5_rp_1991 >= 4,
        baseline_disagree = case_when(
            is.na(seas5_rp_1991) ~ NA,
            TRUE ~ seas5_fires_1984 != seas5_fires_1991
        )
    )

# Show years where baselines disagree on SEAS5 trigger (only for 1991+ years)
df_disagree <- df_comparison |>
    filter(!is.na(baseline_disagree), baseline_disagree) |>
    arrange(desc(year))

if (nrow(df_disagree) > 0) {
    df_disagree |>
        select(year, drought_actual,
               zscore_1984, seas5_rp_1984, seas5_fires_1984,
               zscore_1991, seas5_rp_1991, seas5_fires_1991) |>
        mutate(across(contains("zscore"), ~ round(.x, 2))) |>
        mutate(across(contains("rp"), ~ round(.x, 1))) |>
        gt() |>
        tab_header(
            title = "Years Where SEAS5 Baseline Choice Affects Trigger (RP >= 4)",
            subtitle = "Different trigger decisions under 1984 vs 1991 baseline (1991-2025 only)"
        ) |>
        tab_spanner(label = "1984-2025 Baseline", columns = c(zscore_1984, seas5_rp_1984, seas5_fires_1984)) |>
        tab_spanner(label = "1991-2025 Baseline", columns = c(zscore_1991, seas5_rp_1991, seas5_fires_1991)) |>
        tab_style(
            style = cell_fill(color = "tomato", alpha = 0.3),
            locations = cells_body(columns = drought_actual, rows = drought_actual == "yes")
        ) |>
        cols_label(
            year = "Year",
            drought_actual = "Drought",
            zscore_1984 = "Z-score",
            seas5_rp_1984 = "RP",
            seas5_fires_1984 = "Fires",
            zscore_1991 = "Z-score",
            seas5_rp_1991 = "RP",
            seas5_fires_1991 = "Fires"
        )
} else {
    cat("No years have different SEAS5 trigger decisions between baselines at RP >= 4.")
}
```

## Early Warning Performance Comparison

Following the analysis in the trigger proposal chapter, we evaluate SEAS5's value as an early warning signal under both baselines. The key metric is **early warnings per false positive added** — droughts that get one month earlier warning vs false alerts that SEAS5 adds beyond what CDI already triggers.

First, we show the **operational evaluation** where each baseline is evaluated on its full record. This reflects what would actually happen if we used each baseline:

- 1984-2025 baseline: evaluated on all 42 years (1984-2025)
- 1991-2025 baseline: evaluated on 35 years (1991-2025)

```{r}
#| label: early-warning-analysis

# Calculate early warning metrics for each baseline and RP threshold
calc_ew_metrics <- function(df, seas5_rp_col, baseline_label) {
    map_dfr(3:7, function(rp) {
        seas5_fires <- df[[seas5_rp_col]] >= rp

        droughts_early <- sum(seas5_fires & df$drought_actual == "yes", na.rm = TRUE)
        fp_total <- sum(seas5_fires & df$drought_actual == "no", na.rm = TRUE)
        # FP added = SEAS5 fires, CDI doesn't, not a drought
        fp_added <- sum(seas5_fires & !df$cdi_fires & df$drought_actual == "no", na.rm = TRUE)
        n_drought <- sum(df$drought_actual == "yes", na.rm = TRUE)

        # Years where FP added
        years_added <- df$year[seas5_fires & !df$cdi_fires & df$drought_actual == "no"]
        years_added_str <- if (length(years_added) > 0) paste(sort(years_added), collapse = ", ") else "—"

        tibble(
            baseline = baseline_label,
            seas5_rp = rp,
            n_years = nrow(df),
            droughts_early = droughts_early,
            total_droughts = n_drought,
            pct_early = round(droughts_early / n_drought * 100, 0),
            fp_total = fp_total,
            fp_added = fp_added,
            years_fp_added = years_added_str,
            # When fp_added = 0, ratio is infinite (use Inf for plotting)
            ratio = if (fp_added > 0) round(droughts_early / fp_added, 1) else Inf,
            ratio_label = if (fp_added > 0) as.character(round(droughts_early / fp_added, 1)) else paste0(droughts_early, ":0")
        )
    })
}

# OPERATIONAL: Each baseline on its full period
# 1984 baseline on all years (1984-2025)
df_ew_1984_full <- calc_ew_metrics(df_comparison, "seas5_rp_1984", "1984-2025 (n=42)")

# 1991 baseline on 1991+ years only
df_comparison_1991 <- df_comparison |> filter(!is.na(seas5_rp_1991))
df_ew_1991_full <- calc_ew_metrics(df_comparison_1991, "seas5_rp_1991", "1991-2025 (n=35)")

df_ew_operational <- bind_rows(df_ew_1984_full, df_ew_1991_full)
```

### Operational Evaluation (Each Baseline on Full Record)

```{r}
#| label: early-warning-operational-table

df_ew_operational |>
    filter(seas5_rp %in% c(4, 6)) |>
    select(baseline, seas5_rp, droughts_early, pct_early, fp_added, years_fp_added, ratio_label) |>
    gt() |>
    tab_header(
        title = "SEAS5 Early Warning: Operational Performance",
        subtitle = "Each baseline evaluated on its full record"
    ) |>
    cols_label(
        baseline = "Baseline",
        seas5_rp = "RP Threshold",
        droughts_early = "Droughts Early",
        pct_early = "% Early",
        fp_added = "FP Added",
        years_fp_added = "FP Years",
        ratio_label = "Early:FP"
    ) |>
    tab_style(
        style = cell_fill(color = "#FFF3E0"),
        locations = cells_body(rows = seas5_rp == 4)
    ) |>
    tab_style(
        style = cell_fill(color = "#F3E5F5"),
        locations = cells_body(rows = seas5_rp == 6)
    )
```

::: {.callout-warning}
## Key Difference: 1985

The 1984 baseline includes **1985** as an additional false positive at RP ≥ 4. This year is excluded from the 1991 baseline evaluation entirely because it falls outside that period. This means:

- **1984 baseline, RP ≥ 4**: 2 FPs added (1985, 2010)
- **1991 baseline, RP ≥ 4**: 1 FP added (2010 only)
:::

### Apples-to-Apples Comparison (Same Years)

For a fair comparison of how the **baseline statistics** (mean/SD) affect rankings, we also evaluate both baselines on the **same years** (1991-2025, n=35):

```{r}
#| label: early-warning-same-years

# Filter to years where both baselines have valid data (1991+)
df_comparison_overlap <- df_comparison |>
    filter(!is.na(seas5_rp_1991))

n_years_overlap <- nrow(df_comparison_overlap)
n_droughts_overlap <- sum(df_comparison_overlap$drought_actual == "yes")

# Use operational evaluation: each baseline on its FULL period
# This correctly shows 1984 baseline has 2 FPs at RP4 (1985, 2010)
# Clean up baseline labels for table
df_ew_both <- df_ew_operational |>
    mutate(baseline = gsub(" \\(n=\\d+\\)", "", baseline))
```

**Evaluation period**: Each baseline evaluated on its full record (1984-2025: 42 years, 11 droughts; 1991-2025: 35 years, 10 droughts).

```{r}
#| label: early-warning-table

df_ew_both |>
    select(baseline, seas5_rp, droughts_early, pct_early, fp_added, years_fp_added, ratio_label) |>
    pivot_wider(
        names_from = baseline,
        values_from = c(droughts_early, pct_early, fp_added, years_fp_added, ratio_label),
        names_glue = "{baseline}_{.value}"
    ) |>
    select(
        seas5_rp,
        `1984-2025_droughts_early`, `1991-2025_droughts_early`,
        `1984-2025_pct_early`, `1991-2025_pct_early`,
        `1984-2025_fp_added`, `1991-2025_fp_added`,
        `1984-2025_ratio_label`, `1991-2025_ratio_label`
    ) |>
    gt() |>
    tab_header(
        title = "SEAS5 Early Warning Performance by Baseline",
        subtitle = "Each baseline evaluated on its full record"
    ) |>
    tab_spanner(label = "Droughts Early", columns = contains("droughts_early")) |>
    tab_spanner(label = "% Early", columns = contains("pct_early")) |>
    tab_spanner(label = "FP Added", columns = contains("fp_added")) |>
    tab_spanner(label = "Early:FP Ratio", columns = contains("ratio_label")) |>
    cols_label(
        seas5_rp = "SEAS5 RP",
        `1984-2025_droughts_early` = "1984",
        `1991-2025_droughts_early` = "1991",
        `1984-2025_pct_early` = "1984",
        `1991-2025_pct_early` = "1991",
        `1984-2025_fp_added` = "1984",
        `1991-2025_fp_added` = "1991",
        `1984-2025_ratio_label` = "1984",
        `1991-2025_ratio_label` = "1991"
    )
```

```{r}
#| label: early-warning-plot
#| fig-height: 5
#| fig-width: 10

p_early <- df_ew_both |>
    filter(seas5_rp <= 6) |>
    ggplot(aes(x = seas5_rp, y = droughts_early, color = baseline, group = baseline)) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    scale_color_manual(values = c("1984-2025" = hdx_hex("sapphire-hdx"),
                                   "1991-2025" = hdx_hex("tomato-hdx"))) +
    scale_x_continuous(breaks = 3:6) +
    labs(
        title = "Droughts with Early Warning",
        x = "SEAS5 RP Threshold",
        y = "N Droughts",
        color = "Baseline"
    )

# Filter to RP 3-6, identify infinite ratios
df_ratio_plot <- df_ew_both |>
    filter(seas5_rp <= 6) |>
    mutate(is_inf = is.infinite(ratio))

# Separate finite points (for line/dots) and infinite points (for curve to ∞)
df_finite <- df_ratio_plot |> filter(!is_inf)
df_inf <- df_ratio_plot |> filter(is_inf)

# For baselines that go to infinity at RP 6: get the last finite point to start curve
# Only for baselines that have infinite ratio at RP 6
baselines_with_inf <- df_inf$baseline
df_curve_start <- df_finite |>
    filter(baseline %in% baselines_with_inf) |>
    group_by(baseline) |>
    filter(seas5_rp == max(seas5_rp)) |>
    ungroup()

# Arrow endpoint
y_max <- max(df_finite$ratio, na.rm = TRUE)
y_arrow <- y_max + 3

p_ratio <- ggplot() +
    # Main line and points for all finite data
    geom_line(data = df_finite,
              aes(x = seas5_rp, y = ratio, color = baseline, group = baseline),
              linewidth = 1) +
    geom_point(data = df_finite,
               aes(x = seas5_rp, y = ratio, color = baseline), size = 3) +
    # Curved arrow ONLY for baselines with infinite ratio at RP 6
    geom_curve(data = df_curve_start,
               aes(x = seas5_rp, y = ratio, xend = 6, yend = y_arrow, color = baseline),
               curvature = 0.3, linewidth = 1,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed"),
               show.legend = FALSE) +
    # ∞ label (only if there are infinite points)
    {if(nrow(df_inf) > 0) annotate("text", x = 6.08, y = y_arrow, label = "∞", size = 6, fontface = "bold")} +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
    annotate("text", x = 5.9, y = 1.4, label = "1:1 break-even", hjust = 1, size = 3, color = "grey50") +
    scale_color_manual(values = c("1984-2025" = hdx_hex("sapphire-hdx"),
                                   "1991-2025" = hdx_hex("tomato-hdx"))) +
    scale_x_continuous(breaks = 3:6, limits = c(3, 6.3)) +
    coord_cartesian(ylim = c(0, y_arrow + 1.5)) +
    labs(
        title = "Early Warning Value (Early:FP Added)",
        subtitle = "1991 baseline: zero FPs at RP ≥ 6 (ratio → ∞)",
        x = "SEAS5 RP Threshold",
        y = "Ratio",
        color = "Baseline"
    )

p_early + p_ratio + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

### False Positives Added: Detail

```{r}
#| label: fp-added-detail

# Show which years contribute FP under each baseline at RP4
df_fp_detail <- df_ew_both |>
    filter(seas5_rp == 4) |>
    select(baseline, fp_added, years_fp_added)

df_fp_detail |>
    knitr::kable(
        col.names = c("Baseline", "FP Added", "Years"),
        caption = "False positives added by SEAS5 at RP >= 4 threshold"
    )
```

**Verification: All years where SEAS5 (1991 baseline) fires at RP >= 4:**

```{r}
#| label: fp-verification

# Show all years where SEAS5 1991 baseline fires, with their status
df_comparison_overlap |>
    filter(seas5_rp_1991 >= 4) |>
    select(year, drought_actual, cdi_fires, seas5_fires_1991, cdi_rp, seas5_rp_1991) |>
    mutate(
        fp_added = !cdi_fires & drought_actual == "no",
        category = case_when(
            drought_actual == "yes" ~ "True Positive",
            cdi_fires ~ "FP (already in CDI)",
            TRUE ~ "FP Added by SEAS5"
        )
    ) |>
    arrange(desc(year)) |>
    knitr::kable(
        col.names = c("Year", "Drought", "CDI Fires", "SEAS5 Fires", "CDI RP", "SEAS5 RP", "FP Added?", "Category"),
        caption = "All years where SEAS5 (1991 baseline) fires at RP >= 4"
    )
```

## Apples-to-Apples Comparison: Same RP Denominator

The analysis above calculates RP within each baseline's full period (42 years for 1984, 35 years for 1991). This means RP4 represents slightly different percentiles. For a true apples-to-apples comparison, we recalculate RP for **both** baselines using only the **same 35 years** (1991-2025). This isolates the effect of the climatology choice (different mean/SD) on year rankings.

```{r}
#| label: same-denominator-rp

# Recalculate RPs for both baselines using ONLY the 35 overlapping years
df_same_denom <- df_comparison_overlap |>
    mutate(
        # Both RPs calculated on the same 35 years
        seas5_rp_1984_same = rp_empirical(zscore_1984, direction = "-1"),
        seas5_rp_1991_same = rp_empirical(zscore_1991, direction = "-1")
    )

# Check for disagreements under same-denominator RPs
df_same_denom <- df_same_denom |>
    mutate(
        fires_1984_same = seas5_rp_1984_same >= 4,
        fires_1991_same = seas5_rp_1991_same >= 4,
        disagree_same = fires_1984_same != fires_1991_same
    )

n_disagree_same <- sum(df_same_denom$disagree_same, na.rm = TRUE)

# Create full record (1984-2025) with 1991 baseline values as NA for pre-1991 years
df_full_record <- df_comparison |>
    left_join(
        df_same_denom |> select(year, seas5_rp_1984_same, fires_1984_same,
                                 seas5_rp_1991_same, fires_1991_same),
        by = "year"
    )
```

### Year-by-Year: Same Denominator

```{r}
#| label: same-denom-comparison

df_full_record |>
    select(year, drought_actual, cdi_rp, cdi_fires,
           zscore_1984, seas5_rp_1984, fires_1984 = seas5_fires_1984,
           zscore_1991, seas5_rp_1991_same, fires_1991_same) |>
    arrange(desc(seas5_rp_1984)) |>
    mutate(
        is_pre_1991 = year < 1991,
        # Blank out 1991 baseline values for pre-1991 years (misleading to show)
        zscore_1991 = if_else(is_pre_1991, NA_real_, zscore_1991),
        across(contains("zscore"), ~ round(.x, 2)),
        across(contains("rp"), ~ round(.x, 1))
    ) |>
    gt() |>
    cols_hide(columns = is_pre_1991) |>
    sub_missing(missing_text = "") |>
    tab_header(
        title = "Full Record by SEAS5 Signal (1984-2025)",
        subtitle = "Row colors: TP (teal), TN (light teal), FP (light salmon), FN (coral). Violet = SEAS5 fires when CDI doesn't."
    ) |>
    tab_spanner(label = "CDI (April)", columns = c(cdi_rp, cdi_fires)) |>
    tab_spanner(label = "SEAS5 1984 Baseline (n=42)", columns = c(zscore_1984, seas5_rp_1984, fires_1984)) |>
    tab_spanner(label = "SEAS5 1991 Baseline (n=35)", columns = c(zscore_1991, seas5_rp_1991_same, fires_1991_same)) |>
    # TP: CDI fires AND drought = yes (teal)
    tab_style(
        style = cell_fill(color = "#1EBFB3"),
        locations = cells_body(rows = cdi_fires == TRUE & drought_actual == "yes")
    ) |>
    # TN: CDI doesn't fire AND drought = no (light teal)
    tab_style(
        style = cell_fill(color = "#D2F2F0"),
        locations = cells_body(rows = cdi_fires == FALSE & drought_actual == "no")
    ) |>
    # FN: CDI doesn't fire AND drought = yes (coral red)
    tab_style(
        style = cell_fill(color = "#F2645A"),
        locations = cells_body(rows = cdi_fires == FALSE & drought_actual == "yes")
    ) |>
    # FP: CDI fires AND drought = no (light salmon)
    tab_style(
        style = cell_fill(color = "#F7A29C"),
        locations = cells_body(rows = cdi_fires == TRUE & drought_actual == "no")
    ) |>
    # Light violet for SEAS5 columns when SEAS5 fires but CDI doesn't
    tab_style(
        style = cell_fill(color = "#E1BEE7"),  # light violet
        locations = cells_body(
            columns = c(zscore_1984, seas5_rp_1984, fires_1984),
            rows = fires_1984 == TRUE & cdi_fires == FALSE
        )
    ) |>
    tab_style(
        style = cell_fill(color = "#E1BEE7"),  # light violet
        locations = cells_body(
            columns = c(zscore_1991, seas5_rp_1991_same, fires_1991_same),
            rows = fires_1991_same == TRUE & cdi_fires == FALSE
        )
    ) |>
    # Grey background for NA cells (pre-1991 years) - applied last to override row colors
    tab_style(
        style = cell_fill(color = "#E0E0E0"),
        locations = cells_body(
            columns = c(zscore_1991, seas5_rp_1991_same, fires_1991_same),
            rows = is_pre_1991 == TRUE
        )
    ) |>
    cols_label(
        year = "Year",
        drought_actual = "Drought",
        cdi_rp = "RP",
        cdi_fires = "Fires",
        zscore_1984 = "Z-score",
        seas5_rp_1984 = "RP",
        fires_1984 = "Fires",
        zscore_1991 = "Z-score",
        seas5_rp_1991_same = "RP",
        fires_1991_same = "Fires"
    )
```

```{r}
#| label: same-denom-disagree

# Show disagreement years if any
df_disagree_same <- df_same_denom |>
    filter(disagree_same) |>
    arrange(desc(year))

if (nrow(df_disagree_same) > 0) {
    df_disagree_same |>
        select(year, drought_actual,
               zscore_1984, seas5_rp_1984_same, fires_1984_same,
               zscore_1991, seas5_rp_1991_same, fires_1991_same) |>
        mutate(across(contains("zscore"), ~ round(.x, 2))) |>
        mutate(across(contains("rp"), ~ round(.x, 1))) |>
        gt() |>
        tab_header(
            title = "Years Where Baselines Disagree (Same RP Denominator)",
            subtitle = "Different trigger decisions at RP >= 4"
        ) |>
        tab_spanner(label = "1984-2025 Baseline", columns = c(zscore_1984, seas5_rp_1984_same, fires_1984_same)) |>
        tab_spanner(label = "1991-2025 Baseline", columns = c(zscore_1991, seas5_rp_1991_same, fires_1991_same)) |>
        tab_style(
            style = cell_fill(color = "tomato", alpha = 0.3),
            locations = cells_body(columns = drought_actual, rows = drought_actual == "yes")
        ) |>
        cols_label(
            year = "Year",
            drought_actual = "Drought",
            zscore_1984 = "Z-score",
            seas5_rp_1984_same = "RP",
            fires_1984_same = "Fires",
            zscore_1991 = "Z-score",
            seas5_rp_1991_same = "RP",
            fires_1991_same = "Fires"
        )
} else {
    cat("**No disagreements**: Both baselines trigger the same years at RP >= 4 when using the same 35-year denominator.")
}
```

### Early Warning Performance: Same Denominator

```{r}
#| label: same-denom-ew-analysis

# Calculate early warning metrics with same-denominator RPs
df_ew_1984_same <- calc_ew_metrics(df_same_denom, "seas5_rp_1984_same", "1984-2025")
df_ew_1991_same <- calc_ew_metrics(df_same_denom, "seas5_rp_1991_same", "1991-2025")

df_ew_same <- bind_rows(df_ew_1984_same, df_ew_1991_same)
```

```{r}
#| label: same-denom-ew-table

df_ew_same |>
    select(baseline, seas5_rp, droughts_early, pct_early, fp_added, ratio_label) |>
    pivot_wider(
        names_from = baseline,
        values_from = c(droughts_early, pct_early, fp_added, ratio_label),
        names_glue = "{baseline}_{.value}"
    ) |>
    select(
        seas5_rp,
        `1984-2025_droughts_early`, `1991-2025_droughts_early`,
        `1984-2025_pct_early`, `1991-2025_pct_early`,
        `1984-2025_fp_added`, `1991-2025_fp_added`,
        `1984-2025_ratio_label`, `1991-2025_ratio_label`
    ) |>
    gt() |>
    tab_header(
        title = "SEAS5 Early Warning Performance (Same RP Denominator)",
        subtitle = "Both baselines: RP calculated on 35 years (1991-2025)"
    ) |>
    tab_spanner(label = "Droughts Early", columns = contains("droughts_early")) |>
    tab_spanner(label = "% Early", columns = contains("pct_early")) |>
    tab_spanner(label = "FP Added", columns = contains("fp_added")) |>
    tab_spanner(label = "Early:FP Ratio", columns = contains("ratio_label")) |>
    cols_label(
        seas5_rp = "SEAS5 RP",
        `1984-2025_droughts_early` = "1984",
        `1991-2025_droughts_early` = "1991",
        `1984-2025_pct_early` = "1984",
        `1991-2025_pct_early` = "1991",
        `1984-2025_fp_added` = "1984",
        `1991-2025_fp_added` = "1991",
        `1984-2025_ratio_label` = "1984",
        `1991-2025_ratio_label` = "1991"
    )
```

```{r}
#| label: same-denom-ew-plot
#| fig-height: 5
#| fig-width: 10

p_early_same <- df_ew_same |>
    filter(seas5_rp <= 6) |>
    ggplot(aes(x = seas5_rp, y = droughts_early, color = baseline, group = baseline)) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    scale_color_manual(values = c("1984-2025" = hdx_hex("sapphire-hdx"),
                                   "1991-2025" = hdx_hex("tomato-hdx"))) +
    scale_x_continuous(breaks = 3:6) +
    labs(
        title = "Droughts with Early Warning",
        subtitle = "Same RP denominator (35 years)",
        x = "SEAS5 RP Threshold",
        y = "N Droughts",
        color = "Baseline"
    )

# Filter to RP 3-6, separate finite and infinite
df_ratio_same_plot <- df_ew_same |>
    filter(seas5_rp <= 6) |>
    mutate(is_inf = is.infinite(ratio))

df_finite_same <- df_ratio_same_plot |> filter(!is_inf)
df_inf_same <- df_ratio_same_plot |> filter(is_inf)

# Only curve to infinity for baselines that have infinite ratio at RP 6
baselines_with_inf_same <- df_inf_same$baseline
df_curve_start_same <- df_finite_same |>
    filter(baseline %in% baselines_with_inf_same) |>
    group_by(baseline) |>
    filter(seas5_rp == max(seas5_rp)) |>
    ungroup()

y_max_same <- max(df_finite_same$ratio, na.rm = TRUE)
y_arrow_same <- y_max_same + 3

p_ratio_same <- ggplot() +
    geom_line(data = df_finite_same,
              aes(x = seas5_rp, y = ratio, color = baseline, group = baseline),
              linewidth = 1) +
    geom_point(data = df_finite_same,
               aes(x = seas5_rp, y = ratio, color = baseline), size = 3) +
    # Curved arrow ONLY for baselines with infinite ratio at RP 6
    geom_curve(data = df_curve_start_same,
               aes(x = seas5_rp, y = ratio, xend = 6, yend = y_arrow_same, color = baseline),
               curvature = 0.3, linewidth = 1,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed"),
               show.legend = FALSE) +
    {if(nrow(df_inf_same) > 0) annotate("text", x = 6.08, y = y_arrow_same, label = "∞", size = 6, fontface = "bold")} +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
    annotate("text", x = 5.9, y = 1.4, label = "1:1 break-even", hjust = 1, size = 3, color = "grey50") +
    scale_color_manual(values = c("1984-2025" = hdx_hex("sapphire-hdx"),
                                   "1991-2025" = hdx_hex("tomato-hdx"))) +
    scale_x_continuous(breaks = 3:6, limits = c(3, 6.3)) +
    coord_cartesian(ylim = c(0, y_arrow_same + 1.5)) +
    labs(
        title = "Early Warning Value (Early:FP Added)",
        subtitle = "Same RP denominator (35 years) — At RP ≥ 6: ratio → ∞",
        x = "SEAS5 RP Threshold",
        y = "Ratio",
        color = "Baseline"
    )

p_early_same + p_ratio_same + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

::: {.callout-note}
## Same-Denominator Summary

When both baselines use the **same 35-year period** for RP calculation:

- **Trigger disagreements at RP >= 4**: `r n_disagree_same` year(s)
- This isolates the effect of different climatology mean/SD on rankings
- If disagreements exist, they reveal years where the baseline choice materially changes the trigger decision
:::

## Summary & Conclusion

```{r}
#| label: summary-stats

# Key metrics at RP4 - operational (each baseline on its full period)
ew_1984_rp4 <- df_ew_both |> filter(baseline == "1984-2025", seas5_rp == 4)
ew_1991_rp4 <- df_ew_both |> filter(baseline == "1991-2025", seas5_rp == 4)

# Key metrics at RP4 - same denominator
ew_1984_rp4_same <- df_ew_1984_same |> filter(seas5_rp == 4)
ew_1991_rp4_same <- df_ew_1991_same |> filter(seas5_rp == 4)

n_disagree <- sum(df_comparison$baseline_disagree, na.rm = TRUE)

# Check if 1991 baseline is strictly better at any RP (same denominator)
better_1991_same <- df_ew_same |>
    select(seas5_rp, baseline, ratio) |>
    pivot_wider(names_from = baseline, values_from = ratio) |>
    mutate(diff = `1991-2025` - `1984-2025`) |>
    filter(diff > 0) |>
    nrow()

better_1984_same <- df_ew_same |>
    select(seas5_rp, baseline, ratio) |>
    pivot_wider(names_from = baseline, values_from = ratio) |>
    mutate(diff = `1984-2025` - `1991-2025`) |>
    filter(diff > 0) |>
    nrow()
```
::: {.callout-important}
## Key Findings

**Z-Score Correlation:** r = `r round(cor_zscores, 3)` — baseline choice produces highly correlated but systematically shifted z-scores.

### Original Comparison (Different RP Denominators)

RP calculated within each baseline's full period (42 vs 35 years).

**Trigger Disagreement:** `r n_disagree` year(s) differ at RP >= 4.

| Metric | 1984-2025 (n=42) | 1991-2025 (n=35) |
|--------|------------------|------------------|
| Droughts with early warning | `r ew_1984_rp4$droughts_early` | `r ew_1991_rp4$droughts_early` |
| % Early | `r ew_1984_rp4$pct_early`% | `r ew_1991_rp4$pct_early`% |
| FP Added | `r ew_1984_rp4$fp_added` | `r ew_1991_rp4$fp_added` |
| Early:FP Ratio | `r ew_1984_rp4$ratio`:1 | `r ew_1991_rp4$ratio`:1 |

### Apples-to-Apples Comparison (Same RP Denominator)

Both baselines: RP calculated on the same 35 years (1991-2025).

**Trigger Disagreement:** `r n_disagree_same` year(s) differ at RP >= 4.

| Metric | 1984-2025 | 1991-2025 |
|--------|-----------|-----------|
| Droughts with early warning | `r ew_1984_rp4_same$droughts_early` | `r ew_1991_rp4_same$droughts_early` |
| % Early | `r ew_1984_rp4_same$pct_early`% | `r ew_1991_rp4_same$pct_early`% |
| FP Added | `r ew_1984_rp4_same$fp_added` | `r ew_1991_rp4_same$fp_added` |
| Early:FP Ratio | `r ew_1984_rp4_same$ratio`:1 | `r ew_1991_rp4_same$ratio`:1 |

:::

::: {.callout-note}
## Baseline Recommendation

```{r}
#| label: recommendation
#| results: asis

# Use same-denominator comparison for recommendation (fairer comparison)
ratio_diff_same <- abs(ew_1984_rp4_same$ratio - ew_1991_rp4_same$ratio)

if (ratio_diff_same < 0.5 && n_disagree_same <= 2) {
    cat("The SEAS5 early warning performance is **robust to baseline choice**. Under the apples-to-apples comparison (same RP denominator), both baselines produce similar early warning value at RP >= 4. Given the minimal difference, **either baseline is acceptable**:\n\n")
    cat("- **1991-2025**: Aligns with ECMWF's recommendation about pre-1991 data quality\n")
    cat("- **1984-2025**: Maximizes consistency with CDI baseline\n")
} else if (ew_1991_rp4_same$ratio > ew_1984_rp4_same$ratio) {
    cat("The 1991-2025 baseline produces **better early warning performance** (higher early:FP ratio) under the apples-to-apples comparison. This supports ECMWF's concern about pre-1991 forecast quality.\n\n")
    cat("**Recommend adopting 1991-2025 baseline for SEAS5** while keeping CDI on 1984-2025.")
} else {
    cat("The 1984-2025 baseline produces **better early warning performance** under the apples-to-apples comparison. Despite ECMWF's concerns about pre-1991 data, the different climatology (mean/SD) appears to produce better rankings for drought detection.\n\n")
    cat("**Recommend retaining 1984-2025 baseline for SEAS5** for consistency with CDI.")
}
```
:::

## Framing the Choice: Two Equivalent Paths

The analysis above demonstrates that the **baseline choice** (1984 vs 1991) has minimal impact on performance—the real decision is the **RP threshold** (4 vs 6). A lower threshold (RP≥4) provides more early warnings but adds a false positive; a higher threshold (RP≥6) improves precision but sacrifices early warning coverage.

### The Core Insight

SEAS5 serves one purpose: **early warning**. It fires one month before CDI (March vs April), giving humanitarian actors additional lead time. The CDI remains the primary, higher-confidence trigger. Any year that CDI catches, the Combined trigger catches. SEAS5 only matters for the subset of droughts where it provides that extra month.

### Option A: 1991-2025 Baseline with RP ≥ 4

**Narrative:** *"We use SEAS5 forecasts from 1991 onward, following ECMWF guidance about pre-1991 data quality. A 1-in-4 year forecast anomaly triggers early warning."*

| Aspect | Implication |
|--------|-------------|
| **Scientific defensibility** | Aligns with ECMWF's documented concerns about pre-1991 hindcast quality |
| **Threshold consistency** | Both CDI (~RP4) and SEAS5 (RP4) use the same return period—simple to communicate |
| **Historical record** | 35 years (1991-2025) for SEAS5 evaluation |
| **Messaging** | "1-in-4 year event" is intuitive and matches CDI framing |

### Option B: 1984-2025 Baseline with RP ≥ 6

**Narrative:** *"We use the full 42-year SEAS5 record to maximize sample size. A higher threshold (1-in-6 year) compensates for potential noise in early forecasts."*

| Aspect | Implication |
|--------|-------------|
| **Scientific defensibility** | Maximizes sample size; higher threshold implicitly down-weights noisy early years |
| **Threshold consistency** | CDI uses ~RP4, SEAS5 uses RP6—requires explanation of why they differ |
| **Historical record** | 42 years (1984-2025) for SEAS5 evaluation |
| **Messaging** | "1-in-6 year event" sounds more conservative but requires explaining the asymmetry |

### Option C: 1991-2025 Baseline with RP ≥ 6

**Narrative:** *"We use the reliable forecast period (1991+) AND a conservative threshold. This is our most stringent early warning criterion."*

| Aspect | Implication |
|--------|-------------|
| **Scientific defensibility** | Most conservative: reliable data + high threshold |
| **Threshold consistency** | CDI uses ~RP4, SEAS5 uses RP6—still requires explaining asymmetry |
| **Historical record** | 35 years (1991-2025) for SEAS5 evaluation |
| **Messaging** | Very conservative; may miss early warning opportunities |

### What Actually Differs?

```{r}
#| label: what-differs

# Compare all three configurations on 1991-2025 period (where all have valid data)
df_config_compare <- df_same_denom |>
    mutate(
        # Option A: 1991 baseline, RP4
        fires_a = seas5_rp_1991_same >= 4,
        # Option B: 1984 baseline, RP6 (use 1984 z-scores, same-denom RP)
        fires_b = seas5_rp_1984_same >= 6,
        # Option C: 1991 baseline, RP6
        fires_c = seas5_rp_1991_same >= 6
    )

# Early warning metrics for each option
calc_option_metrics <- function(df, fires_col, label) {
    fires <- df[[fires_col]]
    droughts_early <- sum(fires & df$drought_actual == "yes", na.rm = TRUE)
    fp_added <- sum(fires & !df$cdi_fires & df$drought_actual == "no", na.rm = TRUE)
    n_drought <- sum(df$drought_actual == "yes", na.rm = TRUE)
    n_triggers <- sum(fires, na.rm = TRUE)

    # Combined metrics
    combined <- fires | df$cdi_fires
    combined_tp <- sum(combined & df$drought_actual == "yes", na.rm = TRUE)
    combined_fp <- sum(combined & df$drought_actual == "no", na.rm = TRUE)
    combined_fn <- sum(!combined & df$drought_actual == "yes", na.rm = TRUE)

    tibble(
        option = label,
        seas5_triggers = n_triggers,
        droughts_early = droughts_early,
        pct_early = round(droughts_early / n_drought * 100, 0),
        fp_added = fp_added,
        ratio = if (fp_added > 0) paste0(droughts_early, ":", fp_added) else paste0(droughts_early, ":0"),
        combined_tp = combined_tp,
        combined_fp = combined_fp,
        combined_fn = combined_fn,
        combined_f1 = round(2 * combined_tp / (2 * combined_tp + combined_fp + combined_fn), 2)
    )
}

df_options <- bind_rows(
    calc_option_metrics(df_config_compare, "fires_a", "A: 1991 baseline, RP ≥ 4"),
    calc_option_metrics(df_config_compare, "fires_b", "B: 1984 baseline, RP ≥ 6"),
    calc_option_metrics(df_config_compare, "fires_c", "C: 1991 baseline, RP ≥ 6")
)
```

### Three-Way Comparison (1991-2025 evaluation period)

```{r}
#| label: three-way-table

df_options |>
    gt() |>
    tab_header(
        title = "SEAS5 Early Warning: Three Configuration Options",
        subtitle = "All evaluated on 1991-2025 (n=35 years, 10 droughts)"
    ) |>
    tab_spanner(label = "SEAS5 Standalone", columns = c(seas5_triggers, droughts_early, pct_early, fp_added, ratio)) |>
    tab_spanner(label = "Combined Trigger (OR)", columns = c(combined_tp, combined_fp, combined_fn, combined_f1)) |>
    cols_label(
        option = "Configuration",
        seas5_triggers = "Triggers",
        droughts_early = "Early Warnings",
        pct_early = "% Droughts",
        fp_added = "FP Added",
        ratio = "Early:FP",
        combined_tp = "TP",
        combined_fp = "FP",
        combined_fn = "FN",
        combined_f1 = "F1"
    ) |>
    tab_style(
        style = cell_fill(color = "#D2F2F0"),
        locations = cells_body(rows = 1)
    ) |>
    tab_footnote(
        footnote = "FP Added = false positives from SEAS5 that CDI doesn't already trigger",
        locations = cells_column_labels(columns = fp_added)
    )
```

```{r}
#| label: years-by-option

# Show which years each option triggers (SEAS5 only, not CDI)
df_years_by_option <- df_config_compare |>
    select(year, drought_actual, cdi_fires, fires_a, fires_b, fires_c,
           seas5_rp_1991_same, seas5_rp_1984_same) |>
    filter(fires_a | fires_b | fires_c) |>
    arrange(desc(seas5_rp_1991_same))

df_years_by_option |>
    mutate(
        seas5_rp_1991_same = round(seas5_rp_1991_same, 1),
        seas5_rp_1984_same = round(seas5_rp_1984_same, 1),
        early_warning = !cdi_fires & drought_actual == "yes",
        fp_added = !cdi_fires & drought_actual == "no"
    ) |>
    select(year, drought_actual, cdi_fires, seas5_rp_1991_same, fires_a, fires_c,
           seas5_rp_1984_same, fires_b, early_warning, fp_added) |>
    gt() |>
    tab_header(
        title = "Years Triggered by Each SEAS5 Configuration",
        subtitle = "Green = early warning value (SEAS5 fires, CDI doesn't, drought occurs)"
    ) |>
    tab_spanner(label = "1991 Baseline", columns = c(seas5_rp_1991_same, fires_a, fires_c)) |>
    tab_spanner(label = "1984 Baseline", columns = c(seas5_rp_1984_same, fires_b)) |>
    cols_label(
        year = "Year",
        drought_actual = "Drought",
        cdi_fires = "CDI",
        seas5_rp_1991_same = "RP",
        fires_a = "RP≥4",
        fires_c = "RP≥6",
        seas5_rp_1984_same = "RP",
        fires_b = "RP≥6",
        early_warning = "Early Warn",
        fp_added = "FP Added"
    ) |>
    tab_style(
        style = cell_fill(color = "#1EBFB3"),
        locations = cells_body(rows = drought_actual == "yes")
    ) |>
    tab_style(
        style = cell_fill(color = "#F7A29C"),
        locations = cells_body(rows = drought_actual == "no")
    ) |>
    tab_style(
        style = cell_fill(color = "#E1BEE7"),
        locations = cells_body(columns = c(early_warning), rows = early_warning == TRUE)
    ) |>
    cols_hide(columns = c(early_warning, fp_added))
```

### The Real Tradeoff

The table above reveals that **Options B and C produce identical results**—at RP≥6, the baseline choice doesn't matter. The real decision is between **RP≥4 vs RP≥6**:

| Choice | Early Warnings | FP Added | Combined F1 | Effective RP |
|--------|---------------|----------|-------------|--------------|
| **RP ≥ 4** | 6/10 (60%) | 1 | 0.82 | ~3 years |
| **RP ≥ 6** | 4/10 (40%) | 0 | 0.86 | ~3.6 years |

**The question**: Is getting **2 additional early warnings** worth **1 false positive** and a small F1 reduction?

### Recommendation

::: {.callout-tip}
## Preferred: Option A (1991-2025 baseline, RP ≥ 4)

**Rationale:**

1. **Early warning is the whole point**: SEAS5 exists to provide lead time. At RP6, we lose 2 of our 6 early warnings (33% reduction) to avoid 1 false positive.
2. **6:1 ratio is excellent**: Getting 6 early warnings for 1 added FP is a strong value proposition for humanitarian preparedness.
3. **Simpler communication**: Both CDI and SEAS5 use ~RP4 threshold—one number for stakeholders.
4. **Scientific alignment**: Follows ECMWF guidance on SEAS5 hindcast reliability (1991+).

**The counterargument for RP≥6 (Options B/C):**

- Higher Combined F1 (0.86 vs 0.82)
- Zero added false positives
- More conservative = fewer false alarms

Choose RP≥6 if the cost of a false positive outweighs the value of 2 additional early warnings. Choose RP≥4 if early warning lead time is the priority.
:::

## Final Trigger Recommendation

Based on the analysis in this chapter and the preceding trigger modeling chapters, we recommend the following two-stage trigger system:

```{r}
#| label: final-trigger-summary

# Get CDI metrics
n_years_total <- nrow(df_cdi)
n_droughts_total <- sum(df_cdi$drought_actual == "yes")
n_cdi_triggers <- sum(cdi_fires)
cdi_tp <- sum(cdi_fires & df_cdi$drought_actual == "yes")
cdi_fp <- sum(cdi_fires & df_cdi$drought_actual == "no")
cdi_fn <- sum(!cdi_fires & df_cdi$drought_actual == "yes")
cdi_precision <- round(cdi_tp / (cdi_tp + cdi_fp), 2)
cdi_recall <- round(cdi_tp / (cdi_tp + cdi_fn), 2)
cdi_f1 <- round(2 * cdi_precision * cdi_recall / (cdi_precision + cdi_recall), 2)
cdi_rp_effective <- round((n_years_total + 1) / n_cdi_triggers, 1)

# Combined trigger metrics
# Use df_same_denom (1991-2025, n=35) where both SEAS5 baselines have same-denominator RPs
# This ensures consistent evaluation period and RP calculation for SEAS5 and Combined
df_combined_eval <- df_same_denom |>
    mutate(
        # Use 1984 baseline z-scores with RP calculated on 35-year period (same denom)
        seas5_fires_rp4 = seas5_rp_1984_same >= 4,
        combined_fires = cdi_fires | seas5_fires_rp4
    )

n_combined_triggers <- sum(df_combined_eval$combined_fires, na.rm = TRUE)
combined_tp <- sum(df_combined_eval$combined_fires & df_combined_eval$drought_actual == "yes", na.rm = TRUE)
combined_fp <- sum(df_combined_eval$combined_fires & df_combined_eval$drought_actual == "no", na.rm = TRUE)
combined_fn <- sum(!df_combined_eval$combined_fires & df_combined_eval$drought_actual == "yes", na.rm = TRUE)
combined_precision <- round(combined_tp / (combined_tp + combined_fp), 2)
combined_recall <- round(combined_tp / (combined_tp + combined_fn), 2)
combined_f1 <- round(2 * combined_precision * combined_recall / (combined_precision + combined_recall), 2)
n_years_combined <- nrow(df_combined_eval)
combined_rp_effective <- round((n_years_combined + 1) / n_combined_triggers, 1)
```

::: {.callout-important}
## Recommended Trigger Configuration

### Two-Stage Trigger (OR Logic)

| Stage | Signal | Timing | Threshold | Purpose |
|-------|--------|--------|-----------|---------|
| **1. Early Warning** | SEAS5 MAM Forecast | March | RP ≥ 4 | Early mobilization (1 month lead) |
| **2. Confirmation** | Ridge-based CDI | April | F1-optimized (~RP 4) | Higher confidence trigger |

A year triggers if **either** stage fires.

### Baseline Periods

| Component | Baseline | Rationale |
|-----------|----------|-----------|
| CDI | 1984-2025 | Maximize sample size; ASI/ERA5 observations reliable |
| SEAS5 | 1984-2025 or 1991-2025 | Either acceptable; results robust to choice |
:::

### Performance Characteristics

```{r}
#| label: performance-table

# Helper function to format count with years in parentheses
format_with_years <- function(count, years) {
    if (count == 0) return("0")
    paste0(count, " (", paste(sort(years), collapse = ", "), ")")
}

# ============================================================================
# CDI years for FP and FN
# ============================================================================
cdi_fp_years <- df_cdi$year[cdi_fires & df_cdi$drought_actual == "no"]
cdi_fn_years <- df_cdi$year[!cdi_fires & df_cdi$drought_actual == "yes"]
cdi_fp_str <- format_with_years(cdi_fp, cdi_fp_years)
cdi_fn_str <- format_with_years(cdi_fn, cdi_fn_years)

# ============================================================================
# Configuration A: 1991-2025 (n=35), SEAS5 RP >= 4
# ============================================================================
n_droughts_combined <- sum(df_combined_eval$drought_actual == "yes")

# SEAS5 Only metrics (Config A)
seas5_fires_a <- df_combined_eval$seas5_fires_rp4
n_seas5_triggers_a <- sum(seas5_fires_a, na.rm = TRUE)
seas5_tp_a <- sum(seas5_fires_a & df_combined_eval$drought_actual == "yes", na.rm = TRUE)
seas5_fp_a <- sum(seas5_fires_a & df_combined_eval$drought_actual == "no", na.rm = TRUE)
seas5_fn_a <- sum(!seas5_fires_a & df_combined_eval$drought_actual == "yes", na.rm = TRUE)
seas5_precision_a <- round(seas5_tp_a / (seas5_tp_a + seas5_fp_a), 2)
seas5_recall_a <- round(seas5_tp_a / (seas5_tp_a + seas5_fn_a), 2)
seas5_f1_a <- round(2 * seas5_precision_a * seas5_recall_a / (seas5_precision_a + seas5_recall_a), 2)
seas5_rp_effective_a <- round((n_years_combined + 1) / n_seas5_triggers_a, 1)

# Extract years for Config A SEAS5
seas5_fp_years_a <- df_combined_eval$year[seas5_fires_a & df_combined_eval$drought_actual == "no"]
seas5_fn_years_a <- df_combined_eval$year[!seas5_fires_a & df_combined_eval$drought_actual == "yes"]
seas5_fp_str_a <- format_with_years(seas5_fp_a, seas5_fp_years_a)
seas5_fn_str_a <- format_with_years(seas5_fn_a, seas5_fn_years_a)

# Extract years for Config A Combined
combined_fp_years_a <- df_combined_eval$year[df_combined_eval$combined_fires & df_combined_eval$drought_actual == "no"]
combined_fn_years_a <- df_combined_eval$year[!df_combined_eval$combined_fires & df_combined_eval$drought_actual == "yes"]
combined_fp_str_a <- format_with_years(combined_fp, combined_fp_years_a)
combined_fn_str_a <- format_with_years(combined_fn, combined_fn_years_a)

# ============================================================================
# Configuration B: 1984-2025 (n=42), SEAS5 RP >= 6
# ============================================================================
# Use df_comparison filtered to all years with valid 1984 baseline RP
df_eval_1984 <- df_comparison |>
    filter(!is.na(seas5_rp_1984)) |>
    mutate(
        cdi_fires = cdi >= cdi_threshold,
        seas5_fires_rp6 = seas5_rp_1984 >= 6,
        combined_fires = cdi_fires | seas5_fires_rp6
    )

n_years_1984 <- nrow(df_eval_1984)
n_droughts_1984 <- sum(df_eval_1984$drought_actual == "yes")

# SEAS5 Only metrics (Config B: RP >= 6)
seas5_fires_b <- df_eval_1984$seas5_fires_rp6
n_seas5_triggers_b <- sum(seas5_fires_b, na.rm = TRUE)
seas5_tp_b <- sum(seas5_fires_b & df_eval_1984$drought_actual == "yes", na.rm = TRUE)
seas5_fp_b <- sum(seas5_fires_b & df_eval_1984$drought_actual == "no", na.rm = TRUE)
seas5_fn_b <- sum(!seas5_fires_b & df_eval_1984$drought_actual == "yes", na.rm = TRUE)
seas5_precision_b <- round(seas5_tp_b / (seas5_tp_b + seas5_fp_b), 2)
seas5_recall_b <- round(seas5_tp_b / (seas5_tp_b + seas5_fn_b), 2)
seas5_f1_b <- round(2 * seas5_precision_b * seas5_recall_b / (seas5_precision_b + seas5_recall_b), 2)
seas5_rp_effective_b <- round((n_years_1984 + 1) / n_seas5_triggers_b, 1)

# Combined metrics (Config B)
n_combined_triggers_b <- sum(df_eval_1984$combined_fires, na.rm = TRUE)
combined_tp_b <- sum(df_eval_1984$combined_fires & df_eval_1984$drought_actual == "yes", na.rm = TRUE)
combined_fp_b <- sum(df_eval_1984$combined_fires & df_eval_1984$drought_actual == "no", na.rm = TRUE)
combined_fn_b <- sum(!df_eval_1984$combined_fires & df_eval_1984$drought_actual == "yes", na.rm = TRUE)
combined_precision_b <- round(combined_tp_b / (combined_tp_b + combined_fp_b), 2)
combined_recall_b <- round(combined_tp_b / (combined_tp_b + combined_fn_b), 2)
combined_f1_b <- round(2 * combined_precision_b * combined_recall_b / (combined_precision_b + combined_recall_b), 2)
combined_rp_effective_b <- round((n_years_1984 + 1) / n_combined_triggers_b, 1)

# Extract years for Config B SEAS5
seas5_fp_years_b <- df_eval_1984$year[seas5_fires_b & df_eval_1984$drought_actual == "no"]
seas5_fn_years_b <- df_eval_1984$year[!seas5_fires_b & df_eval_1984$drought_actual == "yes"]
seas5_fp_str_b <- format_with_years(seas5_fp_b, seas5_fp_years_b)
seas5_fn_str_b <- format_with_years(seas5_fn_b, seas5_fn_years_b)

# Extract years for Config B Combined
combined_fp_years_b <- df_eval_1984$year[df_eval_1984$combined_fires & df_eval_1984$drought_actual == "no"]
combined_fn_years_b <- df_eval_1984$year[!df_eval_1984$combined_fires & df_eval_1984$drought_actual == "yes"]
combined_fp_str_b <- format_with_years(combined_fp_b, combined_fp_years_b)
combined_fn_str_b <- format_with_years(combined_fn_b, combined_fn_years_b)

# ============================================================================
# Configuration C: 1991-2025 (n=35), SEAS5 RP >= 6
# ============================================================================
df_eval_1991_rp6 <- df_same_denom |>
    mutate(
        seas5_fires_rp6 = seas5_rp_1991_same >= 6,
        combined_fires_c = cdi_fires | seas5_fires_rp6
    )

# SEAS5 Only metrics (Config C: 1991 baseline, RP >= 6)
seas5_fires_c <- df_eval_1991_rp6$seas5_fires_rp6
n_seas5_triggers_c <- sum(seas5_fires_c, na.rm = TRUE)
seas5_tp_c <- sum(seas5_fires_c & df_eval_1991_rp6$drought_actual == "yes", na.rm = TRUE)
seas5_fp_c <- sum(seas5_fires_c & df_eval_1991_rp6$drought_actual == "no", na.rm = TRUE)
seas5_fn_c <- sum(!seas5_fires_c & df_eval_1991_rp6$drought_actual == "yes", na.rm = TRUE)
seas5_precision_c <- round(seas5_tp_c / (seas5_tp_c + seas5_fp_c), 2)
seas5_recall_c <- round(seas5_tp_c / (seas5_tp_c + seas5_fn_c), 2)
seas5_f1_c <- round(2 * seas5_precision_c * seas5_recall_c / (seas5_precision_c + seas5_recall_c), 2)
seas5_rp_effective_c <- round((n_years_combined + 1) / n_seas5_triggers_c, 1)

# Combined metrics (Config C)
n_combined_triggers_c <- sum(df_eval_1991_rp6$combined_fires_c, na.rm = TRUE)
combined_tp_c <- sum(df_eval_1991_rp6$combined_fires_c & df_eval_1991_rp6$drought_actual == "yes", na.rm = TRUE)
combined_fp_c <- sum(df_eval_1991_rp6$combined_fires_c & df_eval_1991_rp6$drought_actual == "no", na.rm = TRUE)
combined_fn_c <- sum(!df_eval_1991_rp6$combined_fires_c & df_eval_1991_rp6$drought_actual == "yes", na.rm = TRUE)
combined_precision_c <- round(combined_tp_c / (combined_tp_c + combined_fp_c), 2)
combined_recall_c <- round(combined_tp_c / (combined_tp_c + combined_fn_c), 2)
combined_f1_c <- round(2 * combined_precision_c * combined_recall_c / (combined_precision_c + combined_recall_c), 2)
combined_rp_effective_c <- round((n_years_combined + 1) / n_combined_triggers_c, 1)

# Extract years for Config C SEAS5
seas5_fp_years_c <- df_eval_1991_rp6$year[seas5_fires_c & df_eval_1991_rp6$drought_actual == "no"]
seas5_fn_years_c <- df_eval_1991_rp6$year[!seas5_fires_c & df_eval_1991_rp6$drought_actual == "yes"]
seas5_fp_str_c <- format_with_years(seas5_fp_c, seas5_fp_years_c)
seas5_fn_str_c <- format_with_years(seas5_fn_c, seas5_fn_years_c)

# Extract years for Config C Combined
combined_fp_years_c <- df_eval_1991_rp6$year[df_eval_1991_rp6$combined_fires_c & df_eval_1991_rp6$drought_actual == "no"]
combined_fn_years_c <- df_eval_1991_rp6$year[!df_eval_1991_rp6$combined_fires_c & df_eval_1991_rp6$drought_actual == "yes"]
combined_fp_str_c <- format_with_years(combined_fp_c, combined_fp_years_c)
combined_fn_str_c <- format_with_years(combined_fn_c, combined_fn_years_c)

# ============================================================================
# Calculate FPs ADDED by SEAS5 (beyond CDI) for interpretation
# ============================================================================
# Config A: SEAS5 FPs that are not already CDI FPs
seas5_fp_added_a <- sum(seas5_fires_a & !df_combined_eval$cdi_fires & df_combined_eval$drought_actual == "no", na.rm = TRUE)

# Config C: SEAS5 FPs that are not already CDI FPs
seas5_fp_added_c <- sum(seas5_fires_c & !df_eval_1991_rp6$cdi_fires & df_eval_1991_rp6$drought_actual == "no", na.rm = TRUE)

# ============================================================================
# Build combined comparison table
# ============================================================================
tibble(
    Metric = c(
        "Evaluation period",
        "Drought years",
        "Years triggered",
        "True positives",
        "False positives",
        "Missed droughts",
        "Precision",
        "Recall",
        "F1 Score",
        "Effective RP"
    ),
    # CDI Only (same for both configs)
    cdi = c(
        paste0(n_years_total, " (1984-2025)"),
        as.character(n_droughts_total),
        as.character(n_cdi_triggers),
        as.character(cdi_tp),
        cdi_fp_str,
        cdi_fn_str,
        as.character(cdi_precision),
        as.character(cdi_recall),
        as.character(cdi_f1),
        paste0("~", cdi_rp_effective, " yr")
    ),
    # Config A: 1991-2025, RP4
    seas5_a = c(
        paste0(n_years_combined, " (1991-2025)"),
        as.character(n_droughts_combined),
        as.character(n_seas5_triggers_a),
        as.character(seas5_tp_a),
        seas5_fp_str_a,
        seas5_fn_str_a,
        as.character(seas5_precision_a),
        as.character(seas5_recall_a),
        as.character(seas5_f1_a),
        paste0("~", seas5_rp_effective_a, " yr")
    ),
    combined_a = c(
        paste0(n_years_combined, " (1991-2025)"),
        as.character(n_droughts_combined),
        as.character(n_combined_triggers),
        as.character(combined_tp),
        combined_fp_str_a,
        combined_fn_str_a,
        as.character(combined_precision),
        as.character(combined_recall),
        as.character(combined_f1),
        paste0("~", combined_rp_effective, " yr")
    ),
    # Config B: 1984-2025, RP6
    seas5_b = c(
        paste0(n_years_1984, " (1984-2025)"),
        as.character(n_droughts_1984),
        as.character(n_seas5_triggers_b),
        as.character(seas5_tp_b),
        seas5_fp_str_b,
        seas5_fn_str_b,
        as.character(seas5_precision_b),
        as.character(seas5_recall_b),
        as.character(seas5_f1_b),
        paste0("~", seas5_rp_effective_b, " yr")
    ),
    combined_b = c(
        paste0(n_years_1984, " (1984-2025)"),
        as.character(n_droughts_1984),
        as.character(n_combined_triggers_b),
        as.character(combined_tp_b),
        combined_fp_str_b,
        combined_fn_str_b,
        as.character(combined_precision_b),
        as.character(combined_recall_b),
        as.character(combined_f1_b),
        paste0("~", combined_rp_effective_b, " yr")
    ),
    # Config C: 1991-2025, RP6
    seas5_c = c(
        paste0(n_years_combined, " (1991-2025)"),
        as.character(n_droughts_combined),
        as.character(n_seas5_triggers_c),
        as.character(seas5_tp_c),
        seas5_fp_str_c,
        seas5_fn_str_c,
        as.character(seas5_precision_c),
        as.character(seas5_recall_c),
        as.character(seas5_f1_c),
        paste0("~", seas5_rp_effective_c, " yr")
    ),
    combined_c = c(
        paste0(n_years_combined, " (1991-2025)"),
        as.character(n_droughts_combined),
        as.character(n_combined_triggers_c),
        as.character(combined_tp_c),
        combined_fp_str_c,
        combined_fn_str_c,
        as.character(combined_precision_c),
        as.character(combined_recall_c),
        as.character(combined_f1_c),
        paste0("~", combined_rp_effective_c, " yr")
    )
) |>
    gt() |>
    tab_header(
        title = "Trigger Performance Comparison",
        subtitle = "Three configurations: 1991 RP≥4, 1984 RP≥6, 1991 RP≥6"
    ) |>
    tab_spanner(
        label = "CDI Only (April)",
        columns = cdi
    ) |>
    tab_spanner(
        label = "1991, RP ≥ 4",
        columns = c(seas5_a, combined_a)
    ) |>
    tab_spanner(
        label = "1984, RP ≥ 6",
        columns = c(seas5_b, combined_b)
    ) |>
    tab_spanner(
        label = "1991, RP ≥ 6",
        columns = c(seas5_c, combined_c)
    ) |>
    # Column group colors - softer palette
    tab_style(
        style = cell_fill(color = "#D2F2F0"),  # light teal for CDI
        locations = cells_body(columns = cdi)
    ) |>
    tab_style(
        style = cell_fill(color = "#D2F2F0"),
        locations = cells_column_labels(columns = cdi)
    ) |>
    tab_style(
        style = cell_fill(color = "#FFF8E1"),  # very light amber for Config A
        locations = cells_body(columns = c(seas5_a, combined_a))
    ) |>
    tab_style(
        style = cell_fill(color = "#FFF8E1"),
        locations = cells_column_labels(columns = c(seas5_a, combined_a))
    ) |>
    tab_style(
        style = cell_fill(color = "#F3E5F5"),  # light violet for Config B
        locations = cells_body(columns = c(seas5_b, combined_b))
    ) |>
    tab_style(
        style = cell_fill(color = "#F3E5F5"),
        locations = cells_column_labels(columns = c(seas5_b, combined_b))
    ) |>
    # Config C styling - light blue for third config
    tab_style(
        style = cell_fill(color = "#E3F2FD"),  # light blue for Config C
        locations = cells_body(columns = c(seas5_c, combined_c))
    ) |>
    tab_style(
        style = cell_fill(color = "#E3F2FD"),
        locations = cells_column_labels(columns = c(seas5_c, combined_c))
    ) |>
    # Style the spanner headers
    tab_style(
        style = list(
            cell_fill(color = "#1EBFB3"),
            cell_text(weight = "bold", size = px(13)),
            cell_borders(sides = c("left", "right"), color = "white", weight = px(2))
        ),
        locations = cells_column_spanners(spanners = "CDI Only (April)")
    ) |>
    tab_style(
        style = list(
            cell_fill(color = "#FFCC80"),
            cell_text(weight = "bold", size = px(13)),
            cell_borders(sides = c("left", "right"), color = "white", weight = px(2))
        ),
        locations = cells_column_spanners(spanners = "1991, RP ≥ 4")
    ) |>
    tab_style(
        style = list(
            cell_fill(color = "#CE93D8"),
            cell_text(weight = "bold", size = px(13)),
            cell_borders(sides = c("left", "right"), color = "white", weight = px(2))
        ),
        locations = cells_column_spanners(spanners = "1984, RP ≥ 6")
    ) |>
    tab_style(
        style = list(
            cell_fill(color = "#64B5F6"),  # blue for Config C spanner
            cell_text(weight = "bold", size = px(13)),
            cell_borders(sides = c("left", "right"), color = "white", weight = px(2))
        ),
        locations = cells_column_spanners(spanners = "1991, RP ≥ 6")
    ) |>
    # Add vertical borders between column groups
    tab_style(
        style = cell_borders(sides = "right", color = "#9E9E9E", weight = px(2)),
        locations = cells_body(columns = cdi)
    ) |>
    tab_style(
        style = cell_borders(sides = "right", color = "#9E9E9E", weight = px(2)),
        locations = cells_body(columns = combined_a)
    ) |>
    tab_style(
        style = cell_borders(sides = "right", color = "#9E9E9E", weight = px(2)),
        locations = cells_body(columns = combined_b)
    ) |>
    cols_label(
        Metric = "",
        cdi = "CDI",
        seas5_a = "SEAS5",
        combined_a = "Combined",
        seas5_b = "SEAS5",
        combined_b = "Combined",
        seas5_c = "SEAS5",
        combined_c = "Combined"
    )
```

### SEAS5 Early Warning Value

Beyond standalone performance, SEAS5 adds value by providing **one month earlier warning** for droughts that CDI would eventually catch anyway.

```{r}
#| label: seas5-value-add

tibble(
    Metric = c(
        "Droughts with early warning",
        "Extra FP from SEAS5",
        "Early:FP ratio"
    ),
    Value = c(
        paste0(ew_1984_rp4_same$droughts_early, "/", ew_1984_rp4_same$total_droughts, " (", ew_1984_rp4_same$pct_early, "%)"),
        as.character(ew_1984_rp4_same$fp_added),
        paste0(ew_1984_rp4_same$ratio, ":1")
    )
) |>
    gt() |>
    tab_header(
        title = "SEAS5 Early Warning Value-Add",
        subtitle = "Additional benefit when used alongside CDI (1991-2025)"
    )
```

::: {.callout-note}
## Interpretation

**CDI (April trigger)** is the primary trigger with strong performance (evaluated on full 1984-2025 record):

- Captures `r cdi_recall * 100`% of droughts (`r cdi_tp`/`r n_droughts_total`)
- Precision of `r cdi_precision * 100`% (`r cdi_fp` false positive(s) in `r n_years_total` years)
- F1 score of `r cdi_f1`

**SEAS5 (March early warning)** provides lead time options (evaluated on 1991-2025):

- **RP ≥ 4**: Identifies `r seas5_recall_a * 100`% of droughts (`r seas5_tp_a`/`r n_droughts_combined`) with `r seas5_fp_added_a` FP(s) added beyond CDI
- **RP ≥ 6**: Identifies `r seas5_recall_c * 100`% of droughts (`r seas5_tp_c`/`r n_droughts_combined`) with `r seas5_fp_added_c` FP(s) added beyond CDI

**Combined trigger (CDI OR SEAS5)** performance depends on SEAS5 threshold (evaluated on 1991-2025):

- **With SEAS5 RP ≥ 4**: `r combined_recall * 100`% recall, `r combined_precision * 100`% precision, F1 = `r combined_f1`
- **With SEAS5 RP ≥ 6**: `r combined_recall_c * 100`% recall, `r combined_precision_c * 100`% precision, F1 = `r combined_f1_c`

:::
