---
title: "Appendix: In-Sample vs Cross-Validated Performance"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Why Does SEAS5-Only Beat Multivariate in Bootstrap CV?

This appendix investigates a seemingly paradoxical result from the March window analysis: the simple SEAS5-only model outperforms a multivariate model in bootstrap cross-validation (F1 ≈ 0.58 vs 0.47).

**The puzzle**: Shouldn't more variables always help, or at least not hurt?

**The answer**: More variables improve *in-sample* fit (mathematically guaranteed), but can hurt *out-of-sample* performance due to overfitting.

```{r}
#| label: setup

library(tidymodels)
library(dplyr)
library(tidyr)
library(cumulus)
library(knitr)

# Load March data
df_mar <- blob_read(
    name = "ds-aa-afg-drought/processed/vector/2026_mar_pub_feature_set_v1.parquet",
    stage = "dev",
    container_name = "projects"
) |>
    mutate(
        drought = factor(
            if_else(
                outcome_asi_zscore >= quantile(outcome_asi_zscore, 1 - 1/4, na.rm = TRUE),
                "yes", "no"
            ),
            levels = c("yes", "no")
        )
    ) |>
    select(-outcome_asi_zscore, -adm0_name.x, -adm0_name.y, -pub_date, -timestep) |>
    drop_na()
```

## The Data

We have `r nrow(df_mar)` years of data with `r sum(df_mar$drought == "yes")` drought years (~`r round(mean(df_mar$drought == "yes") * 100)`%).

**Predictors available in March window:**

- `seas5 Mar-Apr-May`: Seasonal precipitation forecast
- `asi`, `vhi`: Vegetation indices (February values)
- `snow_cover`, `total_precipitation_sum`, `precip_cumsum`: Climate variables
- `volumetric_soil_water_1m`: Soil moisture

## Experiment: In-Sample vs Bootstrap CV

We compare two models:

1. **SEAS5-only**: `drought ~ seas5 Mar-Apr-May`
2. **Multivariate**: `drought ~ all predictors`

```{r}
#| label: define-models

# Model specifications
spec <- logistic_reg() |> set_engine("glm") |> set_mode("classification")

# Recipes
rec_seas5 <- recipe(drought ~ `seas5 Mar-Apr-May`, data = df_mar)
rec_multi <- recipe(drought ~ ., data = df_mar) |>
    update_role(pub_year, new_role = "ID")

# Workflows
wf_seas5 <- workflow() |> add_recipe(rec_seas5) |> add_model(spec)
wf_multi <- workflow() |> add_recipe(rec_multi) |> add_model(spec)
```

### In-Sample Performance (Train and Test on Same Data)

This is "cheating" - we expect multivariate to win or tie.

```{r}
#| label: insample-performance

# Fit on full data
fit_seas5 <- fit(wf_seas5, data = df_mar)
fit_multi <- fit(wf_multi, data = df_mar)

# Predict on same data (in-sample)
pred_seas5 <- predict(fit_seas5, df_mar, type = "prob") |>
    bind_cols(predict(fit_seas5, df_mar, type = "class")) |>
    bind_cols(df_mar |> select(drought))

pred_multi <- predict(fit_multi, df_mar, type = "prob") |>
    bind_cols(predict(fit_multi, df_mar, type = "class")) |>
    bind_cols(df_mar |> select(drought))

# Calculate metrics
insample_seas5 <- tibble(
    model = "SEAS5-only",
    auc = roc_auc_vec(pred_seas5$drought, pred_seas5$.pred_yes),
    f1 = f_meas_vec(pred_seas5$drought, pred_seas5$.pred_class),
    precision = precision_vec(pred_seas5$drought, pred_seas5$.pred_class),
    recall = recall_vec(pred_seas5$drought, pred_seas5$.pred_class)
)

insample_multi <- tibble(
    model = "Multivariate",
    auc = roc_auc_vec(pred_multi$drought, pred_multi$.pred_yes),
    f1 = f_meas_vec(pred_multi$drought, pred_multi$.pred_class),
    precision = precision_vec(pred_multi$drought, pred_multi$.pred_class),
    recall = recall_vec(pred_multi$drought, pred_multi$.pred_class)
)

insample_results <- bind_rows(insample_seas5, insample_multi)

kable(insample_results, digits = 3, caption = "In-Sample Performance (training = test data)")
```

**Result**: Multivariate has higher AUC (`r round(insample_multi$auc, 3)` vs `r round(insample_seas5$auc, 3)`), confirming that more variables improve in-sample fit.

### Bootstrap Cross-Validation (Honest Out-of-Sample)

Now we evaluate honestly - each prediction is made on data the model never saw.

```{r}
#| label: bootstrap-cv

set.seed(42)
boots <- bootstraps(df_mar, times = 30, strata = drought)

# Fit and evaluate SEAS5
cv_seas5 <- fit_resamples(
    wf_seas5,
    resamples = boots,
    metrics = metric_set(roc_auc, f_meas, precision, recall),
    control = control_resamples(save_pred = TRUE)
)

# Fit and evaluate Multivariate
cv_multi <- fit_resamples(
    wf_multi,
    resamples = boots,
    metrics = metric_set(roc_auc, f_meas, precision, recall),
    control = control_resamples(save_pred = TRUE)
)

# Collect metrics
cv_results <- bind_rows(
    collect_metrics(cv_seas5) |> mutate(model = "SEAS5-only"),
    collect_metrics(cv_multi) |> mutate(model = "Multivariate")
) |>
    select(model, .metric, mean, std_err) |>
    pivot_wider(names_from = .metric, values_from = c(mean, std_err))

kable(
    cv_results |> select(model, mean_roc_auc, mean_f_meas, mean_precision, mean_recall),
    digits = 3,
    col.names = c("Model", "AUC", "F1", "Precision", "Recall"),
    caption = "Bootstrap CV Performance (out-of-sample)"
)
```

**Result**: SEAS5-only now wins! The extra variables that helped in-sample are hurting out-of-sample.

## Side-by-Side Comparison

```{r}
#| label: comparison-table

comparison <- tibble(
    Model = c("SEAS5-only", "Multivariate"),
    `In-Sample AUC` = c(insample_seas5$auc, insample_multi$auc),
    `Bootstrap CV AUC` = c(
        cv_results$mean_roc_auc[cv_results$model == "SEAS5-only"],
        cv_results$mean_roc_auc[cv_results$model == "Multivariate"]
    ),
    `In-Sample F1` = c(insample_seas5$f1, insample_multi$f1),
    `Bootstrap CV F1` = c(
        cv_results$mean_f_meas[cv_results$model == "SEAS5-only"],
        cv_results$mean_f_meas[cv_results$model == "Multivariate"]
    )
)

kable(comparison, digits = 3, caption = "In-Sample vs Bootstrap CV Performance")
```

## Why This Happens

```{r}
#| label: explanation-diagram
#| echo: false

tibble(
    Aspect = c(
        "What it measures",
        "Multivariate advantage?",
        "Problem detected",
        "Use for"
    ),
    `In-Sample` = c(
        "Fit to training data",
        "Yes (mathematically guaranteed)",
        "None - overfitting is invisible",
        "Never use for model selection"
    ),
    `Bootstrap CV` = c(
        "Expected performance on new data",
        "Only if variables have real signal",
        "Overfitting penalized",
        "Model selection, honest evaluation"
    )
) |>
    kable(caption = "In-Sample vs Cross-Validation")
```

### The Overfitting Mechanism

1. **In-sample**: The multivariate model finds patterns in the February variables that happen to correlate with drought in this specific 42-year sample.

2. **Out-of-sample**: These patterns don't generalize - they were fitting noise, not signal.

3. **Physical explanation**: February observations (before growing season) shouldn't strongly predict June agricultural drought. SEAS5 forecasts the actual growing season, so it has real predictive power.

## Key Takeaway

::: {.callout-important}
## In-Sample Performance is Misleading

A model can have **better in-sample fit** but **worse out-of-sample performance**. This is why cross-validation is essential for model selection.

The multivariate model's superior in-sample AUC (`r round(insample_multi$auc, 3)` vs `r round(insample_seas5$auc, 3)`) is an illusion - it's fitting noise that won't generalize.
:::

## Technical Note: Factor Levels in R

When using raw `glm()` for binary classification, be aware that `predict(..., type = "response")` returns P(second factor level). With `levels = c("yes", "no")`, this means P("no"), not P("yes").

**tidymodels avoids this confusion** by providing explicit `.pred_yes` and `.pred_no` columns. This appendix uses tidymodels throughout to ensure correct probability interpretation.
